Chromosome Structure I: Loop extrusion or boundary:boundary pairing?

Version 1

Abstract

Abstract

Two different models have been proposed to explain how the endpoints of chromatin looped domains (“TADs”) in eukaryotic chromosomes are determined. In the first, a cohesin complex extrudes a loop until it encounters a boundary element roadblock, generating a stem-loop (and an unanchored loop). In this model, boundaries are functionally autonomous: they have an intrinsic ability to halt the movement of incoming cohesin complexes that is independent of the properties of neighboring boundaries. In the second, loops are generated by boundary:boundary pairing. In this model, boundaries are functionally non-autonomous, and their ability to form a loop depends upon how well they match with their neighbors. Moreover, unlike the loop-extrusion model, pairing interactions can generate both stem-loops and circle-loops. We have used a combination of MicroC to analyze how TADs are organized and experimental manipulations of the even skipped TAD boundary, homie, to test the predictions of the “loop-extrusion” and the “boundary-pairing” models. Our findings are incompatible with the loop-extrusion model and instead suggest that endpoints of TADs in flies are determined by a mechanism in which boundary elements physically pair with their partners, either head-to-head, or head-to-tail, with varying degrees of specificity. How the partners find each other is not clear but is unlikely to require a loop extrusion mechanism.

Introduction

The chromosomes of multicellular animals have a regular and inheritable physical organization. This was first recognized in studies on the lampbrush chromosomes in amphibian oocytes and the polytene chromosomes of insects (Callan, 1963; Nora et al., 2020; Zhimulev and Belyaeva, 1991; Zhimulev et al., 1983). Subsequent research has shown that the architectural features inferred from analysis of lampbrush and polytene chromosomes are present in chromosomes throughout much of the animal kingdom. The key organizational principle is the subdivision of the chromatin fiber into a series of independent looped domains, commonly called “TADs” (Cavalheiro et al., 2021; Chetverina et al., 2017; Jerković et al., 2020; Matthews and White, 2019; Rowley and Corces, 2018). With important exceptions, the arrangement of TADs along a given chromosome tend to be invariant and are largely independent of the cell type or developmental stage. This regular and inheritable organization is a reflection of the underlying mechanism of TAD formation. TADs are separated from each other by special elements called boundaries or insulators. While these elements have been found in many different species, they have been most fully characterized in Drosophila (Cavalheiro et al., 2021; Chetverina et al., 2017). Fly boundaries span DNA sequences of 150 bp to 1.5 kb in length and contain one or more nucleosome-free nuclease-hypersensitive regions. These nuclease-hypersensitive regions are targets for a large collection of DNA binding proteins that have been implicated in boundary function. Some of the fly boundary factors are widely conserved (CTCF, BEN family proteins, GAF), while others appear to be restricted to insect linages (Su(Hw), Pita, Zw5, Pita, Mod(mdg4), BEAF) (Heger et al., 2013; Heger and Wiehe, 2014; Schoborg and Labrador, 2010).

Boundary elements in flies are not only responsible for organizing the chromatin fiber, they also have genetic activities. When interposed between enhancers or silencers and target promoters, boundary elements block regulatory interactions (Bell et al., 2001; Chetverina et al., 2014; Chetverina et al., 2017). This insulating activity provides a mechanism for delimiting units of independent gene activity: genes located between a pair of compatible boundaries are subject to regulatory interactions with enhancers/silencers present in the same chromosomal interval, while they are insulated from the effects of enhancers/silencers located beyond either boundary in adjacent regulatory neighborhoods. Genetic studies suggest that the insulating activity of boundary elements is a consequence of subdividing the chromosome into a series of topologically independent domains (Cai and Shen, 2001; Gohl et al., 2011; Muravyova et al., 2001). Organizing the chromatin fiber into looped domains enhances contacts between sequences within the loop, while it suppresses contacts with sequences outside of the loop. While it is not known whether a similar insulation mechanism is at play in mammals, the fact that mammalian chromosomes are also subdivided into TADs by boundary-like elements suggests that it might be.

A critical question in chromosome biology is the mechanism(s) responsible for determining the endpoints of the loop domains, the TADs. In mammals, a novel loop-extrusion model has been proposed to not only generate chromosomal loops but also determine the endpoints of those loops (Alipour and Marko, 2012; Fudenberg et al., 2016; Guo et al., 2012; Guo et al., 2015; Nuebler et al., 2018; Sanborn et al., 2015). In this model, a cohesin complex initiates loop formation at a loading site within the “loop-to-be” and then extrudes a chromatin loop until it bumps into CTCF-dependent roadblocks, one on each side of the extruding loop. The location of these roadblocks combined with the processive action of the cohesin complex determines the endpoints of each TAD (Davidson and Peters, 2021; Ghosh and Meyer, 2021; Mirny and Dekker, 2022; Perea-Resa et al., 2021). A key assumption of the loop-extrusion model is that mammalian boundaries are fully autonomous: they are roadblocks, and their physical presence in and of itself is sufficient to define the loop endpoint, independent of the functional properties of neighboring boundaries. In a more specific variant of this model, the relative orientation of the paired CTCF roadblocks is also important—in order to halt cohesin-mediated extrusion, the CTCF sites on each side of the loop must have a convergent orientation. In this model the CTCF roadblocks are able to block an oncoming cohesin complex in only one direction; however, their intrinsic blocking activity is independent of the orientation of other CTCF sites in their neighborhood.

While loop extrusion is widely thought to be operative in mammals, the evidence in flies is inconsistent with this mechanism. Genetics studies have shown that fly boundaries are functionally non-autonomous and that their activities in both loop formation and gene regulation depend upon their ability to engage in direct physical interactions with other boundaries (Chetverina et al., 2014; Chetverina et al., 2017). That fly boundaries might function by physical pairing was first suggested by studies which showed that the gypsy transposon su(Hw) boundary and the bithorax complex (BX-C) Mcp boundary can mediate regulatory interactions (enhancer/silencer: reporter) between transgenes inserted at sites separated by 15-32 MB or more (Muller et al., 1999; Sigrist and Pirrotta, 1997; Vazquez et al., 1993). Consistent with direct physical interactions, these distant transgenes co-localize in vivo, and in living tissue remain in contact for extended periods of time (Chen et al., 2018; Li et al., 2011; Vazquez et al., 2006).

Some of the parameters that govern pairing interactions have been defined in two different transgene assays, insulator bypass and boundary competition. In one version of the bypass assay, a set of enhancers are placed upstream of two different reporters (Cai and Shen, 2001; Kyrchanova et al., 2008a; Muravyova et al., 2001). When a single boundary is introduced between the enhancers and the closest reporter, the enhancers are unable to activate either reporter. However, if the same boundary is placed downstream of the closest reporter, bypass is observed. In this case the closest reporter, which is bracketed by the two boundaries, is still insulated from the enhancers; however, the downstream reporter is activated (Muravyova et al., 2001). Kyrchanova et al. (Kyrchanova et al., 2008a) showed that bypass activity is, in most cases, orientation-dependent. When the same boundary is placed in the upstream and downstream position, they typically need to be introduced in the opposite orientation. The reason for this is that self-pairing is head-to-head and this configuration generates a stem-loop topology, bringing the upstream enhancers in close proximity to the downstream reporter. When the two boundaries are introduced in the same orientation, the enhancers do not activate the downstream reporter because a circle-loop rather than a stem-loop is formed (Chetverina et al., 2017; Kyrchanova et al., 2008a). Head-to-head pairing seems to be a common feature of self-pairing interactions between fly boundaries and was observed for scs, scs’, iA2 and wari (Kyrchanova et al., 2008a). A preference for head-to-head interactions is also observed for heterologous pairing interactions between different boundaries in the Abdominal-B (Abd-B) region of the BX-C (Kyrchanova et al., 2011; Kyrchanova et al., 2008b). There are of course some exceptions. The BX-C Fab-7 boundary pairs with itself and with its neighbor Fab-8 in both orientations (Kyrchanova et al., 2011; Kyrchanova et al., 2008b). Bypass activity also depends upon finding the proper match. For example, when multimerized dCTCF or Zw5 binding sites are placed both upstream and downstream of the closest reporter, bypass is observed. However, there is no bypass when multimerized dCTCF sites are tested with multimerized Zw5 sites (Kyrchanova et al., 2008a). Similar results have been reported for the Zipic and Pita factors (Zolotarev et al., 2016). Boundary:boundary pairing preferences are also observed in competition assays in which different boundaries are introduced into the same transgene (Gohl et al., 2011). In these assays, the insulating activity of a boundary in the blocking position (between an enhancer and a reporter that is flanked by a 3’ boundary) is challenged by placing a heterologous boundary upstream of the enhancer. If the boundary upstream of the enhancer is a better match for the boundary located downstream of the reporter, then the insulating activity of the boundary in the blocking position can be compromised or lost altogether. These and other experiments argue that fly boundaries are functionally non-autonomous—that is, their activities depend upon their ability to physically interact with other boundaries.

However, the differences between the two models are not limited to the mechanisms for determining the endpoints of TADs. The models also differ in the possible topologies of the chromatin loops that form TADs. Only two loop topologies can be generated by the cohesin-extrusion mechanism. One is a stem-loop while the other is an unanchored loop. Depending upon the location of the road-blocks that halt cohesin extrusion and the cohesin loading sites, the chromatin fiber would be organized into a series of stem-loops. These stem-loops will be separated from each other by unanchored loops (see Fig. 1B). This is the configuration that is often illustrated in articles discussing the loop-extrusion model. If the relevant road-blocks are very closely juxtaposed, then the unanchored connecting loops will disappear and be replaced by a series of stem-loops which point in opposite orientations, as illustrated in Fig. 1C. Stem-loops are also possible in the boundary-pairing model, and they will be formed when two heterologous boundaries pair with each other head-to-tail (see Fig 1D) in cis. Like loop-extrusion, the stem-loops could be separated from each other by unanchored loops (Fig. 1E). If the boundaries in the neighborhood pair with each other head-to-tail, a series of connected stem-loops pointing in opposite orientations will be generated (Fig. 1D). In this case, the main axis of the chromosome would be a series of paired boundaries (Fig. 1F).

Diagram of the possible loop topologies generated by loop-extrusion and boundary:boundary pairing.

A) Cohesin embraces a loop at a loading site somewhere within the TAD-to-be and then extrudes a loop at an equal rate on both strands. Extrusion continues until cohesin encounters boundary roadblocks on both strands. B) In one model, the orientation of the roadblock is important. As a consequence, the chromatin fiber will be organized into a series of stem-loops separated from each other by unanchored loops. C) If the presence of a boundary roadblock, but not its orientation, is sufficient to halt extrusion, the chromatin fiber will be organized into a series of linked stem-loops. The main axis of the chromosome will be defined by the cohesin:CTCF roadblocks. D) The pairing of two boundaries head-to-tail generates a stem-loop. E) If boundary pairing interactions are strictly pairwise, head-to-tail pairing will generate a series of stem-loops separated from each other by unanchored loops. F) If boundaries can engage in multiple head-to-tail pairing interaction, the chromosome will be organized into a series of linked stem-loops. The main axis of the chromosome will be defined by a series paired boundaries. G) The pairing of two boundaries head-to-head generates a circle-loop. H) If boundary pairing interactions are strictly pairwise there will be a series of circle-loops separated from each other by unanchored loops. I) If boundaries can engage in multiple head-to-head pairing interactions, the chromatin fiber will be organized into a series of circle-loos connected to each other at their base. Pairing interactions between boundaries #1 and #2 need not be in register with pairing of boundaries #2 and #3. In this case, the main axis of the chromosome may bend and twist and this could impact the relative orientation of the circle-loops.

Stem-loops are not, however, the only loop topology that can be generated in the boundary-pairing model. If the neighboring boundaries pair with each other head-to-head instead of head-to-tail, a circle-loop structure will be generated (Fig 1G). Like stem-loops, the circle-loops could be connected by unanchored loops (Fig. 1H). Alternatively, if boundaries pair with both neighbors head-to-head this will generate a series of linked circle-loops oriented in (more or less) the same direction (see Fig. 1I).

In the studies reported here we have critically evaluated these two models. We have used a combination of MicroC to analyze how TADs are organized and experimental manipulations to test the predictions of the “loop-extrusion” and the “boundary-pairing” models. For these experimental manipulations, we have used the well-characterized homie boundary from the even skipped (eve) locus. homie together with nhomie flank the 16 kb eve locus, and these two elements share many of the properties of other fly boundaries (Fujioka et al., 2013; Fujioka et al., 2009). Like the gypsy and Mcp boundaries, nhomie and homie can mediate long-distance regulatory interactions (Fujioka et al., 2016). For example, a lacZ reporter transgene containing homie (or nhomie) inserted on one homolog can be activated by an eve enhancer in a homie- (or nhomie-) containing transgene inserted on the other homolog over a distance of 2 Mb. Long-distance activation is also observed in heterologous combinations of nhomie and homie, but not in combinations with phage lambda DNA. Similarly, regulatory interactions are observed when a homie- or nhomie- containing reporter is inserted at an attP landing site in the hebe gene, 142 kb upstream of the eve locus (Chen et al., 2018; Fujioka et al., 2016; Fujioka et al., 2009). In this case, the eve enhancers drive reporter expression during development in a pattern that mimics the stage-and tissue-specific expression of the endogenous eve gene. Activation also depends upon the orientation of the boundary (5’→3’) relative to the reporter in the transgene. For homie, the reporter must be located “upstream” of the boundary (5’→3’), just like the eve enhancers (and eve gene) are “upstream” of endogenous homie (see Fig. 2B). Little or no activation is observed when the relative orientation is reversed so that the reporter is located “downstream” of homie. For nhomie, the reporter must be located “downstream”, in the same relative position as that in which the eve enhancers and eve gene are located with respect to endogenous nhomie (5’→3’: Fig. 2B).

TAD organization of the even-skipped locus and neighboring sequences.

A) The eve skipped TAD is a volcano with a plume that is anchored by nhomie and homie. ChIP-seq data below the MicroC map indicate many of the known fly chromosomal architectural proteins proteins are associated with the two eve boundary elements in vivo (Cuartero et al., 2014; Duan et al., 2021; Gaskill et al., 2021; Li et al., 2015; Sun et al., 2015; Ueberschär et al., 2019; Zolotarev et al., 2016). B) The eve locus forms a stem-loop structure. In this illustration nhomie pairs with homie head-to-tail, and this configuration forms a stem-loop which brings sequences upstream nhomie and downstream of homie in contact as is observed in insulator bypass assays (Kyrchanova et al., 2008a). The eve locus is shown assembled into a coiled “30 nM” chromatin fiber. C) MicroC contact pattern for the chromosomal region spanning the attP site at –142 kb and the eve locus. Like the eve volcano, this contact pattern was generated using aggregated previously published NC14 data (Batut et al., 2022; Levo et al., 2022). A black arrow indicates the –142kb locus where transgenes are integrated into the genome. A blue arrow indicates position of the Etf-QO gene. Note the numerous TADs separating the –142kb hebe locus from the eve “volcano TAD”. Individual TADs are labeled TA—TM. Black boxes indicate positions of sequences that, based on ChIP experiments, are bound by one or more known insulator proteins in vivo.

Results

The eve TAD

We used MicroC to probe the TAD organization of the eve locus and the genomic region that extends from eve to just beyond the attP site at –142 kb. For this purpose, we chose nuclear cycle 14 embryos. At this stage of development, most nuclei in our sample will have completed replication and will be arrested in G2. As sister chromosomes in flies pair with each other in register, there will be two juxtaposed copies of this region of the chromosome. The sister chromosomes should be linked to each other by cohesin in preparation for the asynchronous nuclear divisions at the end of NC14 (Collier and Nasmyth, 2022; Gligoris et al., 2014). While homolog pairing is limited in pre-cellular blastoderm embryos, there are regions of the chromosome that already show evidence of pairing (Fung et al., 1998). In fact, transvection between gypsy insulator-containing transgenes has been observed in live imaging studies of nuclear cycle 14 embryos (Lim et al., 2018). Since the NC14 nuclei are just emerging from an earlier mitosis and a round of DNA synthesis, one would imagine that other key features of chromosome organization might also be in the process of being established. Thus, it should be possible to capture evidence of intermediates like those expected to be generated as the extruding cohesin complexes move through the eve TAD and other nearby loci at this particular stage of development.

Shown in Fig. 2A is a blow-up of the TAD organization of the eve locus. As expected, the left and right boundaries of the eve TAD correspond to nhomie and homie, respectively, and sequences containing these two boundaries converge at the apex of the eve interaction triangle. All of the sequences within the TAD appear to come in contact with each other and define a high density of internal contact domain (HDIC). These contacts are expected to be generated, at least in part, by the sliding of the chromatin fiber within the TAD against itself, much like sequences in circular/supercoiled DNA can interact with each other (Fig. 1E). There are also sequences within the eve TAD that show enhanced interactions. For example, there is a darker interaction triangle connecting nhomie and the promoter region, while another triangle connects homie (and the PRE adjacent to homie) to the promoter region. These enhanced internal contacts would be expected to help insulate the eve gene from external regulatory elements and at the same time facilitate interactions between eve and its upstream and downstream enhancers.

As predicted by both the loop-extrusion and boundary-pairing models, the eve TAD is a stem-loop, not a circle-loop (Fig. 2B). This is evident from the prominent “volcano plume” at the apex of the eve TAD (volcano triangle). A volcano plume is observed because sequences flanking eve are brought into close proximity when homie and nhomie are linked to each other by either cohesin or by head-to-tail pairing (see Fig.1: see also Ke et al.). Fig. Supplemental 1 shows the hierarchy of interactions between TADs to either side of eve (L-M, K-M and J-M in Sup. Fig. 1). This conclusion is supported by previous transgene studies. As first described in the boundary bypass experiments of Cai and Shen (Cai and Shen, 2001) and Muravyova et al. (Muravyova et al., 2001), the formation of a stem-loop structure when two boundaries are linked together brings sequences immediately beyond the linked boundaries into close proximity with each other. In this assay, a stem-loop structure is required, as it enables enhancers flanking the “upstream” boundary to interact with reporters flanking the downstream boundary (Kyrchanova et al., 2008a). In contrast, a circle-loop configuration does not bring the neighbors to either side of eve into close proximity.

While a stem-loop is the topology expected for the eve TAD in both models, one feature of the TAD is inconsistent with the loop-extrusion model. According to this model, loops (TADs) are generated by a cohesin complex that forms a small bubble at a loading site somewhere in the TAD-to-be and then extrudes both strands of the growing loop at an equal rate until it bumps into the boundary road-blocks (Davidson and Peters, 2021; Ghosh and Meyer, 2021; Mirny and Dekker, 2022; Perea-Resa et al., 2021). If the loading site were located in the middle of the TAD-to-be, and the extrusion intermediates in a population of nuclei were captured by crosslinking, one should observe a crosslink-generated stripe that begins at the loading site and extends perpendicular to the chromosomal DNA until it reaches the apex that links the two TAD boundaries with imaginary lines of 45° (Fig. Supplemental 2B). This is the pattern that would be generated by a “broken zipper” when it transiently links the zipper teeth equidistant from the initial “loading site”. If the loading site is off center to the left, the crosslinking of the passing cohesin complex in a population of nuclei will generate a perpendicular stripe until cohesin come to a halt when it encounters the roadblock on the left (Fig. Supplemental 2C). Assuming that the complex continues to extrude the right strand, it will then generate a stripe at 45° degrees that extends to the apex of the TAD. If the loading site corresponds to one of the CTCF roadblocks or is located close to the roadblock, the cohesin complex will generate a 45° stripe that comes to a halt at the apex of the triangle when it encounters the neighboring CTCF roadblock (Fig. Supplemental 2D and E). If there are multiple loading sites within the TAD, there should be a series of perpendicular stripes that generate and sequentially reinforce the stripe(s) at 45° (Fig. Supplemental 2F). However, in spite of the fact that NC14 fly embryos should provide by far the best prospect of actually capturing loop extrusion intermediates, there are no vertical stripes in the eve TAD, nor are there stripes at 45° that extend to the apex of the eve TAD. Nor do we observe a population of perpendicular stripes that convert into a series of reinforced 45° stripes. There is also no evidence of stripes that are initially tilted to the left or right of a “loading site”, as might be observed if the rate of extrusion were unequal on the two strands, and then subsequently convert to a 45° stripe as cohesin comes to a halt at the first boundary encountered.

TADs in the eve environs

The attP insertion site at –142 kb is near the end of the first exon of the hebe transcription unit (Fig. 2C). It is located just within a ∼14 kb TAD, TB, that has a high density of internal contacts, and extends from the hebe promoter to the 3’ end of the neighboring dila gene. In between hebe and the eve locus there at least are ten distinct TADs, TC-TL (Fig. 2C). The endpoints of most of these TADs correspond to sequences that are associated with one or more known chromosomal architectural factors (dots in Fig. 2C).

These TADs correspond to the fundamental building blocks for organizing the 3D architecture of this chromosomal segment. Superimposed upon these TADs are regions that exhibit lower density of contacts (LDC). For example, the mef2 gene spans two TADs, TJ and TK. TK contains the mef2 distal promoter region and extends to the internal mef2 promoter, while TJ extends to near the promoter of the divergently transcribed Etf-QO gene (blue arrow). Both of these TADs are linked together by a rectangular LDC domain J-K (see Fig. Supplemental 2). Likewise, TJ is linked to its immediate neighbor TI by the LDC domain I-J. In the next level, TADs separated by a single TAD are linked together. In the region immediately above TJ, the LDC domain I-K links TI to TK (Fig. Supplemental 2). Similarly, TJ is linked to TL by J-L while TJ is also linked to the large TAD TM (which contains TER94 and the neighboring gene, pka-R2) on the other side of eve by J-M. The same pattern of a hierarchical series of LDC domains linking TAD neighbors, TAD next-door neighbors and next-next-door neighbors is observed in the DNA segment that includes the hebe gene.

--cohesin-mediated loop extrusion

Like eve, the TADs in the region between eve and the attP site in the hebe locus are defined by right angle triangles (volcanos) with a high density of internal contacts. However, unlike eve, these volcano triangles do not have a plume, and instead are surrounded by a series of LDC rectangular clouds. This is not the contact pattern that one might expect for either a series of stem-loops connected by unanchored loops or a series of stem-loops connected to each other. Moreover, there are no perpendicular or angled crosslinking stripes (broken zippers) emanating from the base of these triangles, nor are their stripes along their 45° legs. Thus, in spite of the fact that the NC14 nuclei used in this analysis are just emerging from an earlier mitosis and a round of DNA synthesis and should be in the process of assembling TADs, the crosslinking signatures that should be generated by the embrace of passing cohesin complexes are completely absent.

In the loop extrusion model, the LDC domains would arise in a subset of nuclei because the extruding cohesin complex breaks through one or more boundary roadblocks before its progression is halted (Fig. Supplemental 3A versus B, C and D) (Hsieh et al., 2020; Krietenstein et al., 2020). Breakthroughs could occur at only one roadblock on one side of the extruding loop (Fig. Supplemental 3B and C), or at multiple roadblocks on one or both sides (Fig. Supplemental 3D), giving rise to a set of overlying LDC domains like that evident in the region between hebe and eve. However, while there would have to be multiple breakthrough events to account for the hierarchical array of LDC domains observed in these NC14 embryos, there is no evidence of the 45° crosslinking stripes that would be expected to mark the legs of the LDC domains following the breakthrough. For example, the I-J LDC domain (Fig. Supplemental 2) could be generated by cohesin complexes that initiated extrusion in either TI or TJ and then failed to halt at the TI:TJ boundary (green arrowhead). In either case, there should be a 45° stripe of crosslinking that marks (at least part of) the left (TI) and/or the right (TJ) leg of the I-J LDC domain and extends to the apex of the TI-TJ LDC triangle; however, there is no evidence of such stripes. Nor is there evidence of crosslinking stripes marking one or both legs of the J-K or I-K LDC domains (Fig. Supplemental 2).

--boundary:boundary pairing

At least two different mechanisms are expected to account for the TADs and LDC domains in the region between eve and the attP insertion site at -142 kb. First, since none of the TADs in this DNA segment are topped by volcano plumes, they could correspond to circle-loops. As circle-loops are generated by head-to-head pairing, the boundaries in the region between nhomie and hebe would be expected to pair with each other head-to-head. (The TAD to TAD contact maps generated by circle-loops and stem-loops are considered further in the accompanying paper, Ke et al.)

As illustrated in Fig. Supplemental 3E-G, the coiled circle-loop TADs, though topologically independent, are expected to be in relatively close proximity to each other. This means that in addition to crosslinking events within the TADs, there will be crosslinking events between neighboring TADs. These crosslinking events could generate the LDCs seen in Fig. 2C and Fig. Supplemental 2. Since TADs next to each other (Fig. Supplemental 3E and F; Fig. Supplemental 2 I-J or J-K) would typically be expected to interact more frequently than TADs separated by one, two or more TADs (Fig. Supplemental 3G; Fig. Supplemental 2 I-K or J-L), there should be a progressive reduction in contact frequency at each step. This is generally what is seen (see Fig. Supplemental 2). While connected stem-loops generated by cohesin extrusion should also bump into one another, contacts between next-door neighbors might be expected to be less frequent than contacts between next-next-door neighbors (Fig. 1B, C and Fig. Supplemental 3A). (TAD-to-TAD crosslinking and the impact of loop topology is further examined in Ke et al., 2023.)

The second mechanism would be switching and/or combining pairing partners (Fig. Supplemental 3H-J). Though imaging studies suggest that pairing interactions can be of relatively long duration (>30 min: Chen et al., 2018; Vazquez et al., 2006)), they are not permanent, and other nearby boundaries can compete for pairing interactions (Gohl et al., 2011). If switching/combining occurs in NC14 embryos (see Fig. Supplemental 3H-J), then the precise pattern of TADs and LDC domains could also be impacted by the relative avidity of potential pairing partners in the neighborhood and also the distances separating potential partners. For example, the rectangle linking TH and TI (H-I in Fig. Supplemental 2) could represent partner switching so that the boundary between TI and TJ (green arrowhead in Fig. Supplemental 2) sometimes pairs with the boundary separating TH and TI (blue arrowhead) and sometimes with the boundary separating TG and TH (red arrowhead in Fig. Supplemental 1). Distant boundary elements could also preferentially interact with each other, generating a supra-TAD that contains multiple TAD domains. For example, at the apex of the LDC domain B-I there is an interaction dot (purple arrow, Fig. Supplemental 2) that links the TA-TB boundary to the TI-TJ boundary (green arrowhead).

Activation of a distant reporter by the eve enhancers

In previous studies we found that a minimal 367 bp homie fragment can orchestrate regulatory interactions between the eve enhancers and reporters inserted at an attP site in the first intron of the hebe transcription unit, 142 kb upstream of the eve gene (Fujioka et al., 2016). This minimal fragment lacks the single homie CTCF site, but has sites for several other generic boundary proteins including BEAF, Zipic, Su(Hw), Pita and Ibf2. In blastoderm stage embryos, lacZ reporter expression was observed in a 7-stripe pattern that coincided with the stripes of the endogenous gene. However, at this stage only a subset of eve-expressing nuclei expressed the reporter. Later in development, during mid-embryogenesis, reporter expression was observed in cells in the dorsal mesoderm, the ventral CNS, and in the anal plate region, in a pattern that recapitulated eve gene expression (Fujioka et al., 2016). At these later stages, most of the eve- expressing cells also appear to express the reporter. The eve-dependent activation of the reporter in a stripe pattern has also been visualized in live imaging of pre-cellular blastoderm-stage embryos (Chen et al., 2018). Within the 7 eve stripes, the majority of the nuclei (∼80-85%) did not express the lacZ reporter. In these nuclei, the transgene was found to be at a considerable average physical distance from the eve locus (∼700 nM). In the remaining nuclei, the transgene was located in closer proximity to the eve locus (∼350 nM, on average). Within this subset, the transgene was expressed in most of the nuclei. Moreover, in those cases in which the transgene was active, it remained in close proximity and continued to express lacZ for the duration of the experiment (∼30 min).

In the two models for TAD formation, quite different mechanisms must be invoked to account for activation of the reporter at –142 kb by the eve enhancers. In the boundary-pairing model, the transgene homie boundary at –142 kb loops over the intervening TADs and pairs with the nhomie:homie complex flanking the eve TAD (c.f., Fig. 3B and C). In the loop-extrusion model, a cohesin complex initiating loop extrusion in the eve TAD must break through the nhomie roadblock at the upstream end of the eve TAD. It must then make its way past the boundaries that separate eve from the attP site in the hebe gene, and come to a halt at the homie boundary associated with the lacZ reporter. This would generate a novel TAD, eveMammoth (eveMa), that extends from the eve homie all the way to the homie fragment at –142 kb, and encompasses both the reporter and the eve gene, including its enhancers (c.f., Fig. 3D). Of course, the eveMa TAD could also be generated by a cohesin complex that initiated in, for example, TF. However, in this case, the runaway cohesin complex would have to break through the intervening boundary roadblocks in both directions. In both the boundary-pairing and loop-extrusion models, the configuration of the chromatin fiber would lead to the activation of lacZ expression by the eve enhancers, while the reporter would be protected from the hebe enhancers by the homie boundary.

Schematics of boundary pairing and loop extrusion.

A) LhomieG Z5 and GhomieL Z5 transgenes and the eve locus on a linear map. The same color codes are used throughout. In the eve locus, nhomie (blue arrow) and homie (red arrow) are oriented, by convention, so that they are pointing towards the right. This convention is maintained in the two transgenes. In LhomieG Z5 homie (top) is in the opposite orientation from homie in the eve locus is pointing away from the eve locus. In GhomieL Z5 homie (bottom) homie is in the same orientation as the homie in the eve locus and is pointing towards the eve locus. B) Predicted boundary pairing interactions between LhomieG Z5 and the eve locus. homie in the transgene pairs with homie in the eve locus head-to-head. Since homie in the transgene is pointing in the opposite orientation from homie in the eve locus, a stem-loop will be generated. If homie in the transgene also pairs with nhomie in the eve locus head-to-tail, a loop structure like that shown in B will be generated. In this topology, eve-lacZ is in close proximity to eve enhancers and eve-gfp is in contact with the hebe enhancer. C) Predicted boundary pairing between GhomieL Z5 and the eve locus. The GhomieL Z5 transgene is inserted in the same chromosomal orientation as LhomieG Z5; however, the homie boundary is inverted so that it is pointing towards the eve in the same orientation as the homie in the eve locus. Homie in the transgene will pair with homie in the eve locus head-to-head, and this generates a circle-loop. If homie in the transgene also pairs with nhomie in the eve loucs a loop structure like that shown in C will be generated. In this topology eve-gfp will be activated by both the eve and hebe enhancers. D) Loop extrusion model for LhomieG Z5. Transgene homie and endogenous homie determine the endpoints of the extruded eveMammoth (eveMa) loop. In this topology, eve-lacZ is in close proximity to the eve enhancers, while eve-gfp is close to the hebe enhancers. E) Loop extrusion model for GhomieL Z5. The cohesin complex bypasses transgenic homie and is stopped at an upstream boundary X to generate a novel eveGarantuan (eveGa) loop. Both eve-gfp and eve-lacZ are located within the same eveGa TAD, and thus should interact with eve. Since the hebe enhancers are within this TAD as well, they would also be able to activate both reporters.

To test these two models, we used a transgene that has two divergently transcribed reporters, lacZ and gfp, that are each under the control of an eve promoter (Fig. 3A). Two different versions of the transgene were generated. In the first, LhomieG, the eve-lacZ reporter is located “upstream” of homie (with respect to the 5’→3’orientation of homie in the eve locus), while eve-gfp is located downstream (Fig. 3A: top). In the second, the orientation of homie within the transgene is reversed, so that gfp is upstream of homie, while lacZ is downstream (GhomieL) (Fig. 3A: bottom). These two transgenes were then individually inserted into the attP site at –142 kb, in either the Z5 or Z3 orientation. In the Z5 orientation, the eve-lacZ reporter is on the eve side of the homie boundary, so that it is closer to the eve locus (Fig. 3A). The eve-gfp reporter is on the opposite side of the homie boundary and is farther away from the eve locus. This places the eve-gfp reporter next to a series of hebe enhancers located farther upstream, in the intron of the hebe gene, and thus it should be subject to regulation by these enhancers (Fig. 2C). In the Z3 orientation, the entire transgene is inserted in the opposite orientation so that eve-gfp is on the eve side of homie, while eve-lacZ is next to the hebe enhancers (Fig. 6A). As a control, we also generated a Z5 transgene in which lambda DNA instead of homie was inserted between the eve-gfp and eve-lacZ reporters.

LhomieG Z5 and GhomieL Z5

The results for transgenes LhomieG Z5 and the control LlambdaG Z5 are most straightforward and will be considered first. In both the loop-extrusion and boundary-pairing models, the LlambdaG Z5 control is not expected to display any regulatory interactions with the eve locus. Fig. 4C shows that this is the case: the eve-lacZ and eve-gfp reporters in LlambdaG Z5 embryos are silent at the blastoderm stage, and transcripts are only very rarely detected. Later in development, both reporters are expressed in a repeating pattern along the dorsal midline (Fig. 6E). This expression is driven by the hebe gene enhancers located just beyond the gfp-reporter (Fujioka et al., 2009).

The lacZ reporter is activated by eve enhancers in the LhomieG Z5 transgene.

A) Expression patterns of endogenous eve in early (stage 4-5) and late (stage 14-16) embryos by smFISH. eve (Atto 633) probes were used to hybridize with eve mRNA. eve is shown in yellow and DAPI in blue. N>3. B) Schematics of transgenes. C) Expression patterns of transgenic reporters in early stages embryos. Top: LlambdaG Z5. Bottom: LhomieG Z5. GFP is in green, lacZ is in orange, and DAPI is in blue, here and in E. D) Quantification of normalized stripe signals for transgenes shown in C. The reporter smFISH intensity (signal to background), here and in F, was measured, normalized, and plotted as described in Methods. N>3. n=27 for LlambdaG Z5 and n=42 for LhomieG Z5. E) Expression patterns of transgenic reporters in late stages embryos. Top: LlambdaG Z5. Bottom: LhomieG Z5. F) Quantification of normalized stripe signals for transgenes shown in E. N>3. n=27 for LlambdaG Z5 and n=56 for LhomieG Z5. Scale bars = 200μm. N = # independent biological replicates. n = # animals. The paired two-tailed t-test were used for statistical analysis. *p < 0.05, **p < 0.01, ***p < 0.001, ****p < 0.0001, ns: not significant. Raw measurements are available in the Source Data files.

In the case of the LhomieG Z5 insert, regulatory interactions between the transgene and the eve locus will be observed if a stem-loop is established that links the transgene to eve. In the loop-extrusion model, the breakthrough cohesin complexes would generate a stem-loop, eveMammoth (eveMa), by coming to a halt at the homie boundary in the transgene and the homie boundary in eve, as illustrated in Fig. 3D. This will place the lacZ reporter in close proximity to the eve enhancers. It will also disrupt the eve TAD, as nhomie is no longer linked to the homie boundary by cohesin. In the boundary-pairing model, the lacZ reporter in the transgene is “upstream” of homie just like the eve gene and its enhancers are “upstream” of homie in the eve locus. Since homie pairs with itself head-to-head, pairing between homie in the transgene and homie in the eve locus will generate the same stem-loop as that generated by loop-extrusion. However, because homie pairs with nhomie head-to-tail, a more complex structure like that in Fig. 3B would be expected if the pairing interactions with the transgene do not disrupt the eve TAD. Consistent with a tripartite structure of the sort shown here, previous studies have shown that three Mcp transgenes on three different chromosomes can interact with each other genetically (Muller et al., 1999). Subsequent direct visualization of Mcp-mediated pairing interactions in imaginal discs by Vazquez et al. (Vazquez et al., 2006) showed that four Mcp- containing transgene separated by Mbs (at Bridges cytogenetic intervals ∼65 on the left arm of chromosome 3, and ∼83 and ∼95 on the right arm of chromosome 3) and/or located on different homologs, clustered in the same nuclear foci in 94% of the nuclei.

As predicted from the formation of a stem-loop linking the transgene to the eve locus, lacZ is expressed in blastoderm-stage embryos in 7 stripes that coincide with the 7 stripes of the endogenous eve gene. However, unlike eve, not all nuclei in the seven stripes express lacZ (compare with the eve control in Fig. 4A). Later, during mid-embryogenesis, lacZ is expressed in the mesoderm, the CNS, and the anal plate in a pattern that mimics the endogenous eve gene. At this stage almost all eve positive cells are also positive for lacZ. Unlike the LlambdaG Z5 transgene, the hebe enhancers do not drive eve-lacZ expression, as they are beyond the homie boundary (Fig. 4E). The eve-gfp reporter in the LhomieG Z5 transgene responds differently. As was observed for the LlambdaG Z5 control, the eve enhancers drive little if any gfp expression, either at the blastoderm stage or later in development (Fig. 5C, E). Instead, gfp is expressed in the midline during mid-embryogenesis under the control of the hebe enhancers (Fig. 5E), which should be in the same TAD as eve-gfp. These results would be consistent with both the loop-extrusion (eveMa) and boundary-pairing models (Fig. 3B and D).

The GFP reporter is activated by both eve and hebe enhancers in the LhomieG Z5 transgene.

A) Schematics of transgenes. B) Expression patterns of transgenic reporters in early stage embryos. Top: LlambdaG Z5. Bottom: GhomieL Z5. GFP is in green, lacZ is in orange, and DAPI is in blue, here and in D. C) Quantification of normalized stripes signal as represented in B. The reporter smFISH intensity (signal to background) was measured, normalized, and plotted as described in Methods, here and in E. N>3. n=27 for LlambdaG Z5, and n=46 for LhomieG Z5. D) Expression patterns of transgenic reporters in late stage embryos. Top: LlambdaG Z5. Bottom: GhomieL Z5. E) Quantification of normalized stripe signals for transgenes shown in D. N>3. n=27 for LlambdaG Z5, and n=59 for LhomieG Z5. Scale bars = 200μm. N = # independent biological replicates. n = # animals. The paired two-tailed t-test were used for statistical analysis. *p < 0.05, **p < 0.01, ***p < 0.001, ****p < 0.0001, ns: not significant. Raw measurements are available in the Source Data files.

GhomieL Z5

The situation is more complicated for the GhomieL Z5 transgene. This transgene is inserted in the same chromosomal orientation as LhomieG Z5: eve-lacZ is on the eve side of the homie boundary and eve-gfp is on the hebe enhancer side of the boundary. However, the orientation of the boundary within the transgene is switched so that eve-gfp rather than eve-lacZ is “upstream” of homie. In the boundary-pairing model, flipping the orientation of homie in the transgene but keeping the same orientation of the transgene in the chromosome will have two consequences. First, as illustrated in Fig. 3C, the topology of the chromatin loop connecting the homie boundary in the transgene and the eve locus will change from a stem-loop to a circle-loop. The reason for the change in loop topology is that homie pairs with homie head-to-head, and with nhomie head-to-tail (Fujioka et al., 2016). As before, the transgene homie will pair with eve homie head-to-head to generate a simple circle-loop; however, if, as expected, it also pairs with nhomie so that the eve TAD remains intact, a more complicated loop structure would be generated (Fig. 3C). Second, the reporter that is preferentially activated by the eve enhancers will switch from eve-lacZ to eve-gfp. In order to be activated by the eve enhancer, the reporter must be “upstream” of the transgene homie. This physical relationship means that the eve-gfp reporter will be activated by the eve enhancers independent of the orientation of the transgene in the chromosome. On the other hand, the eve-lacZ reporter will not be activated by the eve enhancers. In addition, it will still be insulated from the hebe enhancers by the homie boundary.

It is not possible to form a circle-loop in the loop extrusion model. Instead, the breakthrough cohesin complex will come to a halt as before when it encounters the inverted homie boundary at –142 kb (Fig. 3D). This means that the eve-lacZ reporter will be in the same TAD (eveMa) as the eve enhancers and will be activated by the eve enhancers. In contrast, the eve-gfp reporter will be in the neighboring TAD and will not be activated by the eve enhancers. It will, however, be regulated by the nearby hebe enhancers.

Fig. 5 shows that the four predictions of the boundary-pairing model are correct: a) the eve enhancers drive eve-gfp expression, b) the hebe enhancers drive eve-gfp expression, c) eve-lacZ is not subject to regulation by the eve enhancers, and d) eve-lacZ is insulated from the hebe enhancers. This would also mean that pairing of the transgene homie in the GhomieL Z5 insert with homie (and nhomie) in the eve locus generates a circle-loop, not a stem-loop, as was the case for LhomieG Z5. In contrast, since loop-extrusion cannot generate circle-loops, the key predictions of this model, namely that eve-gfp will be silent while eve-lacZ will be regulated by the eve enhancers, are not satisfied. Instead, eve-gfp is activated by the eve enhancers, while eve-lacZ is not.

While the results described above for the GhomieL Z5 transgene are inconsistent with a “simple” loop-extrusion model, there is one potential caveat: the 5’→3’ orientation of the homie boundary is inverted so that it is “pointing” towards the eve locus rather than away from it like the endogenous homie boundary. In this orientation, it is possible that homie no longer functions as an effective roadblock, and instead the cohesin complex emanating from the eve locus (or from a loading site somewhere in between) slips past this homie fragment and continues until it is blocked by a properly oriented boundary further upstream. In this case the newly formed stem-loop, eveGargantuan (eveGa) (Fig. 3E), would include the eve-gfp reporter, and it would then be activated by the eve enhancers. However, there are two problems with postulating this novel eveGa TAD. First, eve-lacZ would be in the same eveGa TAD as eve-gfp, and it should also be activated by the eve enhancers. Second, since the homie boundary was bypassed by the cohesin complex to form the larger eveGa TAD, it should not be able to block the hebe enhancers from activating eve-lacZ. Neither of these predictions is correct.

GhomieL Z3 and LhomieG Z3

To confirm these conclusions, we examined the expression patterns of the GhomieL and LhomieG transgenes when they are inserted in the opposite orientation (Z3) in the –142 kb attP site. In the boundary-pairing model, the orientation of homie in the transgene determines which reporter will be preferentially activated, while the orientation of the transgene homie in the chromosome determines whether a stem-loop or a circle loop will be formed. In GhomieL Z3 a stem-loop will be formed, and eve-gfp will be activated by the eve enhancers; however, since eve-lacZ is adjacent to the hebe enhancers, they will activate it rather than the eve-gfp reporter (Fig. 6B). In LhomieG Z3 the topology of the loop will switch to a circle-loop, and eve-lacZ will be activated by both the eve and hebe enhancers (Fig. 6C). In the first version of the loop-extrusion model, the eve-gfp reporter in both GhomieL Z3 (Fig. 6D) and LhomieG Z3 (not shown), should be activated by the eve enhancers, since it is included in the eveMa TAD when the transgenes are inserted in the Z3 orientation, while eve-lacZ will be regulated by the hebe enhancers (Fig. 6D). In the second version of the loop-extrusion model, the functioning of the homie roadblock depends upon its orientation relative to endogenous homie. In this model, GhomieL Z3 is expected to form the eveMa TAD (Fig. 6D), while LhomieG Z3 will form the larger eveGa TAD (Fig. 6E).

Expression of reporters in Z3 transgenes.

A) Schematics of Z3 transgenes. B) Boundary pairing for GhomieL Z3 transgene long-range interactions. In this topology, GFP is in close proximity to eve enhancers, and lacZ is in contact with the hebe enhancer region. C) Boundary pairing for the LhomieG Z3 transgene. In this topology, lacZ is in close proximity to both the eve enhancers and the hebe enhancers. D) Loop extrusion model for GhomieL Z3. Transgenic homie and endogenous homie flank the eveMammoth loop. In this topology, GFP is in close proximity to eve enhancers, and lacZ is close to the hebe enhancers. E) Loop extrusion model for the LhomieG Z3 transgene. The cohesin complex bypasses transgenic homie and is stopped at an upstream boundary X to create the eveGarantuan loop. Both GFP and lacZ are at a similar physical distance from eve enhancers and hebe enhancers in this topology. F) Expression patterns of transgenic reporters in early stage embryos. Top: GhomieL Z3. Bottom: LhomieG Z3. GFP is in green, lacZ is in orange, and DAPI is in blue. N=3, n=24 for both GhomieL Z3 and LhomieG Z3. G) Expression patterns of transgenic reporters in late stage embryos. Top: GhomieL Z3. Bottom: LhomieG Z3. N=3, n=24 for both GhomieL Z3 and LhomieG Z3. Scale bars = 200μm, N = # independent biological replicates, n = # animals.

As would be predicted by all three models, the eve enhancers activate the eve-gfp reporter in the GhomieL Z3 insert (Fig. 6F and G). Likewise, homie blocks the hebe enhancers from activating the eve-gfp reporter, while they turn on eve-lacZ expression instead. As predicted by the boundary-pairing model, eve-dependent expression of the reporters switches from gfp to lacZ in the LhomieG Z3 insert (Fig. 6F and G). This would not be expected in the first version of the loop-extrusion model, but it is predicted in the second version. However, as indicated in the diagram in Fig. 6E, the two reporters, along with the hebe enhancers, are included in the larger eveGa TAD, and for this reason both reporters should be activated equally by both the eve and hebe enhancers. This is not the case. The hebe enhancers are blocked by the intervening homie boundary, while there is little or no eve-dependent gfp expression.

TAD formation by transgenes inserted at–142 kb

We used MicroC to examine the TAD organization of the genomic region extending from upstream of the attP site through the eve locus in 12-16 hr embryos carrying different transgene insertions. At this stage, eve is expressed in only a small number of cells in the CNS, mesoderm and anal plate, and this is also true for hebe. This means that most of the interactions detected by MicroC are in cells in which the eve gene and the two reporters are inactive. In addition, while the transgene co-localizes with eve in less than about a fifth of blastoderm stage nuclei, the frequency of physical contact is expected to be much higher in 12-16 hr embryos (c.f., Vazquez et al. Vazquez et al., 2006)). This is because at this stage of development, cell cycles are much longer, and homologs are fully paired. The key findings are summarized below.

LlambdaG Z5

While there is no evidence of contact between sequences in eve and sequences around the attP site at –142 kb in wild-type embryos, this not true for embryos carrying the dual reporter transgene. Perhaps the most unexpected finding is a weak, but clearly discernable pattern of interaction linking the LlambdaG Z5 transgene to the eve locus (Fig. 7A). However, since the LlambdaG Z5 transgene does not respond to the eve enhancers, this physical interaction is not sufficient to drive detectable expression of the reporters. Though our experiments do not allow us to unequivocally identify which sequences in the transgene are responsible for this long-distance interaction, the most likely candidates are the two eve promoters in the LlambdaG Z5 transgene. As noted above, interaction triangles are observed within the eve TAD spanning the regions between nhomie and homie (and/or the eve 3’ PRE) and the eve promoter. It seems possible that the factors supporting these intra-TAD interactions are also responsible for connecting LlambdaG Z5 to the eve locus. In this respect it is interesting to note that homie and nhomie stimulate enhancer-promoter interactions in a transgene assay (Yokoshi et al., 2020).

homie in Z5 transgene LhomieG and GhomieL Z3 mediates long-range interactions with the eve locus

Small dark blue arrows – insulators upstream and downstream of hebe that appear important to demarcate interaction domains between –142kb and eve. Large blue arrow – endogenous nhomie. Large red arrow – homie, either endogenous or in the transgene. The direction of arrows follow established convention on nhomie/homie, and do not reflect orientation of insulator protein binding motifs per se. Gray boxes – enhancers. A) MicroC map of the control line LlambdaG Z5. Scaled cartoons of the two loci of interest are shown directly below the MicroC map, and an unscaled blowup of the elements of interest at each locus is provided. Within the MicroC map of the entire locus, a zoom-in of the off-diagonal interaction between the –142kb insertion cassette and the endogenous eve locus is shown. Note a slight increase in interaction frequency (compare to Figure 1A). B, C) MicroC map of LhomieG Z5 and GhomieL Z5, respectively. The only difference between the two lines is the orientation of transgenic homie, as indicate below each MicroC map. Blow up of the off-diagonal interaction between –142kb and eve, including scaled cartoons denoting features of interest in the top right corner. Note the changed pattern of interaction with the endogenous eve locus due to the orientation switch. D) “Virtual 4C” maps obtained from MicroC maps of “B” (top panel), and “C” (bottom panel). Viewpoints are shown in both panels from either the lacZ gene (orange) or the GFP gene (green). Note in each cassette the altered interaction of either gene with endogenous eve, depending on the change of orientation of homie.

LhomieG Z5

Unlike LlambdaG Z5, the eve-lacZ reporter in the LhomieG Z5 insert is activated by the eve enhancers, and this regulatory interaction is perfectly paralleled by the strong pattern of crosslinking between the transgene and the eve locus (Fig. 7B). On the eve side, the interactions between the transgene and the eve locus generate a heavily populated band of crosslinked sequences that span the 16 kb eve TAD. On the transgene side, there is an unequal distribution of crosslinked sequences to either side of the transgene homie boundary. As shown in the inset, the heaviest density of crosslinked sequences is on the eve-lacZ side of the transgene, consistent with the activation of this reporter by eve enhancers. To confirm that the interaction bias mimics the differences in activity of the two reporters in cells in which the reporters are expressed, we analyzed the interaction profiles from viewpoints within either the eve-lacZ or eve-gfp genes. Fig. 7D shows that eve-lacZ interacts with sequences extending across the eve TAD, while eve-gfp interactions are largely restricted to the eve homie element. While these results would be seemingly be consistent with both the boundary-pairing and loop-extrusion models, there is no evidence of cohesin breaking through multiple intervening boundaries into order to establish the novel eve(Mam) TAD. Instead, the organization of the TADs and LDC domains in the interval in between the transgene and eve (Fig. 7B) resemble that seen in wild-type NC14 embryos (see Fig. 2) (with the caveat that there are significantly fewer reads in the transgene experiment). Equally important, the same is true for the eve TAD. In order to generate the eve(Mam) TAD, the cohesin complex must break through the nhomie boundary and, in so doing, disrupt the eve TAD. However, there are no obvious alterations in either the eve volcano or plume with respect to their structure or relative contact density compared to the LlambdaG Z5 and wild type NC14 embryos. Likewise, there is no evidence of novel vertical and 45° crosslinking stripes that would reflect the boundary breakthrough events that are expected in the loop-extrusion model to accompany the formation of a physical connection between the transgene and the eve locus.

We next used the transgene homie as the viewpoint to test the predictions of the boundary-pairing and loop-extrusion models. Since homie can pair with itself (head-to-head) and with nhomie (head-to-tail) in transvection assays (Fujioka et al., 2016), a prediction of the boundary-pairing model is that there will be physical interactions between the transgene homie and both nhomie and homie in the eve locus (c.f., Vazquez et al., Vazquez et al., 2006)). Also based on what is known about how fly boundaries pair with each other (Chetverina et al., 2014; Chetverina et al., 2017; Kyrchanova et al., 2008a), self-interactions are expected to be stronger than heterologous interactions. In contrast, in the loop-extrusion model, the cohesin complex forms the eveMa TAD by linking homie in the transgene to homie in the eve locus, and this requires the cohesin complex to break through nhomie and disrupt the connections between nhomie and homie in the eve locus. In this case, transgene:homie← →eve:homie interactions should be observed, while transgene:homie← →-eve:nhomie should not. Fig. 8A shows that, as predicted by the boundary-pairing model, the transgene homie interacts with both homie and nhomie in the eve locus, and the cross-linking frequency is greater for homie:homie interactions.

homie in the transgene interacts with eve homie and nhomie. “Virtual 4C” maps of the LhomieG (A) or GhomieL (B) transgenes, for the Z5 (blue) and Z3 (orange) orientations. Viewpoints are taken from the transgenic homie sequence. Importantly, note that transgenic homie preferentially interacts with endogenous homie no matter the orientation. Intriguingly, the orientation switch of transgenic homie leads to an apparent change in its degree of preference for endogenous homie over nhomie.

Since the eve TAD does not appear to be disturbed by the presence of the LhomieG Z5 transgene, these findings would suggest that the transgene homie likely pairs with both nhomie and homie in a pattern somewhat like that illustrated in Fig. 3B. While the interactions with the transgene are inconsistent with the loop-extrusion model, one could make the ad hoc assumption that runaway cohesin complexes form not one but two different novel TADs. One would be eveMa. In the other, eveElephant (eveEl), a cohesin complex that initiates in, for example, the TE TAD would come to a halt on one side at the homie boundary in the transgene, and on the other side at the nhomie boundary in the eve locus (Fig. Supplemental 4). However, this configuration would not explain how the transgene reporter is activated by eve enhancers. Since the eve enhancers in eveEl are located in a different TAD than the TADs containing eve-lacZ and eve-gfp, they would not be able to activate either reporter.

GhomieL Z5

When homie is reversed in the GhomieL Z5 transgene, the eve-gfp reporter responds to the eve enhancers instead of eve-lacZ. In the boundary-pairing model, eve-gfp is activated because the self-and heterologous-pairing interactions between nhomie and homie are orientation-dependent. As a result, the topology of the loop formed between the transgene and eve switches from a stem-loop to a circle-loop, and this brings the eve-gfp reporter in close proximity to the eve enhancers instead of the eve-lacZ reporter. Since circle-loops cannot be formed in the loop-extrusion model, activation of the distal eve-gfp reporter would require the formation of a larger eveGa TAD in order to encompass the reporters in the transgene and the eve enhancers in the same looped domain (Fig. 3E). However, as shown in Fig. 7C, we do not observe the cross-linking pattern expected for a new eveGa TAD that includes the entire transgene at –142 kb and then extends to an even more distal boundary. Nor are there any novel vertical/45° crosslinking stripes or obvious perturbations in the basic TAD organization in the region beyond –142 kb, or between –142 kb and the eve locus, as might be expected for cohesin complexes breaking through multiple intervening roadblocks. Instead, as was the case for LhomieG Z5, the GhomieL Z5 transgene is physically linked to the eve locus, and there is a prominent band of crosslinked transgene sequences that spans the eve TAD (Fig. 7C). Consistent with the expression patterns of the two reporters, the heaviest density of crosslinked sequences is on the eve-gfp side of the transgene (see inset). This is confirmed by the interaction profiles generated using either the eve-gfp or eve-lacZ gene body as the viewpoint: the most frequent contacts are between sequences within eve and the eve-gfp reporter (Fig. 7D). In addition, as was the case for LhomieG Z5, when the transgene homie is used as the viewpoint, interactions are observed between both homie and nhomie in the eve locus, with the most frequent corresponding to self-pairing interactions (Fig. 8B). Other than postulating the ad hoc eveEl TAD (Fig. Supplemental 4), this would not be expected for either version of the loop-extrusion model.

LhomieG Z3 and GhomieL Z3

To confirm the findings for transgenes inserted in the Z5 orientation, we used MicroC to analyze the TAD organization of LhomieG Z3 and GhomieL Z3. (Sup. Figs. 5 and 6). Like their Z5 counterparts, both show a strong band of interaction between the transgene and sequences spanning the eve locus. In the case of GhomieL Z3, which is predicted to form a stem-loop, bringing eve-gfp in contact with the eve enhancers, the eve-gfp sequences interact more frequently with the eve locus (see inset in Fig. Supplemental 5A). homie is in the opposite orientation in the transgene and in the chromosome in LhomieG Z3, and in the boundary pairing model is expected to generate a circle-loop. While the sample preparation for MicroC was of poorer quality than the others, the inset in Sup. Fig.6A shows that, as expected, eve-lacZ interacts more frequently with sequences in the eve locus than eve-gfp. On the other hand, there is no evidence of a larger evaGa TAD even though eve-lacZ is activated by the eve enhancers. These interactions are confirmed when either eve-lacZ or eve-GFP are used as viewpoints to plot the physical contacts between the eve TAD and the two reporters in LhomieG Z3 and GhomieL Z3 (Sup. Figs. 5B and 6B). In addition, as was the case for the two Z5 transgenes, when the transgene homie is used as the viewpoint, interactions are observed between homie in the transgene and both homie and nhomie in the eve locus (Fig. 8A and B). Again, the most frequent interactions are self-pairing interactions between homie in the transgene and the eve homie.

Loop topology and interactions with neighbors

The results in the previous sections demonstrate that the topology of the loop formed between the transgene and the eve locus depends upon the orientation of homie in the transgene and the orientation of the transgene in the chromosome. When homie in the transgene is pointing away from the eve locus, the topology is a stem-loop. When homie in the transgene is pointing towards the eve locus, the topology is a circle-loop.

Changing the topology also impacts how sequences flanking the transgene and flanking eve interact with each other. With the important caveat that the depth of our MicroC sequencing is limited, the interaction profiles reflect the topology of the transgene-induced loop. For homie pointing away from eve (LhomieG Z5 and GhomieL Z3), sequences to the TER94 side of eve and to the hebe enhancer side of the transgene are more frequently crosslinked with each other (Fig. Supplemental 7). These interactions would be expected for a stem-loop structure (Fig. 3B). For homie pointing towards eve, sequences on the TER94 side of eve interact with sequences on the eve side of the transgene. This would be expected for a circle-loop structure (Fig. 3C).

Discussion

Two different mechanisms have been proposed to explain how TADs are formed and their endpoints determined. In the first, a cohesin complex first induces a small loop at a loading site within the TAD-to-be and then progressively extrudes one or both strands until the complex encounters boundary roadblocks. If the loading site is located asymmetrically within the TAD-to-be, extrusion will continue on the unblocked strand until a roadblock is encountered. The endpoints of the loops were initially thought to be determined by the location of the first two roadblocks encountered by the cohesin complex (Alipour and Marko, 2012). In a refinement of this model, the orientation of a given roadblock relative to the direction of movement of the cohesin complex determines whether it will come to a halt (Davidson and Peters, 2021; Fudenberg et al., 2016; Guo et al., 2012; Guo et al., 2015; Sanborn et al., 2015). It has also been suggested that cohesin complexes can break through a roadblock and come to a halt at more distant roadblocks, so that in a population of cells analyzed by Hi-C or MicroC, the basic units of organization, the TADs, are overlaid by a hierarchy of LDC domains (Hsieh et al., 2020; Krietenstein et al., 2020). However, the only loop topologies that are possible in the loop-extrusion model are a stem-loop and an unanchored loop.

In the other mechanism, TAD formation is governed by boundary:boundary pairing interactions. In this case, the critical factors are the pairing and orientation preferences of potential partners, together with proximity. Since boundary pairing interactions often exhibit an orientation preference, the topology of the loop formed by paired boundaries can be either a stem-loop or a circle-loop, depending on their relative orientation in the chromosome. In this respect it differs from the loop-extrusion model in that switching the orientation of the boundary does not preclude the formation of a TAD, whereas it can in some versions of the loop-extrusion model.

In the studies reported here, we have used MicroC to analyze the TAD organization of a ∼150 kb chromosomal segment flanking the eve locus, and have experimentally tested the predictions of these two models for the mechanisms of TAD formation. Our studies do not support a loop-extrusion mechanism, and are inconsistent with this model on multiple levels.

Manipulating “TAD” formation

In our experimental paradigm, a transgene containing two divergently transcribed reporters and the eve boundary homie is inserted 142 kb upstream of the eve TAD. Depending upon the orientation of the homie boundary within the transgene, either eve-gfp or eve-lacZ is activated by developmentally regulated enhancers in the eve TAD. Since there are a series of well-defined TADs and their associated boundary elements separating the transgene from eve, a version of the loop-extrusion model in which the cohesin complex invariably arrests upon encountering nhomie and homie at the ends of the eve TAD cannot account for this long-distance regulation. Instead, the extruding cohesin complex must be able to break through the boundaries located between homie in the transgene and homie in the eve TAD. This would generate a novel TAD, eveMa, with endpoints corresponding to the two homie boundaries (Fig. 3D). In this case, the reporter on the eve side of the transgene homie would be in the same domain as the eve enhancers, and thus potentially subject to regulation. While eveMa could account for reporter expression in inserts in which the transgene homie is pointing away from eve (and also away from the reporter on the eve side of the transgene insert), it does not explain reporter expression when the transgene homie is pointing towards eve (and in this case towards the reporter on the eve side of the transgene insert). In the latter case, the reporter activated by the eve enhancers lies outside of the eveMa (cf., Fig. 5D), and thus is not included in the same TAD as the eve enhancers. In order to account for activation of the distal reporter, one must imagine that the cohesin complex breaks through the homie boundary in the transgene and continues on until extrusion is halted by an even more distal boundary. In this revision of the loop-extrusion model, both reporters would be included in the same TAD, eveGa (c.f., Fig. 5E), as the eve enhancers, and both reporters should be activated. However, this is not observed.

TAD organization: loop-extrusion versus boundary pairing

Analysis of the TAD organization in the region spanning the transgene insertion site and eve also does not fit with the expectations of the loop-extrusion model. To account for reporter activation, the cohesin complex must be able to break through the nhomie boundary. This breakthrough could be engineered by a cohesin complex that initiated loop extrusion either from within the eve TAD or from one of the TADs located between the nhomie boundary and the transgene at –142 kb. However, in the absence of the transgene, there is no indication that nhomie is particularly prone to break-through events. In fact, the eve TAD differs from most of the TADs in between eve and the transgene in that it is incorporated into larger LDC interaction triangles much less frequently. There is also no reason to imagine that the presence of a homie (or nhomie: see Fujioka et al., 2016)) transgene at –142 kb would somehow induce nhomie break-through events.

As expected from the regulatory interactions, homie-containing transgenes at –142 kb generate a strong band of crosslinking events that physically link the transgene to the eve locus. In the loop-extrusion model, this strong band of interaction should generate a novel interaction triangle (LDC or HDIC) with endpoints corresponding to the homie elements in the transgene and eve. Contrary to this expectation, a triangle of increased interaction frequencies spanning the entire region between –142 kb and homie is not observed for any of the transgene inserts. In viewpoints from the homie element in the LhomieG Z5 and GhomieL Z3 transgenes, there are two peaks in the eve TAD. The more prominent peak maps to homie, while the less prominent peak maps to nhomie. To account for these interaction peaks, the loop-extrusion model would have to assume that the presence of the LhomieG Z5 and GhomieL Z3 transgenes induces the formation of two novel TADs, the eveMa (homie:homie) TAD in Fig. 5 and another TAD, eveElephant, that links the transgene homie to the nhomie element in the eve locus (Fig. Supplemental 4). However, while the eveMa TAD could potentially explain the activation of reporters on the proximal side of the transgene homie by the eve enhancers, these TADs do not explain how the eve enhancers are able to activate reporters on the distal (hebe enhancer) side of the GhomieL Z5 and LhomieG Z3 transgene. This would require the formation of yet another TAD, eveGa, that encompasses eve, the transgene, and one or more TADs beyond the hebe gene (Fig. 3E). However, there is no indication of physical interactions between the homie boundary in eve and a boundary located beyond the hebe gene. Moreover, like eveMa, eveGa would also have to be accompanied by the formation of eveEl, since the homie element in GhomieL Z5 and LhomieG Z3 interacts with both nhomie and homie.

While these results are inconsistent with the expectations of the loop-extrusion model, they dovetail nicely with many of the predictions of the boundary-pairing model. First, in all four transgenes, homie interacts with both homie and nhomie in the eve locus, but does not interact with the multiple boundary elements located in between. This is consistent with other studies which show that boundaries often have strong partner preferences (Gohl et al., 2011; Kyrchanova et al., 2011; Kyrchanova et al., 2008b). Second, interaction frequencies are greater for self-pairing than for heterologous pairing. This fits with what is known about how partner preferences are determined (Blanton et al., 2003; Chetverina et al., 2017; Erokhin et al., 2021; Kyrchanova et al., 2008a; Zolotarev et al., 2016). Third, the pattern of reporter activation in the four transgenes is precisely as expected from previous studies on the orientation dependence of homie pairing with itself (head-to-head) and with nhomie (head-to-tail) (Chen et al., 2018; Fujioka et al., 2016). This orientation dependence means that the reporter located “upstream” of homie is the one that is activated, independent of the orientation of the transgene in the chromosome. Fourth, the pattern of physical interactions between the eve locus and the transgenes recapitulates the pattern of reporter activation. That is, when eve-lacZ sequences are preferentially crosslinked to sequences in the eve locus, then the eve-lacZ reporter is activated by the eve enhancers. Likewise, eve-gfp is preferentially crosslinked to sequences in eve when it is activated by the eve enhancers. Again this is dependent on the orientation of homie in the transgene, but not on the orientation of the transgene in the chromosome. Fifth, unlike the loop-extrusion model where only stem-loops or disconnected loops can be generated, both stem-loops and circle-loops can be, and, as we have shown here, actually are generated by boundary:boundary pairing. The topology of the loop formed between the transgene and eve depends on the orientation of the transgene homie in the chromosome. When homie in the transgene insert is pointing away from the eve locus, a stem-loop is generated. When homie in the transgene insert is pointing towards the eve locus, a circle-loop is generated. That both loop topologies can be generated by boundary:boundary pairing is confirmed by not only by the pattern of expression of the two reporters in the four transgene inserts but also by their pattern of physical interactions evident in our MicroC experiments.

Are the pairing interactions of homie and nhomie unusual?

There are two reasons to think that the properties of most fly boundaries are similar to those of homie and nhomie. First, all of the fly boundaries that have been tested in assays that are expected to require pairing do have this activity (Blanton et al., 2003; Cai and Shen, 2001; Erokhin et al., 2011; Gohl et al., 2011; Kyrchanova et al., 2008a; Kyrchanova et al., 2011; Kyrchanova et al., 2007; Kyrchanova et al., 2008b; Muravyova et al., 2001). Second, multimerized binding sites for generic polydactyl zinc finger proteins like Su(Hw), Pita, dCTCF, Zw5, and Zipic that are found associated with many fly boundaries are not only able to block enhancer/silencer action but can also mediate pairing interactions in insulator bypass and transvection assays (Erokhin et al., 2021; Kyrchanova et al., 2008a; Zolotarev et al., 2016). Moreover, as is the case for homie and nhomie, pairing interactions require the appropriate partners: multimerized sites for Zw5 do not pair with multimerized sites for dCTCF (Erokhin et al., 2021; Kyrchanova et al., 2008a; Zolotarev et al., 2016). This makes good sense, since these proteins form dimers/multimers, as do other fly boundary factors like BEAF, Mod(mdg4), the BEN-domain protein Insensitive and the Elba complex (Avva and Hart, 2016; Bonchuk et al., 2021; Bonchuk et al., 2011; Bonchuk et al., 2020; Fedotova et al., 2018; Fedotova et al., 2019; Fedotova et al., 2017; Hart et al., 1997). The ability to multimerize means that boundaries that share binding sites for the same protein, i.e., for Pita, Zipic or CTCF, can be physically linked to each other by a single protein complex.

Boundary pairing orientation and loop topology

Our analysis of the TAD organization in the eve neighborhood suggests that the predominant loop topology may be a circle-loop, not a stem-loop. Only the eve TAD (and perhaps the very small TAD, TA) is clearly a stem-loop. It is topped by a distinctive volcano plume and is not overlaid by a hierarchy of LDC domains, indicating that it must be “insulated” from crosslinkable interactions with other nearby TADs. By contrast, none of the TADs in the region spanning eve and the hebe gene have volcano plumes at their apex. Instead, all of these TADs are overlaid by a hierarchal organization of LDC domains (clouds) that diminish in crosslinking intensity as the number of in-between TADs increases. As discussed above, the contact pattern of the TADs to the left of eve are likely to arise from partner switching and/or head-to-head boundary pairing interactions that generate circle-loops. Pairing orientation versus loop topology is considered further in the accompanying paper (Ke et al.).

Orientation and long-distance regulatory interactions

Our experiments highlight an important constraint on long-distance regulatory interactions that are mediated not only by boundary elements like homie and nhomie but also by the GAF-dependent tethering and bypass elements (Batut et al., 2022; Kyrchanova et al., 2023; Kyrchanova et al., 2018; Kyrchanova et al., 2019a; Kyrchanova et al., 2019b; Levo et al., 2022; Li et al., 2023; Mohana et al., 2023), namely the orientation of the physical interactions that bring distant enhancers (or silencers) into close proximity with their gene targets. As we have shown here, the orientation of the physical interactions that bring distant sequences close together determines whether the enhancer will be able to activate its target gene. In principle the effects of orientation could be mitigated by increasing the distance between the boundary/tethering elements and either the enhancers (silencers) or the promoter. However, distance can be a critical if not a rate limiting step in transcriptional activation. For example, live imaging experiments have shown that the frequency and amplitude of transcriptional bursts depend on the distance between the enhancer and target promoter. Nearly continuous bursting and high amplitude transcription was observed when the enhancer was located close the promoter (Fukaya et al., 2016). However, at distances as little as 6 kb, transcription became discontinuous with discrete bursts, while at 9 kb, the bursts were much less frequent, and the amplitude/output was substantially reduced (Yokoshi et al., 2020). If similar distance-dependent reductions in transcriptional efficiency are common, then it would not be possible to overcome the topological constraints on regulatory interactions by simply increasing the distance between the boundary/tethering elements and the enhancers/promoters. While our studies do not directly address how chromosome architecture impacts regulatory interactions in vertebrates, it seems clear that long-distance regulatory interactions in vertebrates (c.f., Tan et al., 2023) will be subject to the same types of topological constraints as those that apply in flies.

eLife Assessment

This valuable work presents elegant experimental data from the Drosophila embryo supporting the notion that interactions among specific loci, called boundary elements, contribute to topologically associated domain (TAD) formation and gene regulation. Although the evidence supporting boundary elements as determinants of 3D structures is compelling, the evidence rejecting loop extrusion is incomplete. This study will be of interest to the nuclear structure community, particularly those using Drosophila as a model.

Reviewer #1 (Public Review):

Summary:

The authors addressed how long-range interactions between boundary elements are established and influence their function in enhancer specificity. Briefly, the authors placed two different reporters separated by a boundary element. They inserted this construct ectopically ~140 kb away from an endogenous locus that contains the same boundary element. The authors used expression patterns driven by nearby enhancers as an output to determine which enhancers the reporters interact with. They complemented this analysis with 3D DNA contact mapping. The authors found that the orientation of the boundary element determined which enhancers each reporter interacted with. They proposed that the 3D interaction topology, whether being circular or stem configuration, distinguished whether the interaction was cohesin mediated or through an independent mechanism termed pairing.

Strengths:

The transgene expression assays are built upon prior knowledge of the enhancer activities. The 3D DNA contacts confirm that transgene expression correlates with the contacts. Using 4 different orientations covers all combinations of the reporter genes and the boundary placement.

Weaknesses:

The interpretation of the data as a refusal of loop extrusion playing a role in TAD formation is not warranted, as the authors did not deplete the loop extruders to show that what they measure is independent. As the authors show, the single long DNA loop mediated by cohesin loop extrusion connecting the ectopic and endogenous boundary is clearly inconsistent with the results, therefore the main conclusion of the paper that the 3D topology of the boundary elements a consequence of pairing is strong. However, the loop extrusion and pairing are not mutually exclusive models for the formation of TADs. Loop-extruding cohesin complexes need not make a 140 kb loop, multiple smaller loops could bring together the two boundary elements, which are then held together by pairing proteins that can make circular topologies.

Reviewer #2 (Public Review):

In Bing et al, the authors analyze micro-C data from NC14 fly embryos, focusing on the eve locus, to assess different models of chromatin looping. They conclude that fly TADs are less consistent with conventional cohesin-based loop extrusion models and instead rely more heavily on boundary-boundary pairings in an orientation-dependent manner.

Overall, I found the manuscript to be interesting and thought-provoking. However, this paper reads much more like a perspective than a research article. I strongly suggest the authors spend some time editing their introduction to the most salient points as well as organizing their results section in a more conventional way with conclusion-based titles. It was very difficult to follow the authors' logic throughout the manuscript as written. It was also not clear as written which experiments were performed as part of this study and which were reanalyzed but published elsewhere. This should be made clearer throughout.

It has been shown several times that Drosophila Hi-C maps do not contain all of the features (frequent corner peaks, stripes, etc.) observed when compared to mammalian cells. Considering these features are thought to be products of extrusion events, it is not an entirely new concept that Drosophila domains form via mechanisms other than extrusion. That being said, the authors' analyses do not distinguish between the formation and the maintenance of domains. It is not clear to this reviewer why a single mechanism should explain the formation of the complex structures observed in static Hi-C heatmaps from a population of cells at a single developmental time point. For example, how can the authors rule out that extrusion initially provides the necessary proximity and possibly the cis preference of contacts required for boundary-boundary pairing whereas the latter may more reflect the structures observed at maintenance? Future work aimed at analyzing micro-C data in cohesin-depleted cells might shed additional light on this.

Additional mechanisms at play include compartment-level interactions driven by chromatin states. Indeed, in mammalian cells, these interactions often manifest as a "plume" on Hi-C maps similar to what the authors attribute to boundary interactions in this manuscript. How do the chromatin states in the neighboring domains of the eve locus impact the model if at all?

How does intrachromosomal homolog pairing impact the models proposed in this manuscript (Abed et al. 2019; Erceg et al., 2019). Several papers recently have shown that somatic homolog pairing is not uniform and shows significant variation across the genome with evidence for both tight pairing regions and loose pairing regions. Might loose pairing interactions have the capacity to alter the cis configuration of the eve locus?

In summary, the transgenic experiments are extensive and elegant and fully support the authors' models. However, in my opinion, they do not completely rule out additional models at play, including extrusion-based mechanisms. Indeed, my major issue is the limited conceptual advance in this manuscript. The authors essentially repeat many of their previous work and analyses. The authors make no attempt to dissect the mechanism of this process by modifying extrusion components directly. Some discussion of Rollins et al., 1999 on the discovery of Nipped-B and its role in enhancer-promoter communication should also be made to reconcile their conclusions in the proposed absence of extrusion events.

Reviewer #3 (Public Review):

Bing et al. attempt to address fundamental mechanisms of TAD formation in Drosophila by analyzing gene expression and 3D conformation within the vicinity of the eve TAD after insertion of a transgene harboring a Homie insulator sequence 142 kb away in different orientations. These transgenes along with spatial gene expression analysis were previously published in Fujioka et al. 2016, and the underlying interpretations regarding resulting DNA configuration in this genomic region were also previously published. This manuscript repeats the expression analysis using smFISH probes in order to achieve more quantitative analysis, but the main results are the same as previously published. The only new data are the Micro-C and an additional modeling/analysis of what they refer to as the 'Z3' orientation of the transgenes. The rest of the manuscript merely synthesizes further interpretation with the goal of addressing whether loop extrusion may be occurring or if boundary:boundary pairing without loop extrusion is responsible for TAD formation. The authors conclude that their results are more consistent with boundary:boundary pairing and not loop extrusion; however, most of this imaging data seems to support both loop extrusion and the boundary:boundary models. This manuscript lacks support, especially new data, for its conclusions. Furthermore, there are many parts of the manuscript that are difficult to follow. There are some minor errors in the labelling of the figures that if fixed would help elevate understanding. Lastly, there are several major points that if elaborated on, would potentially be helpful for the clarity of the manuscript.

Major Points:

1. The authors suggest and attempt to visualize in the supplemental figures, that loop extrusion mechanisms would appear during crosslinking and show as vertical stripes in the micro-C data. In order to see stripes, a majority of the nuclei would need to undergo loop extrusion at the same rate, starting from exactly the same spots, and the loops would also have to be released and restarted at the same rate. If these patterns truly result from loop extrusion, the authors should provide experimental evidence from another organism undergoing loop extrusion.

2. On lines 311-314, the authors discuss that stem-loops generated by cohesin extrusion would possibly be expected to have more next-next-door neighbor contacts than next-door neighbor contacts and site their models in Figure 1. Based on the boundary:boundary pairing models in the same figure would the stem-loops created by head-to-tail pairing also have the same phenotype? Making possible enrichment of next-next-door neighbor contacts possible in both situations? The concepts in the text are not clear, and the diagrams are not well-labeled relative to the two models.

3. The authors appear to cite Chen et al., 2018 as a reference for the location of these transgenes being 700nM away in a majority of the nuclei. However, the exact transgenes in this manuscript do not appear to have been measured for distance. The authors could do this experiment and include expression measurements.

4. The authors discuss the possible importance of CTCF orientation in forming the roadblock to cohesin extrusion and discuss that Homie orientation in the transgene may impact Homie function as an effective roadblock. However, the Homie region inserted in the transgene does not contain the CTCF motif. Can the authors elaborate on why they feel the orientation of Homie is important in its ability to function as a roadblock if the CTCF motif is not present? Trans-acting factors responsible for Homie function have not been identified and this point is not discussed in the manuscript.

5. The imaging results seem to be consistent with both boundary:boundary interaction and loop extrusion stem looping.

6. The authors suggest that the eveMa TAD could only be formed by extrusion after the breakthrough of Nhomie and several other roadblocks. Additionally, the overall long-range interactions with Nhomie appear to be less than the interactions with endogenous Homie (Figures 7, 8, and supplemental 5). Is it possible that in some cases boundary:boundary pairing is occurring between only the transgenic Homie and endogenous Homie and not including Nhomie?

7. In Figure 4E, the GFP hebe expression shown in the LhomieG Z5 transgenic embryo does not appear in the same locations as the LlambdaG Z5 control. Is this actually hebe expression or just a background signal?

8. Figure 6- The LhomieG Z3 late-stage embryo appears to be showing the ventral orientation of the embryo rather than the lateral side of the embryo as was shown in the previous figure. Is this for a reason? Additionally, there are no statistics shown for the Z3 transgenic images. Were these images analyzed in the same way as the Z5 line images?

9. Do the Micro-C data align with the developmental time points used in the smFISH probe assays?

Author response:

The following is the authors’ response to the original reviews.

Reviewer #1 (Public Review):

Summary:

The authors addressed how long-range interactions between boundary elements are established and influence their function in enhancer specificity. Briefly, the authors placed two different reporters separated by a boundary element. They inserted this construct ectopically ~140 kb away from an endogenous locus that contains the same boundary element. The authors used expression patterns driven by nearby enhancers as an output to determine which enhancers the reporters interact with. They complemented this analysis with 3D DNA contact mapping. The authors found that the orientation of the boundary element determined which enhancers each reporter interacted with. They proposed that the 3D interaction topology, whether being circular or stem configuration, distinguished whether the interaction was cohesin mediated or through an independent mechanism termed pairing.

Strengths:

The transgene expression assays are built upon prior knowledge of the enhancer activities. The 3D DNA contacts confirm that transgene expression correlates with the contacts. Using 4 different orientations covers all combinations of the reporter genes and the boundary placement.

Weaknesses:

The interpretation of the data as a refusal of loop extrusion playing a role in TAD formation is not warranted, as the authors did not deplete the loop extruders to show that what they measure is independent.

(1.1) To begin with, our findings do not exclude the possibility that cohesin loop extrusion has some sort of role in the formation or maintenance of TADs in flies or other aspects of chromosome structure. On the other hand, it clearly is not determinative in defining the end-points of TADs or in generating the resulting topology (stem-loop or circle-loop). Our main point, which we feel we have established unequivocally, is that it can’t explain many essential features of TADs or chromosome loops (see below) in Drosophila. This reviewer agrees with this point in their next paragraph (below). We also think that the loop extrusion model’s general acceptance as THE driving force behind TAD formation in mammals is unwarranted and not fully consistent with the available data, as explained below.

As to the reviewer’s specific point regarding depletion of loop extruders, we first note that completely eliminating factors encoding cohesin subunits in fly embryos isn’t readily feasible. As cohesin is essential starting at the beginning of embryonic development, and is maternally deposited, knockdowns/depletions would likely be incomplete and there would always be some remaining activity. As long as there is some residual activity—and no disruption in TAD formation is observed—this experimental test would be a failure. In addition, any defects that are observed might arise not from a failure in TAD formation via loop extrusion but rather because the rapid mitotic cycles would be disrupted. A far better approach would be to deplete/knockdown cohesin subunits in tissue culture cells, as there is no requirement for the cells to undergo embryonic development. Moreover, since cell division is relatively slow, the depletion would likely eliminate much if not all of the activity before a checkpoint is reached.

While a drastic depletion of cohesin is not feasible in our model organism, we would draw the reviewer’s attention to an experiment of this type which has already been done in mammalian tissue culture cells by Goel et al. (Goel et al. 2023). Unlike most Hi-C studies in mammals, the authors used region capture MicroC (RCMC). In contrast to published genome-wide mammalian MicroC experiments (c.f., (Hsieh et al. 2020; Krietenstein et al. 2020)) which require large bin sizes to visualize mammalian “TADs,” the resolution of the experiments in Goel et al. (Goel et al. 2023) is similar to the resolution in our MicroC experiments (200-400 bp). A MicroC contact map from Goel et al. shows the Pdm1g locus on chromosome 5 before and after Rad21 depletion. The contact map visualizes a 250 kb DNA segment, which is only slightly larger than the ~230 kb DNA segment in Fig. 2C in our paper.

In this experiment, there was a 97% reduction in the amount of Rad21. However, as can be seen by comparing the contact profiles above and below the diagonal, there is little or no difference in TAD organization after cohesin depletion when individual TADs are visualized with a bin size of 250 bp. These results would indicate that mammalian TADs do not require cohesin.

Note also that the weak 45o stripes connecting different TADs (c.f. blue/green arrowheads) are still present after Rad21 depletion. In the most popular version of the loop extrusion model, cohesin loads at a site(s) somewhere in the TAD-to-be, and then extrudes both strands until it bumps into CTCF roadblocks. As illustrated in Figure Sup 2, this mechanism generates a vertical stripe originating at the cohesin loading site and extending until cohesin bumps into the left or right roadblock, at which point the stripe transitions into 45o stripe that ends when cohesin bumps into the other roadblock. While 45o stripes are visible, there is no hint of a vertical stripe. This suggests that the mechanism for generating stripes, if it is an active mechanism (rather than passive diffusion) may be quite different. The 45o stripes must be generated by a factor(s) that is anchored to one (blue arrowhead) or both (green arrowhead) boundaries. In addition, this factor, whatever it is, is not cohesin. The reason for this is that the 45o stripes are present both before and after Rad21 depletion. Moreover, if one were to imagine that the stripes represent a process involved in TAD formation, this process does not require cohesin (see Goel et al 2023).

It is worth noting another observation that is inconsistent with the cohesin loop extrusion/CTCF roadblock model for TAD formation/maintenance. CTCF is not found at all of the TAD boundaries in this 250 kb DNA region. This would suggest that there are other DNA binding proteins that have chromosomal architectural functions besides CTCF. In flies, many of the chromosomal architectural proteins are, like CTCF, polydactyl zinc finger (PZF) proteins (Bonchuk et al. 2021; Bonchuk et al. 2022; Fedotova et al. 2017). These include Su(Hw), CTCF, Pita, Zipic and CLAMP. The PZF family in flies is quite large. There are ~250 different PZF genes, and since only a handful of these have been characterized, it seems likely that additional members of this family will have architectural functions. Thus far, only one boundary protein, CTCF, has received attention in studies on mammalian chromosome architecture. As the mammalian genome is much larger and more complicated than the fly genome, it is difficult to believe that CTCF is the sole chromosomal architectural protein in mammals. In this respect, it is worth noting that there are ~800 members of the PZF family in mammalian genomes (Fedotova et al. 2017).

Goel et al. (Goel et al. 2023) did observe alterations in the contact profiles after Rad21 depletion when they visualized the Ppm1g region at much lower resolution (bin sizes of 5 kb and 1 kb). The 5 kb bin size visualizes a region of ~1.2 Mb, while the 1 kb bin size visualizes a region that spans ~800 kb. These large triangular units do not correspond to the individual TADs seen when Goel et al. visualized the Ppm1g locus at 250 bp resolution.

Nor do they correspond to TADs in Fig. 2 of our paper. Instead they represent TAD neighborhoods which, likely consist of 20-30 or more individual TADs. Consequently the alterations in contact patterns seen after Rad21 depletion are occurring at the level of TAD neighborhoods. This can be seen by comparing pixel density inside the blue lines before (above the diagonal) and after Rad21 depletion (below the diagonal) (Goel et al 2023). The more distant contacts between individual TADs within this neighborhood are preferentially reduced by Rad21 depletion (the region below and to the left of the double arrowhead). By contrast, the TADs themselves are unaffected, as are contacts between individual TADs and their immediate neighbors (see purple and light green asterisk). The other interesting feature is the loss of contacts between what appears to be partially overlapping neighborhoods. This loss of neighborhood-toneighborhood contacts can be seen in the region located between the green and blue lines. The neighborhood that appears to partially overlap the Ppm1g neighborhood is outlined in purple.

It worth noting that, with the exception of the high resolution experiments in Goel et al., all of the other studies on cohesin (and CTCF) have examined the effects on contact maps within (and between) large neighborhoods (bin sizes >1 kb). In most cases, these large neighborhoods are likely to be composed of many individual TADs like those seen in Goel et al. and in Fig. 2 of our paper. We also observe larger neighborhoods in the fly genome, though they do not appear to be as large as those in mammals. Our experiments do not address what role cohesin might have in facilitating contacts between more distant TADs located within the same neighborhoods, or between TADs in different neighborhoods, or whether loop extrusion is involved.

We would also note that the Drosophila DNA segment in Fig. 2C contains 35 different genes, while the mammalian DNA segment shown in Fig. 1 has only 9. Thus, in this part of the fly genome, Pol II genes are more densely packed than in the mammalian DNA segment. Much of the fly genome is also densely packed, and the size of individual TADs will likely be smaller, on average, than in mammals. Nevertheless, the MicroC profiles are not all that different. As is also common in flies, each TAD in the Ppm1g region only encompasses one or two genes. Note also that there are no volcano triangles with plumes as would be predicted for TADs that have a stem-loop topology.

In fact, as shown in Author response image 1, the high-resolution contact profile for the Ppm1g region shows a strong resemblance to that observed for the fly Abd-B regulatory domains. These regulatory domains are part of larger neighborhood that encompasses the abd-A and Abd-B genes and their regulatory domains.

Abd-B regulatory domains

As the authors show, the single long DNA loop mediated by cohesin loop extrusion connecting the ectopic and endogenous boundary is clearly inconsistent with the results, therefore the main conclusion of the paper that the 3D topology of the boundary elements a consequence of pairing is strong. However, the loop extrusion and pairing are not mutually exclusive models for the formation of TADs. Loop-extruding cohesin complexes need not make a 140 kb loop, multiple smaller loops could bring together the two boundary elements, which are then held together by pairing proteins that can make circular topologies.

(1.2) In the pairing model, distant boundaries bump into each other (by random walks or partially constrained walks), and if they are “compatible” they pair with each other, typically in an orientation-dependent manner. As an alternative, the reviewer argues that cohesin need not make one large 140 kb loop. Instead it could generate a series of smaller loops (presumably corresponding to the intervening TADs). These smaller loops would bring homie in the transgene in close proximity to the eve locus so that it could interact with the endogenous homie and nhomie elements in the appropriate orientation, and in this way only one of the reporters would be ultimately activated.

There are two problems with the idea that cohesin-dependent loop extrusion brings transgene homie into contact with homie/nhomie in the eve locus by generating a series of small loops (TADs). The first is the very large distances over which specific boundary:boundary pairing interactions can occur. The second is that boundary:boundary pairing interactions can take place not only in cis, but also in trans.

We illustrate these points with several examples.

Fujioka et al. 2016, Fig 7 shows an experiment in which attP sites located ~2 Mb apart were used to insert two different transgenes, one containing a lacZ reporter and the other containing the eve anal plate enhancer (AP) (Fujioka et al. 2016). If the lacZ reporter and the AP transgenes also contain homie, the AP enhancer can activate lacZ expression (panel A,). On the other hand, if one of the transgenes has lambda DNA instead of homie, no regulatory interactions are observed (panel A,). In addition, as is the case in our experiments using the -142 kb platform, orientation matters. In the combination on the top left, the homie boundary is pointing away from both the lacZ reporter and the AP enhancer. Since homie pairs with itself head-tohead, pairing brings the AP enhancer into contact with the lacZ reporter. A different result is obtained for the transgene pair in panel A on the top right. In this combination, homie is pointing away from the lacZ reporter, while it is pointing towards the AP enhancer. As a consequence, the reporter and enhancer are located on opposite sides of the paired homie boundaries, and in this configuration they are unable to interact with each other.

On the top left of panel B, the homie element in the AP enhancer transgene was replaced by a nhomie boundary oriented so that it is pointing towards the enhancer. Pairing of homie and nhomie head-to-tail brings the AP enhancer in the nhomie transgene into contact with the lacZ reporter in the homie transgene, and it activates reporter expression. Finally, like homie, nhomie pairs with itself head-to-head, and when the nhomie boundaries are pointing towards both the AP reporter and the lacZ reporter, reporter expression is turned on.

Long distance boundary-dependent pairing interactions by the bithorax complex Mcp boundary have also been reported in several papers. Fig. 6 from Muller et al. (Muller et al. 1999) shows the pattern of regulatory interactions (in this case PRE-dependent “pairing-sensitive silencing”) between transgenes that have a mini-white reporter, the Mcp and scs’ boundaries and a PRE that is located close to Mcp. In this experiment flies carrying transgenes inserted at the indicated sites on the left and right arms of the 3rd chromosome were mated in pairwise combinations, and their trans-heterozygous progeny examined for pairing-sensitive silencing of the mini-white reporter.

Two examples of long-distance pairing-sensitive silencing mediated by Mcp/scs’ are shown in Fig. 5b from Muller et al. 1999. The transgene inserts in panel A are w#12.43 and ff#10.5. w#12.43 is inserted close to the telomere of 3R at 99B. ff10.5 is inserted closer to the middle of 3R at 91A. The estimate distance between them is 11.3 Mb. The transgene inserts in panel B are ff#10.5 and ff#11.102. ff#11.102 is inserted at 84D, and the distance between them is 11 Mb. Normally, the eye color phenotype of the mini-white reporter is additive: homozygyous inserts have twice as dark eye color as hemizygous inserts, while in _trans-_heterozygous flies the eye color would be the sum of the two different transgenes. However, when a PRE is present and the transgene can pair, silencing is observed. In panel A, the t_rans-_heterozygous combination has a lighter eye color than either of the parents. In panel B, the _trans-_heterozygous combination is darker than one of the parents (ff#10.5) but much lighter than the other (ff#11.102).

All ten of the transgenes tested were able to engage in long distance (>Mbs) _trans_regulatory interactions; however, likely because of how the chromosome folds on the Mb scale (e.g., the location of meta-loops: see #2.1 and Author response image 3) not all of the possible pairwise silencing interactions are observed. The silencing interactions shown in Muller et.al. are between transgenes inserted on different homologs. Mcp/scs'-dependent silencing interactions can also occur in cis. Moreover, just like the homie and nhomie experiments described above, Muller et.al. (Muller et al. 1999) found that Mcp could mediate long-distance activation of mini-white and yellow by their respective enhancers.

The pairing-sensitive activity of the PRE associated with the Mcp boundary is further enhanced when the mini-white transgene has the scs boundary in addition to Mcp and scs’. In the experiment shown in Fig. 8 from Muller et al. 1999, the pairing-sensitive silencing interactions of the Mcp/scs’/scs transgene are between transgenes inserted on different chromosomes. Panel A shows pairing-sensitive silencing between w#15.60, which is on the X chromosome, and w#15.102, which is on the 2nd chromosome. Panel B shows pairing-sensitive silencing between the 2nd chromosome insert w#15.60 and a transgene, w#15.48, which is inserted on the 3rd chromosome.

The long-distance trans and cis interactions described here are not unique to homie, nhomie, Mcp, scs’, or scs. Precisely analogous results have been reported by Sigrist and Pirrotta (Sigrist and Pirrotta 1997) for the gypsy boundary when the bxd PRE was included in the mini-white transgene. Also like the Mcp-containing transgenes in Muller et al. (Muller et al. 1999), Sigrist and Pirrotta observed pairing-sensitive silencing between _gypsy bxd_PRE mini-white transgenes inserted on different chromosomes. Similar long-distance (Mb) interactions have been reported for Fab-7 (Bantignies et al. 2003; Li et al. 2011). In addition, there are examples of “naturally occurring” long-distance regulatory and/or physical interactions. One would be the regulatory/physical interactions between the p53 enhancer upstream of reaper and Xrp1 which was described by Link et al. (Link et al. 2013). Another would be the nearly 60 meta-loops identified by Mohana et al. (Mohana et al. 2023).

Like homie at -142 kb, the regulatory interactions (pairing-sensitive silencing and enhancer activation of reporters) reported in Muller et al. (Muller et al. 1999) involve direct physical interactions between the transgenes. Vazquez et al. (Vazquez et al. 2006) used the lacI/lacO system to visualize contacts between distant scs/Mcp/scs’-containing transgenes in imaginal discs. As indicated in Vasquez et al. 2006, Table 3 lines #4-7, when both transgenes have Mcp and were inserted on the same chromosome, they colocalized in _trans-_heterozygotes (single dot) in 94% to 97% of the disc nuclei in the four pairwise combinations they tested. When the transgenes both lacked Mcp (Vasquez et al. 2006, Table 3 #1), co-localization was observed in 4% of the nuclei. When scs/Mcp/scs’-containing transgenes on the 2nd and 3rd chromosome were combined (Vasquez et al. 2006, Table 3 #8), colocalization was observed in 96% of the nuclei. They also showed that four different scs/Mcp/scs’ transgenes (two at the same insertion site but on different homologs, and two at different sites on different homologs) co-localized in 94% of the eye imaginal disc nuclei (Vasquez et al. 2006, Table 3 #9). These pairing interactions were also found to be stable over several hours. Similar co-localization experiments together with 3C were reported by Li et al. (Li et al. 2011).

The de novo establishment of trans interactions between compatible boundary elements has been studied by Lim et al. (Lim et al. 2018). These authors visualized transvection (enhancer activation of a MS2 loop reporter in trans) mediated by the gypsy insulator, homie and Fab-8 in NC14 embryos. When both transgenes shared the same boundary element, transvection/physical pairing was observed in a small subset of embryos. The interactions took place after a delay and increased in frequency as the embryo progressed into NC14. As expected, transvection was specific: it was not observed when the transgenes had different boundaries. For homie it was also orientation-dependent. It was observed when homie was orientated in the same direction in both transgenes, but not when homie was orientated in opposite directions in the two transgenes.

While one could imagine that loop extrusion-dependent compaction of the chromatin located between eve and the transgene at -142 kb into a series of small loops (the intervening TADs) might be able to bring homie in the transgene close to homie/nhomie in the eve locus, there is no cohesinbased loop extrusion scenario that would bring transgenes inserted at sites 6 Mb, 11 Mb, on different sides of the centromere, or at opposite ends of the 3rd chromosome together so that the distant boundaries recognize their partners and physically pair with each other. Nor is there a plausible cohesin-based loop extrusion mechanism that could account for the fact that most of the documented long-distance interactions involve transgenes inserted on different homologs. This is not to mention the fact that long-distance interactions are also observed between boundarycontaining transgenes inserted on different chromosomes.

In fact, given these results, one would logically come to precisely the opposite conclusion. If boundary elements inserted Mbs apart, on different homologs and on different chromosomes can find each other and physically pair, it would be reasonable to think that the same mechanism (likely random collisions) is entirely sufficient when they are only 142 kb apart.

Yet another reason to doubt the involvement or need for cohesin-dependent loop extrusion in bringing the transgene homie in contact with the eve locus comes from the studies of Goel et al. (Goel et al. 2023). They show that cohesin has no role in the formation of TADs in mammalian tissue culture cells. So if TADs in mammals aren’t dependent on cohesin, there would not be a good reason to think at this point that the loops (TADs) that are located between eve and the transgene are generated by, or even strongly dependent on, cohesin-dependent loop extrusion.

It is also important to note that even if loop-extrusion were to contribute to chromatin compaction in this context and make the looping interactions that lead to orientation-specific pairing more efficient, the role of loop extrusion in this model is not determinative of the outcome, it is merely a general compaction mechanism. This is a far cry from the popular concept of loop extrusion as being THE driving force determining chromosome topology at the TAD level.

Reviewer #2 (Public Review):

In Bing et al, the authors analyze micro-C data from NC14 fly embryos, focusing on the eve locus, to assess different models of chromatin looping. They conclude that fly TADs are less consistent with conventional cohesin-based loop extrusion models and instead rely more heavily on boundaryboundary pairings in an orientation-dependent manner.

Overall, I found the manuscript to be interesting and thought-provoking. However, this paper reads much more like a perspective than a research article. Considering eLIFE is aimed at the general audience, I strongly suggest the authors spend some time editing their introduction to the most salient points as well as organizing their results section in a more conventional way with conclusion-based titles. It was very difficult to follow the authors' logic throughout the manuscript as written. It was also not clear as written which experiments were performed as part of this study and which were reanalyzed but published elsewhere. This should be made clearer throughout.

It has been shown several times that Drosophila Hi-C maps do not contain all of the features (frequent corner peaks, stripes, etc.) observed when compared to mammalian cells. Considering these features are thought to be products of extrusion events, it is not an entirely new concept that Drosophila domains form via mechanisms other than extrusion.

(2.1) While there are differences between the Hi-C contact profiles in flies and mammals, these differences likely reflect in large part the bin sizes used to visualize contact profiles. With the exception of Goel et al. (Goel et al. 2023), most of the mammalian Hi-C studies have been low resolution restriction enzyme-based experiments, and required bin sizes of >1 kb or greater to visualize what are labeled as “TADs.” In fact, as shown by experiments in Goel et al., these are not actually TADs, but rather a conglomeration of multiple TADs into a series of TAD neighborhoods. The same is true for the MicroC experiments of Krietenstein et al. and Hsieh et al. on human and mouse tissue culture cells (Hsieh et al. 2020; Krietenstein et al. 2020). This is shown in Author response image 2. In this image, we have compared the MicroC profiles generated from human and mouse tissue culture cells with fly MicroC profiles at different levels of resolution.

For panels A-D, the genomic DNA segments shown are approximately 2.8 Mb, 760 kb, 340 kb, and 190 kb. For panels E-H, the genomic DNA segments shown are approximately 4.7 Mb, 870 kb, 340 kb and 225 kb. For panels I-L, the genomic DNA segments shown are approximately 3 Mb, 550 kb, 290 kb and 175 kb.

As reported for restriction enzyme-based Hi-C experiments, a series of stripes and dots are evident in mammalian MicroC profiles. In the data from Krietenstein et al., two large TAD “neighborhoods” are evident with a bin size of 5 kb, and these are bracketed by 45o stripes (A: black arrows). At 1 kb (panel B), the 45o stripe bordering the neighborhood on the left no longer defines the edge of the neighborhood (blue arrow: panel B), and both stripes become discontinuous (fuzzy dots). At 500 (panel C) and 200 bp (panel D) bin sizes, the stripes largely disappear (black arrows) even though they were the most prominent feature in the TAD landscape with large bin sizes. At 200 bp, the actual TADs (as opposed to the forest) are visible, but weakly populated. There are no stripes, and only one of the TADs has an obvious “dot” (green asterisk: panel C).

Mammalian MicroC profiles different bin sizes.

Large TAD neighborhoods bordered by stripes are also evident in the Hsieh et al. data set in Author response image 2 panels E and F (black arrows in E and F and green arrow in F). At 400 bp resolution (panel G), the narrow stripe in panel F (black arrows) becomes much broader, indicating that it is likely generated by interactions across one or two small TADs that can be discerned at 200 bp resolution. The same is true for the broad stripe indicated by the green arrows in panels F, G and H. This stripe arises from contacts between the TADs indicated by the red bar in panels G and H and the TADs to the other side of the volcano triangle with a plume (blue arrow in panel H). As in flies, we would expect that this volcano triangle topped by a plume corresponds to a stem-loop. However, the resolution is poor at 200 bp, and the profiles of the neighboring TADs are not very distinct.

For the fly data set, stripes can be discerned when analyzed at 800 bp resolution (see arrows in Author response image 3); however, these stripes are flanked by regions of lower contact, and represent TAD-TAD interactions. At 400 bp, smaller neighborhoods can be discerned, and these neighborhoods exhibit a complex pattern of interaction with adjacent neighborhoods. With bin sizes of 200 bp, individual TADs are observed, as are TAD-TAD interactions like those seen near eve. Some of the TADs have dots at their apex, while others do not—much like what is seen in the mammalian MicroC studies.

Mammalian MicroC profiles different bin sizes.

Stripes: As illustrated in Author response image 2 A-D and E-H, the continuous stripes seen in low resolution mammalian studies (>1 kb bins) would appear to arise from binning artefacts. At high resolution where single TADs are visible, the stripes seem to be generated by TAD-TAD interactions, and not by some type of “extrusion” mechanism. This is most clearly seen for the volcano with plume TAD in Author response inage 2 G and H. While stripes in Author response image 2 disappear at high resolution, this is not always true. There are stripes that appear to be “real” in Geol et al. 2023 for the TADs in the Ppm1g region, and in Author response image 1 for the Abd-B regulatory domain TADs. Since the stripes in the Ppm1g region are unaffected by Rad21 depletion, some other mechanism must be involved (c.f. (Shidlovskii et al. 2021)).

Dots: The high resolution images of mammalian MicroC experiments in Author response image 2D and H show that, like Drosophila (Author response image 3L), mammalian TADs don’t always have a “dot” at the apex of the triangle. This is not surprising. In the MicroC procedure, fixed chromatin is digested to mononucleosomes with MNase. Since most TAD boundaries in flies, and presumably also in mammals, are relatively large (150-400 bp) nuclease hypersensitive regions, extensive MNase digestion will typically reduce the boundary element sequences to oligonucleotides.

In flies, the only known sequences (at least to date) that end up giving dots (like those seen in Author response image 1) are bound by a large (>1,000 kd) GAF-containing multiprotein complex called LBC. In the Abd-B region of BX-C, LBC binds to two ~180 bp sequences in Fab-7 (dHS1 and HS3: (Kyrchanova et al. 2018; Wolle et al. 2015), and to the centromere proximal (CP) side of Fab-8. The LBC elements in Fab-7 (dHS1) and Fab-8 (CP) have both blocking and boundary bypass activity (Kyrchanova et al. 2023; Kyrchanova et al. 2019a; Kyrchanova et al. 2019b; Postika et al. 2018). Elsewhere, LBC binds to the bx and bxd PREs in the Ubx regulatory domains, to two PREs upstream of engrailed, to the hsp70 promoter, the histone H3-H4 promoters, and the eve promoter (unpublished data). Based on ChIP signatures, it likely binds to most PREs/tethering elements in the fly genome (Batut et al. 2022; Li et al. 2023). Indirect end-labeling experiments (Galloni et al. 1993; Samal et al. 1981; Udvardy and Schedl 1984) indicate that LBC protects an ~150-180 bp DNA segment from MNase digestion, which would explain why LBC-bound sequences are able to generate dots in MicroC experiments. Also unlike typical boundary elements, the pairing interactions of the LBC elements we’ve tested appear to be orientation-independent (unpublished data).

The difference in MNase sensitivity between typical TAD boundaries and LBC-bound elements is illustrated in the MicroC of the Leukocyte-antigen-related-like (Lar) meta-loop in Author response image 4 panels A and B. Direct physical pairing of two TAD boundaries (blue and purple) brings two TADs encompassing the 125 kb lar gene into contact with two TADs in a gene poor region 620 kb away. This interaction generates two regions of greatly enhanced contact: the two boxes on either side of the paired boundaries (panel A). Note that like transgene homie pairing with the eve boundaries, the boundary pairing interaction that forms the lar meta-loop is orientation-dependent. In this case the TAD boundary in the Lar locus pairs with the TAD boundary in the gene poor region head-to-head (arrow tip to arrow tip), generating a circle-loop. This circle-loop configuration brings the TAD upstream of the blue boundary into contact with the TAD upstream of the purple boundary. Likewise, the TAD downstream of the blue boundary is brought into contact with the TAD downstream of the purple boundary.

In the MicroC procedure, the sequences that correspond to the paired boundaries are not recovered (red arrow in Author response image 4 panel B). This is why there are vertical and horizontal blank stripes (red arrowheads) emanating from the missing point of contact. Using a different HiC procedure (dHS-C) that allows us to recover sequences from typical boundary elements (Author response image 4 panels C and D), there is a strong “dot” at the point of contact which corresponds to the pairing of the blue and purple boundaries.

There is a second dot (green arrow) within the box that represents physical contacts between sequences in the TADs downstream of the blue and purple boundaries. This dot is resistant to MNase digestion and is visible both in the MicroC and dHS-C profiles. Based on the ChIP signature of the corresponding elements in the two TADs downstream of the blue and purple boundaries, this dot represents paired LBC elements.

Lar metaloop.

Panels A & bB: MicroC. Panels C & D: dHS-C

That being said, the authors' analyses do not distinguish between the formation and the maintenance of domains. It is not clear to this reviewer why a single mechanism should explain the formation of the complex structures observed in static Hi-C heatmaps from a population of cells at a single developmental time point. For example, how can the authors rule out that extrusion initially provides the necessary proximity and possibly the cis preference of contacts required for boundaryboundary pairing whereas the latter may more reflect the structures observed at maintenance?

(2.2) The MicroC profiles shown in Fig. 2 of our paper were generated from nuclear cycle (NC) 14 embryos. NC14 is the last nuclear cycle before cellularization (Foe 1989). After the nuclei exit mitosis, S-phase begins, and because satellite sequences are late replicating in this nuclear cycle, S phase lasts 50 min instead of only 4-6 min during earlier cycles (Shermoen et al. 2010). So unlike MicroC studies in mammals, our analysis of chromatin architecture in NC14 embryos likely offers the best opportunity to detect any intermediates that are generated during TAD formation. In particular, we should be able to observe evidence of cohesin linking the sequences from the two extruding strands together (the stripes) as it generates TADs de novo. However, there are no vertical stripes in the eve TAD as would be expected if cohesin entered at a few specific sites somewhere within the TAD and extruded loops in opposite directions synchronously, nor are their stripes at 45o as would be expected if it started at nhomie or homie (see Figure Supplemental 1). We also do not detect cohesin-generated stripes in any of the TADs in between eve and the attP site at -142 kb. Note that in some models, cohesin is thought to be continuously extruding loops. After hitting the CTCF roadblocks, cohesin either falls off after a short period and starts again or it breaks through one or more TAD boundaries generating the LDC domains. In this dynamic model, stripes of crosslinked DNA generated by the passing cohesin complex should be observed throughout the cell cycle. They are not.

As for formation versus maintenance, and the possible involvement of cohesin loop extrusion in the former, but not the latter: This question was indirectly addressed in point #1.2 above. In this point we described multiple examples of specific boundary:boundary pairing interactions that take place over Mbs, in cis and in trans and even between different chromosomes. These long-distance interactions don’t preexist; instead they must be established de novo and then maintained. This process was actually visualized in the studies of Lim et al. (Lim et al. 2018) on the establishment of trans boundary pairing interactions in NC14 embryos. There is no conceivable mechanism by which cohesin-based loop extrusion could establish the long or short distance trans interactions that have been documented in many studies on fly boundary elements. Also as noted above, its seems unlikely that it is necessary for long-range interactions in cis.

A more plausible scenario is that cohesin entrapment helps to stabilize these long-distance interactions after they are formed. If this were true, then one could argue that cohesin might also function to maintain TADs after boundaries have physically paired with their neighbors in cis. However, the Rad21 depletion experiments of Goel et al. (Goel et al. 2023) would rule out an essential role for cohesin in maintaining TADs after boundary:boundary pairing. In short, while we cannot formally rule out that loop extrusion might help bring sequences closer together to increase their chance of pairing, neither the specificity of that pairing, nor its orientation can be explained by loop extrusion. Furthermore, since pairing in trans cannot be facilitated by loop extrusion, invoking it as potentially important for boundary-boundary pairing in cis can only be described as a potential mechanism in search of a function, without clear evidence in its favor.

On the other hand, the apparent loss of contacts between TADs within large multi-TAD neighborhoods (Geol et al. 2023) would suggest that there is some sort of decompaction of neighborhoods after Rad21 depletion. It is possible that this might stress interactions that span multiple TADs as is the case for homie at -142, or for the other examples described in #1.2 above. This kind of involvement of cohesin might or might not be associated with a loop extrusion mechanism.

Future work aimed at analyzing micro-C data in cohesin-depleted cells might shed additional light on this.

(2.3) This experiment has been done by Goel et al. (Goel et al. 2023) in mammalian tissue culture cells. They found that TADs, as well as local TAD neighborhoods, are not disrupted/altered by Rad21 depletion (see Geol at al. 2023 and our response to point #1.1 of reviewer #1).

Additional mechanisms at play include compartment-level interactions driven by chromatin states. Indeed, in mammalian cells, these interactions often manifest as a "plume" on Hi-C maps similar to what the authors attribute to boundary interactions in this manuscript. How do the chromatin states in the neighboring domains of the eve locus impact the model if at all?

(2.4) Chromatin states have been implicated in driving compartment level interactions.

Compartments as initially described were large, often Mb sized, chromosomal segments that “share” similar chromatin marks/states, and are thought to merge via co-polymer segregation. They were visualized using large multi-kb bin sizes. In the studies reported here, we use bin sizes of 200 bp to examine a DNA segment of less than 200 kb which is subdivided into a dozen or so small TADs. Several of the TADs contain more than one transcription unit, and they are expressed in quite different patterns, and thus might be expected to have different “chromatin states” at different points in development and in different cells in the organism. However, as can be seen by comparing the MicroC patterns in our paper that are shown in Fig. 2 with Fig. 7, Figure Supplemental 5 and Figure Supplemental 6, the TAD organization in NC14 and 12-16 hr embryos is for the most part quite similar. There is no indication that these small TADs are participating in liquid phase compartmentalization that depends upon shared chromatin/transcriptional states in NC14 and then again in 12-16 hr embryos.

In NC14 embryos, eve is expressed in 7 stripes, while it is potentially active throughout much of the embryo. In fact, the initial pattern in early cycles is quite broad and is then refined during NC14. In 12-16 hr embryos, the eve gene is silenced by the PcG system in all but a few cells in the embryo. However, here again the basic structure of the TAD, including the volcano plume, looks quite similar at these different developmental stages.

As for the suggestion that the plume topping the eve volcano triangle is generated because the TADs flanking the eve TAD share chromatin states and coalesce via some sort of phase separation:

This model has been tested directly in Ke et al. (Ke et al. 2024). In Ke et al., we deleted the nhomie boundary and replaced it with either nhomie in the reverse orientation or homie in the forward orientation. According to the compartment model, changing the orientation of the boundaries so that the topology of the eve TAD changes from a stem-loop to a circle-loop should have absolutely no effect on the plume topping the eve volcano triangle. The TADs flanking the eve TAD would still be expected to share the same chromatin states and would still be able to coalesce via phase transition. However, this is not what is observed. The plume disappears and is replaced by “clouds” on both sides of the eve TAD. The clouds arise because the eve TAD bumps into the neighboring TADs when the topology is a circle-loop.

We would also note that “compartment-level” interactions would not explain the findings presented in Muller at al. 1999, in Table 1 or in Author response image 4. It is clear that the long distant (Mb) interactions observed for Mcp, gypsy, Fab-7, homie, nhomie and the blue and purple boundaries in Author response image 4 arise by the physical pairing of TAD boundary elements. This fact is demonstrated directly by the MicroC experiments in Fig. 7 and Fig Supplemental 4 and 5, and by the MicroC and dHS-C experiments in Author response image 4. There is no evidence for any type of “compartment/phase separation” driving these specific boundary pairing interactions.

In fact, given the involvement of TAD boundaries in meta-loop formation, one might begin to wonder whether some of the “compartment level interactions” are generated by the specific pairing of TAD boundary elements rather than by “shared chromatin” states. For example, the head-tohead pairing of the blue and purple boundaries generates a Lar meta-loop that has a circle-loop topology. As a consequence, sequences upstream of the blue and purple boundary come into contact, generating the small dark rectangular box on the upper left side of the contact map. Sequences downstream of the blue and purple boundary also come into contact, and this generates the larger rectangular box in the lower right side of the contact map. A new figure, Fig. 9, shows that the interaction pattern flips (lower left and top right) when the meta-loop has a stem-loop topology. If these meta-loops are visualized using larger bin sizes, the classic “compartment” patchwork pattern of interactions emerges. Would the precise patchwork pattern of “compartmental” interactions involving the four distant TADs that are linked in the two meta-loops shown in Fig. 9 persist as is if we deleted one of the TAD boundaries that forms the meta-loop? Would the precise patchwork pattern persist if we inverted one of the meta-loop boundaries so that we converted the topology of the loop from a circle-loop to a stem-loop or vice versa? We haven’t used MicroC to compare the compartment organization after deleting or inverting a meta-loop TAD boundary; however, a comparison of the MicroC pattern in WT in Fig. 1C with that for the homie transgenes in Fig. 7 and Figs. Supplemental 5, 6 and 7 indicates a) that novel patterns of TAD:TAD interactions are generated by this homie dependent mini-meta-loop and b) that the patterns of TAD:TAD interactions depend upon loop topology. Were these novel TAD:TAD interactions generated instead by compartment level interactions/shared chromatin states, they should be evident in WT as well (Fig. 1). They are not.

How does intrachromosomal homolog pairing impact the models proposed in this manuscript (Abed et al. 2019; Erceg et al., 2019). Several papers recently have shown that somatic homolog pairing is not uniform and shows significant variation across the genome with evidence for both tight pairing regions and loose pairing regions. Might loose pairing interactions have the capacity to alter the cis configuration of the eve locus?

(2.5) At this point it is not entirely clear how homolog pairing impacts the cis configuration/MicroC contact maps. We expect that homolog pairing is incomplete in the NC14 embryos we analyzed; however, since replication of eve and the local neighborhood is likely complete, sister chromosomes should be paired. So we are likely visualizing the 3D organization of paired TADs.

In summary, the transgenic experiments are extensive and elegant and fully support the authors' models. However, in my opinion, they do not completely rule out additional models at play, including extrusion-based mechanisms. Indeed, my major issue is the limited conceptual advance in this manuscript. The authors essentially repeat many of their previous work and analyses.

(2.6) In our view, the current paper makes a number of significant contributions that go well beyond those described in our 2016 publication. These are summarized below.

A) While our 2016 paper used transgenes inserted in the -142 kb attP site to study pairing interactions of homie and nhomie, we didn’t either consider or discuss how our findings might bear on the loop extrusion model. However, since the loop extrusion model is currently accepted as established fact by many labs working on chromosome structure, it is critically important to devise experimental approaches which test the predictions of this particular model. One approach would be to deplete cohesin components; however, as discussed in #1.1, our experimental system is not ideal for this type of approach. On the other hand, there are other ways to test the extrusion model. Given the mechanism proposed for TAD formation—extruding a loop until cohesin bumps into CTCF/boundary road blocks—it follows that only two types of loop topologies are possible: stemloop and unanchored loop. The loop extrusion model, as currently conceived, can’t account for the two cases in this study in which the reporter on the wrong side of the homie boundary from the eve locus is activated by the eve enhancers. In contrast, our findings are completely consistent with orientation-specific boundary:boundary pairing.

B) In the loop extrusion model, cohesin embraces both of the extruded chromatin fibers, transiently bringing them into close proximity. As far as we know, there have been no (high resolution) experiments that have actually detected these extruding cohesin complexes during TAD formation. In order to have a chance of observing the expected signatures of extruding cohesin complexes, one would need a system in which TADs are being formed. As described in the text, this is why we used MicroC to analyze TADs in NC14 embryos. We do not detect the signature stripes that would be predicted (see Figure Supp 2) by the current version of the loop extrusion model.

C) Reporter expression in the different -142 kb transgenes provides only an indirect test of the loop extrusion and boundary:boundary pairing models for TAD formation. The reporter expression results need to be confirmed by directly analyzing the pattern of physical interactions in each instance. While we were able to detect contacts between the transgenes and eve in our 2016 paper, the 3C experiments provided no information beyond that. By contrast, the MicroC experiments in the current paper give high resolution maps of the physical contacts between the transgene and the eve TAD. The physical contacts track completely with reporter activity. Moreover, just as is the case for reporter activity, the observed physical interactions are inconsistent with the loop extrusion model.

D) Genetic studies in Muller et al. (Muller et al. 1999) and imaging in Vazquez et al. (Vazquez et al. 2006) suggested that more than two boundaries can participate in pairing interactions. Consistent with these earlier observations, viewpoint analysis indicates the transgene homie interacts with both eve boundaries. While this could be explained by transgene homie alternating between nhomie and homie in the eve locus, this would require the remodeling of the eve TAD each time the pairing interaction switched between the three boundary elements. Moreover, two out of the three possible pairing combinations would disrupt the eve TAD, generating an unanchored loop (c.f., the lambda DNA TAD in Ke et al., (Ke et al. 2024)). However, the MicroC profile of the eve TAD is unaffected by transgenes carrying the homie boundary. This would suggest that like Mcp, the pairing interactions of homie and nhomie might not be exclusively pairwise. In this context is interesting to compare the contact profiles of the lar meta-loop shown in Author response image 4 with the different 142 kb homie inserts. Unlike the homie element at -142 kb, there is clearly only a single point of contact between the blue and purple boundaries.

E) Chen et al. (Chen et al. 2018) used live imaging to link physical interactions between a homie containing transgene inserted at -142 kb and the eve locus to reporter activation by the eve enhancers. They found that the reporter was activated by the eve enhancers only when it was in “close proximity” to the eve gene. “Close proximity” in this case was 331 nM. This distance is equivalent to ~1.1 kb of linear duplex B form DNA, or ~30 nucleosome core particles lined up in a row. It would not be possible to ligate two DNAs wrapped around nucleosome core particles that are located 330 nM apart in a fixed matrix. Since our MicroC experiments were done on embryos in which the gene is silent in the vast majority of cells, it is possible that the homie transgene only comes into close enough proximity for transgene nucleosome: eve nucleosome ligation events when the eve gene is off. Alternatively, and clearly more likely, distance measurements using imaging procedures that require dozens of fluorescent probes may artificially inflate the distance between sequences that are actually close enough for enzymatic ligation.

F) The findings reported in Goel et al. (Goel et al. 2023) indicate that mammalian TADs don’t require cohesin activity; however, the authors do not provide an alternative mechanism for TAD formation/stability. Here we have suggested a plausible mechanism.

The authors make no attempt to dissect the mechanism of this process by modifying extrusion components directly.

(2.7) See point #1.1

Some discussion of Rollins et al. on the discovery of Nipped-B and its role in enhancer-promoter communication should also be made to reconcile their conclusions in the proposed absence of extrusion events.

(2.8) The reason why reducing nipped-B activity enhances the phenotypic effects of gypsy-induced mutations is not known at this point; however, the findings reported in Rollins et al. (Rollins et al. 1999) would appear to argue against an extrusion mechanism for TAD formation.

Given what we know about enhancer blocking and TADs, there are two plausible mechanisms for how the Su(Hw) element in the gypsy transposon blocks enhancer-promoter interactions in the gypsy-induced mutants studied by Rollins et al. First, the Su(Hw) element could generate two new TADs through pairing interactions with boundaries in the immediate neighborhood. This would place the enhancers in one TAD and the target gene in another TAD. Alternatively, the studies of Sigrist and Pirrotta (Sigrist and Pirrotta 1997) as well as several publications from Victor Corces’ lab raise the possibility that the Su(Hw) element in gypsy-induced mutations is pairing with gypsy transposons inserted elsewhere in the genome. This would also isolate enhancers from their target genes. In either case, the loss of nipped-B activity increases the mutagenic effects of Su(Hw) element presumably by strengthening its boundary function. If this is due to a failure to load cohesin on to chromatin, this would suggest that cohesin normally functions to weaken the boundary activity of the Su(Hw) element, i.e., disrupting the ability of Su(Hw) elements to interact with either other boundaries in the neighborhood or with themselves. Were this a general activity of cohesin (to weaken boundary activity), one would imagine that cohesin normally functions to disrupt TADs rather than generate/stabilize TADs.

An alternative model is that Nipped-B (and thus cohesion) functions to stabilize enhancerpromoter interactions within TADs. In this case, loss of Nipped-B would result in a destabilization of the weak enhancer:promoter interactions that can still be formed when gypsy is located between the enhancer and promoter. In this model the loss of these weak interactions in nipped-b mutants would appear to increase the “blocking” activity of the gypsy element. However, this alternative model would also provide no support for the notion that Nipped-B and cohesin function to promote TAD formation.

Reviewer #3 (Public Review):

Bing et al. attempt to address fundamental mechanisms of TAD formation in Drosophila by analyzing gene expression and 3D conformation within the vicinity of the eve TAD after insertion of a transgene harboring a Homie insulator sequence 142 kb away in different orientations. These transgenes along with spatial gene expression analysis were previously published in Fujioka et al. 2016, and the underlying interpretations regarding resulting DNA configuration in this genomic region were also previously published. This manuscript repeats the expression analysis using smFISH probes in order to achieve more quantitative analysis, but the main results are the same as previously published. The only new data are the Micro-C and an additional modeling/analysis of what they refer to as the 'Z3' orientation of the transgenes. The rest of the manuscript merely synthesizes further interpretation with the goal of addressing whether loop extrusion may be occurring or if boundary:boundary pairing without loop extrusion is responsible for TAD formation. The authors conclude that their results are more consistent with boundary:boundary pairing and not loop extrusion; however, most of this imaging data seems to support both loop extrusion and the boundary:boundary models. This manuscript lacks support, especially new data, for its conclusions.

(3.1) The new results/contributions of our paper are described in #2.6 above.

Although there are (two) homie transgene configurations that give expression patterns that would be consistent with the loop extrusion model, that is not quite the same as strong evidence supporting loop extrusion. On the contrary, key aspects of the expression data are entirely inconsistent with loop extrusion, and they thus rule out the possibility that loop extrusion is sufficient to explain the results. Moreover, the conclusions drawn from the expression patterns of the four transgenes are back up by the MicroC contact profiles—profiles that are also not consistent with the loop extrusion model. Further, as documented above, loop extrusion is not only unable to explain the findings reported in this manuscript, but also the results from a large collection of published studies on fly boundaries. Since all of these boundaries function in TAD formation, there is little reason to think that loop extrusion makes a significant contribution at the TAD level in flies. Given the results reported by Goel et al. (Goel et al. 2023), one might also have doubts about the role of loop extrusion in the formation/maintenance of mammalian TADs.

To further document these points, we’ve included a new figure (Fig. 9) that shows two meta-loops. Like the loops seen for homie-containing transgenes inserted at -142 kb, meta-loops are formed by the pairing of distant fly boundaries. As only two boundaries are involved, the resulting loop topologies are simpler than those generated when transgene homie pairs with nhomie and homie in the eve locus. The meta-loop in panel B is a stem-loop. While a loop with this topology could be formed by loop extrusion, cohesion would have to break through dozens of intervening TAD boundaries and then somehow know to come to a halt at the blue boundary on the left and the purple boundary on the right. However, none of the mechanistic studies on either cohesin or the mammalian CTCF roadblocks have uncovered activities of either the cohesin complex or the CTCF roadblocks that could explain how cohesin would be able to extrude hundreds of kb and ignore dozens of intervening roadblocks, and then stop only when it encounters the two boundaries that form the beat-IV meta-loop. The meta-loop in panel A is even more problematic in that it is a circle-loop--a topology that can’t be generated by cohesin extruding a loop until comes into contact with CTCF roadblocks on the extruded strands.

Furthermore, there are many parts of the manuscript that are difficult to follow. There are some minor errors in the labelling of the figures that if fixed would help elevate understanding. Lastly, there are several major points that if elaborated on, would potentially be helpful for the clarity of the manuscript.

Major Points:

(1) The authors suggest and attempt to visualize in the supplemental figures, that loop extrusion mechanisms would appear during crosslinking and show as vertical stripes in the micro-C data. In order to see stripes, a majority of the nuclei would need to undergo loop extrusion at the same rate, starting from exactly the same spots, and the loops would also have to be released and restarted at the same rate. If these patterns truly result from loop extrusion, the authors should provide experimental evidence from another organism undergoing loop extrusion.

(3.2) We don’t know of any reports that actually document cohesion extrusion events that are forming TADs (TADs as defined in our paper, in the RCMC experiments of Goel et al. (Goel et al. 2023), in response #1.1, or in the high-resolution images from the MicroC data of Krietenstein et al (Krietenstein et al. 2020) and Hseih et al. (Hsieh et al. 2020). However, an extruding cohesin complex would be expected to generate stripes because it transiently brings together the two chromatin strands as illustrated by the broken zipper in Figure Supplemental 2 of our paper. While stripes generated by cohesin forming a TAD have not to our knowledge ever been observed, Fig. 4 in Goel et al. (Goel et al. 2023)) shows 45o stripes outlining TADs and connecting neighboring TADs. These stripes are visible with or without Rad21.

In some versions of the loop extrusion model, cohesin extrudes a loop until it comes to a halt at both boundaries, where it then remains holding the loop together. In this model, the extrusion event would occur only once per cell cycle. This is reason we selected NC14 embryos as this point in development should provide by far the best opportunity to visualize cohesin-dependent TAD formation. However, the expected stripes generated by cohesin embrace of both strands of the extruding loop were not evident. Other newer versions of the loop extrusion model are much more dynamic—cohesin extrudes the loop, coming to a halt at the two boundaries, but either doesn’t remain stably bound or breaks through one or both boundaries. In the former case, the TAD needs to be reestablished by another extrusion event, while in the latter case LDC domains are generated. In this dynamic model, we should also be able to observe vertical and 45o stripes (or stripes leaning to one side or another of the loading site if the extrusion rates aren’t equal on both fibers) in NC14 embryos corresponding to the formation of TADs and LDC domains. However, we don’t.

(2) On lines 311-314, the authors discuss that stem-loops generated by cohesin extrusion would possibly be expected to have more next-next-door neighbor contacts than next-door neighbor contacts and site their models in Figure 1. Based on the boundary:boundary pairing models in the same figure would the stem-loops created by head-to-tail pairing also have the same phenotype? Making possible enrichment of next-next-door neighbor contacts possible in both situations? The concepts in the text are not clear, and the diagrams are not well-labeled relative to the two models.

(3.3) Yes, we expect that stem-loops formed by cohesin extrusion or head-to-tail pairing would behave in a similar manner. They could be stem-loops separated by unanchored loops as shown in Fig. 1B and E. Alternatively, adjacent loops could be anchored to each other (by cohesin/CTCF road blocks or by pairing interactions) as indicated in Fig. 1C and F. In stem-loops generated either by cohesin extrusion or by head-to-tail pairing, next-next door neighbors should interact with each other, generating a plume above the volcano triangle. In the case of circle-loops, the volcano triangle should be flanked by clouds that are generated when the TAD bumps into both next-door neighbors. In the accompanying paper, we test this idea by deleting the nhomie boundary and then a) inserting nhomie back in the reverse orientation, or b) by inserting homie in the forward orientation. The MicroC patterns fit with the predictions that were made in this paper.

(3) The authors appear to cite Chen et al., 2018 as a reference for the location of these transgenes being 700nM away in a majority of the nuclei. However, the exact transgenes in this manuscript do not appear to have been measured for distance. The authors could do this experiment and include expression measurements.

(3.4) The transgenes used in Chen et al. are modified versions of a transgene used in Fujioka et al. (2016) inserted into the same attP site. When we visualize reporter transcription in NC14 embryos driven by the eve enhancers using smFISH, HCR-FISH or DIG, only a subset of the nuclei at this stage are active. The number of active nuclei we detect is similar to that observed in the live imaging experiments of Chen et al. The reason we cited Chen et al. (Chen et al. 2018) was that they found that proximity was a critical factor in determining whether the reporter was activated or not in a given nucleus. The actual distance they measured wasn’t important. Moreover, as we discussed in response #2.6 above, there are good reasons to think that the “precise” distances measured in live imaging experiments like those used in Chen et al. are incorrect. However, their statements are certainly correct if one considers that a distance of ~700 nM or so is “more distant” relative to a distance of ~300 nM or so, which is “closer.”

(4) The authors discuss the possible importance of CTCF orientation in forming the roadblock to cohesin extrusion and discuss that Homie orientation in the transgene may impact Homie function as an effective roadblock. However, the Homie region inserted in the transgene does not contain the CTCF motif. Can the authors elaborate on why they feel the orientation of Homie is important in its ability to function as a roadblock if the CTCF motif is not present? Trans-acting factors responsible for Homie function have not been identified and this point is not discussed in the manuscript.

We discussed the “importance” of CTCF orientation in forming roadblocks because one popular version of the cohesin loop extrusion/CTCF roadblock model postulates that CTCF must be oriented so that the N-terminus of the protein is facing towards the oncoming cohesin complex, otherwise it won’t be able to halt extrusion on that strand. When homie in the transgene is pointing towards the eve locus, the reporter on the other side (farther from eve) is activated by the eve enhancers. One possible way to explain this finding (if one believes the loop extrusion model) is that when homie is inverted, it can’t stop the oncoming cohesin complex, and it runs past the homie boundary until it comes to a stop at a properly oriented boundary farther away. In this case, the newly formed loop would extend from the boundary that stopped cohesin to the homie boundary in the eve locus, and would include not only the distal reporter, but also the proximal reporter. If both reporters are in the same loop with the eve enhancers (which they would have to be given the mechanism of TAD formation by loop extrusion), both reporters should be activated. They are not.

For the boundary pairing model, the reporter that will be activated will depend upon the orientation of the pairing interaction—which can be either head-to-head or head-to-tail (or both: see discussion of LBC elements in #2.1). For an easy visualization of how the orientation of pairing interactions is connected to the patterns of interactions between sequences neighboring the boundary, please look at Fig. 9. This figure shows two different meta-loops. In panel A, head-tohead pairing of the blue and purple boundaries brings together, on the one hand, sequences upstream of the blue and purple boundary, and on the other hand, sequences downstream of the blue and purple boundaries. In the circle loop configuration, the resulting rectangular boxes of enhanced contact are located in the upper left and lower right of the contact map. In panel B, the head-to-tail pairing of the blue and purple boundary changes how sequences upstream and downstream of the blue and purple boundaries interact with each other. Sequences upstream of the blue boundary interact with sequences downstream of the purple boundary, and this gives the rectangular box of enhanced interactions on the top right. Sequences downstream of the blue boundary interact with sequences upstream of the purple boundary, and this gives the rectangular box of enhanced contact on the lower left.

CTCF: Our analysis of the homie boundary suggests that CTCF contributes little to its activity. It has an Su(Hw) recognition sequence and a CP190 “associated” sequence. Mutations in both compromise boundary activity (blocking and -142 kb pairing). Gel shift experiments and ChIP data indicate there are half a dozen or more additional proteins that associate with the 300 bp homie fragment used in our experiments.

Orientation of CTCF or other protein binding sites: The available evidence suggests that orientation of the individual binding sites is not important (Kyrchanova et al. 2016; Lim et al. 2018)). Instead, it is likely that the order of binding sites affects function.

(5) The imaging results seem to be consistent with both boundary:boundary interaction and loop extrusion stem looping.

It is not clear whether the reviewer is referring to the different patterns of reporter expression— which clearly don’t fit with the loop extrusion model in the key cases that distinguish the two models—or the live imaging experiments in Chen et al. (Chen et al. 2018).

(6) The authors suggest that the eveMa TAD could only be formed by extrusion after the breakthrough of Nhomie and several other roadblocks. Additionally, the overall long-range interactions with Nhomie appear to be less than the interactions with endogenous Homie (Figures 7, 8, and supplemental 5). Is it possible that in some cases boundary:boundary pairing is occurring between only the transgenic Homie and endogenous Homie and not including Nhomie?

Yes, it is possible. On the other hand, the data that are currently available supports the idea that transgene homie usually interacts with endogenous homie and nhomie at the same time. This is discussed in #2.6D above. The viewpoints indicate that crosslinking occurs more frequently to homie than to nhomie. This could indicate that when there are only pairwise interactions, these tend to be between homie and homie. Alternatively, this could also be explained by a difference in relative crosslinking efficiency.

(7) In Figure 4E, the GFP hebe expression shown in the LhomieG Z5 transgenic embryo does not appear in the same locations as the LlambdaG Z5 control. Is this actually hebe expression or just a background signal?

The late-stage embryos shown in E are oriented differently. For GlambdaL, the embryo is oriented so that hebe-like reporter expression on the ventral midline is readily evident. However, this orientation is not suitable for visualizing eve enhancer-dependent expression of the reporters in muscle progenitor cells. For this reason, the 12-16 hr GeimohL embryo in E is turned so that the ventral midline isn’t readily visible in most of the embryo. As is the case in NC14 embyros, the eve enhancers drive lacZ but not gfp expression in the muscle progenitor cells.

(8) Figure 6- The LhomieG Z3 (LeimohG) late-stage embryo appears to be showing the ventral orientation of the embryo rather than the lateral side of the embryo as was shown in the previous figure. Is this for a reason? Additionally, there are no statistics shown for the Z3 transgenic images.

Were these images analyzed in the same way as the Z5 line images?

The LeimohG embryo was turned so that the hebe enhancer-dependent expression of lacZ is visible. While the eve enhancer-dependent expression of lacZ in the muscle progenitor cells isn’t visible with this orientation, eve enhancer-dependent expression in the anal plate is.

(9) Do the Micro-C data align with the developmental time points used in the smFISH probe assays?

The MicroC data aligns with the smFISH images of older embryos: 12-14 hour embryos or stages 14-16.

Recommendations for the authors:

Reviewer #1 (Recommendations For The Authors):

This was a difficult paper to review. It took me several hours to understand the terminology and back and forth between different figures to put it together. It might be useful to put the loop models next to the MicroC results and have a cartoon way of incorporating which enhancers are turning on which reporters.

I also found the supercoiled TAD models in Figure 1 not useful. These plectoneme-type of structures likely do not exist, based on the single-cell chromosome tracing studies, and the HiC structures not showing perpendicular to diagonal interactions between the arms of the plectonemes.

We wanted to represent the TAD as a coiled 30nM fiber, as they are not likely to resemble the large loops like those shown in Fig. 1 A, D, and G.

There are no stripes emerging from homies, which is consistent with the pairing model, but there seem to be stripes from the eve promoter. I think these structures may be a result of both the underlying loop extruders + pairing elements.

There are internal structures in the eve TAD that link the upstream region of the eve promoter to the eve PRE and sequences in nhomie. All three of these sequences are bound by LBC. Each of the regulatory domains in BX-C also have LBC elements and, as shown in Author response image 1, you can see stripes connecting some of these LBC elements to each other. Since the stripes that Goel et al. (Goel et al. 2023) observed in their RCMC analysis of Ppm1g didn’t require cohesin, how these stripes are generated (active: e.g, a chromatin remodeler or passive: e.g., the LBC complex has non-specific DNA binding activity that can be readily crosslinked as the chromatin fiber slides past) isn’t clear.

The authors say there are no TADs that have "volcano plumes" but the leftmost TAD TA appears to have one. What are the criteria for calling the plumes? I am also not clear why there is a stripe off the eve volcano. It looks like homie is making a "stripe" loop extrusion type of interaction with the next TAD up. Is this maybe cohesin sliding off the left boundary?

The reviewer is correct, the left-most TAD TA appears to have a plume. We mentioned TA seems to have a plume in the original text, but it was inadvertently edited out.

Two different types of TADßàTAD interactions are observed. In the case of eve, the TADs to either side of eve interact more frequently with each other than they do with eve. This generates a “plume” above the eve volcano triangle. The TADs that comprise the Abd-B regulatory domains (see Author response image 1) are surrounded by clouds of diminishing intensity. Clouds at the first level represent interactions with both next-door neighbors; clouds at the second level represent interactions with both next-next-door neighbors; clouds at the third level represent interactions with next-next-next door neighbors. The Abd-B TADs are close to the same size, so that interactions with neighbors are relatively simple. However, this is not always the case. When there are smaller TADs near larger TADs the pattern of interaction can be quite complicated. An example is indicated by the red bar in Author response image 2

The authors state "In the loop-extrusion model, a cohesin complex initiating loop extrusion in the eve TAD must break through the nhomie roadblock at the upstream end of the eve TAD. It must then make its way past the boundaries that separate eve from the attP site in the hebe gene, and come to a halt at the homie boundary associated with the lacZ reporter." Having multiple loops formed by cohesin would also bring in the 142kb apart reporter and homie. Does cohesin make 140 kb long loops in flies?

A mechanism in which cohesin brings the reporter close to the eve TAD by generating many smaller loops (which would be the intervening TADs) was discussed in #1.2.

Figure 5 title mistakes the transgene used?

Fixed.

In figure 6, the orientation of the embryos does not look the same for the late-stage panels. So it was difficult to tell if the eve enhancer was turning the reporter on.

Here we were focusing mainly on the AP enhancer activation of the reporter, as this is most easily visualized. It should be clear from the images that the appropriate reporter is activated by the AP enhancer for each of the transgene inserts.

It is not clear to me why the GFP makes upstream interactions (from the 4C viewpoint) in GhomileLZ5 but not in LhomieGZ5? Corresponding interactions for Fig Supp 5 & 6 are not the same. That is, LacZ in the same place and with the same homie orientation does not show a similar upstream enrichment as the GFP reporter does.

We are uncertain as to whether we understand this question/comment. In GhomieLZ5 (now GhomieL, the lacZ reporter is on the eve side of the homie boundary while gfp is on the hebe enhancer side of the homie boundary. Since homie is pointing away from gfp, pairing interactions with homie and nhomie in the eve locus bring the eve enhancers in close proximity with the gfp reporter. This is what is seen in Fig. 7 panel D—lower trace. In LhomieGZ5 (now GeimohL) the lacZ reporter is again on the eve side of the homie boundary while gfp is on the hebe enhancer side of the homie boundary. However, in this case homie is inverted so that it is points away from lacZ (towards gfp). In this orientation, pairing brings the lacZ reporter into contact with the eve enhancers. This is what is seen in the upper trace in Fig. 7 panel D.

The orientation of the transgene is switch in Fig. Supp 5 and 6. For these “Z3) transgenes (now called LeimohG and LhomieG the gfp reporter is on the eve side of homie while the lacZ reporter is on the hebe enhancer side of homie. The interactions between the reporters and eve are determined by the orientation of homie in the transgene. When homie is pointing away from gfp (as in LeimohG), gfp is activated and that is reflected in the trace in Supp Fig. 5. When homie is pointing away from lacZ, lacZ is activated and this is reflected (though not as cleanly as in other cases) in the trace in Supp Fig. 6.

I did not see a data availability statement. Is the data publicly available? The authors also should consider providing the sequences of the insertions, or provide the edited genomes, in case other researchers would like to analyze the data.

Data have been deposited.

Reviewer #3 (Recommendations For The Authors):

Minor Points:

(1) There is an inconsistency in the way that some of the citations are formatted. Some citations have 'et al' italicized while others do not. It seems to be the same ones throughout the manuscript. Some examples: Chetverina et al 2017, Chetverina et al 2014, Cavalheiro et al 2021, Kyrchanova et al 2008a, Muravyova et al 2001.

Fixed

(2) Pita is listed twice in line 48.

Fixed

(3) Line 49, mod(mdg4)67.2 is written just as mod(mdg4). The isoform should be indicated.

This refers to all Mod isoforms.

(4) Homie and Nhomie are italicized throughout the manuscript and do not need to be.

This is the convention used previously.

(5) The supplemental figure captions 1 and 2 in the main document are ordered differently than in the supplemental figures file. This caused it to look like the figures are being incorrectly cited in lines 212-214 and 231-232.

Fixed

(6) Is the correct figure being cited in line 388-389? The line cites Figure 6E when mentioning LlambdaG Z5; however, LlambdaG Z5 is not shown in Figure 6.

Fixed

(7) Section heading 'LhomieG Z5 and GhomieL Z5' could be renamed for clarity. GhomieL Z5 results are not mentioned until the next section, named 'GhomieL Z5'.

Fixed

(8) Can the authors provide better labeling for control hebe expression? This would help to determine what is hebe expression and what is background noise in some of the embryos in Figures 4-6.

Author response image 5 shows expression of the lacZ reporter in GeimohL and GlambdaL. For the GlambdaL transgene, the hebe enhancers drive lacZ expression in 1216 hr embryos. Note that lacZ expression is restricted to a small set of quite distinctive cells along the ventral midline. lacZ is also expressed on the ventral side of the GeimohL embryo (top panel). However, their locations are quite different from those of the lacZ positive cells in the GlambdaL transgene embryo. These cells are displaced from the midline, and are arranged as pairs of cells in each hemisegment, locations that correspond to eve-expressing cells in the ventral nerve cord. The eve enhancers also drive lacZ expression elsewhere in the GeimohL embryo, including the anal plate and dorsal muscle progenitor cells (seen most clearly in the lower left panel).

lacZ expression in Giemohl and Glambdal embryos

(9) The Figure 5 title is labeled with the wrong transgene.

Fixed

(10) Heat map scales are missing for Figures 7, supplemental 5, and supplemental 6.

Fixed

(11) Did the authors check if there was a significant difference in the expression of GFP and lacZ from lambda control lines to the Homie transgenic lines?

Yes. Statistical analysis added in Table Supplemental #1

(12) The Figure 7 title references that these are Z3 orientations, however, it is Z5 orientations being shown.

Fixed

(13) The virtual 4C data should include an axis along the bottom of the graphs for better clarity. An axis is missing in all 4C figures.

References:

Bantignies F, Grimaud C, Lavrov S, Gabut M, Cavalli G. 2003. Inheritance of polycomb-dependent chromosomal interactions in drosophila. Genes Dev. 17(19):2406-2420.

Batut PJ, Bing XY, Sisco Z, Raimundo J, Levo M, Levine MS. 2022. Genome organization controls transcriptional dynamics during development. Science. 375(6580):566-570.

Bonchuk A, Boyko K, Fedotova A, Nikolaeva A, Lushchekina S, Khrustaleva A, Popov V, Georgiev P. 2021. Structural basis of diversity and homodimerization specificity of zinc-fingerassociated domains in drosophila. Nucleic Acids Res. 49(4):2375-2389.

Bonchuk AN, Boyko KM, Nikolaeva AY, Burtseva AD, Popov VO, Georgiev PG. 2022. Structural insights into highly similar spatial organization of zinc-finger associated domains with a very low sequence similarity. Structure. 30(7):1004-1015.e1004.

Chen H, Levo M, Barinov L, Fujioka M, Jaynes JB, Gregor T. 2018. Dynamic interplay between enhancer–promoter topology and gene activity. Nat Genet. 50(9):1296.

Fedotova AA, Bonchuk AN, Mogila VA, Georgiev PG. 2017. C2h2 zinc finger proteins: The largest but poorly explored family of higher eukaryotic transcription factors. Acta Naturae. 9(2):4758.

Foe VE. 1989. Mitotic domains reveal early commitment of cells in drosophila embryos. Development. 107(1):1-22.

Fujioka M, Mistry H, Schedl P, Jaynes JB. 2016. Determinants of chromosome architecture: Insulator pairing in cis and in trans. PLoS Genet. 12(2):e1005889.

Galloni M, Gyurkovics H, Schedl P, Karch F. 1993. The bluetail transposon: Evidence for independent cis-regulatory domains and domain boundaries in the bithorax complex. The EMBO Journal. 12(3):1087-1097.

Goel VY, Huseyin MK, Hansen AS. 2023. Region capture micro-c reveals coalescence of enhancers and promoters into nested microcompartments. Nat Genet. 55(6):1048-1056.

Hsieh TS, Cattoglio C, Slobodyanyuk E, Hansen AS, Rando OJ, Tjian R, Darzacq X. 2020. Resolving the 3d landscape of transcription-linked mammalian chromatin folding. Mol Cell. 78(3):539553.e538.

Ke W, Fujioka M, Schedl P, Jaynes JB. 2024. Chromosome structure ii: Stem-loops and circle-loops. eLife.

Krietenstein N, Abraham S, Venev SV, Abdennur N, Gibcus J, Hsieh TS, Parsi KM, Yang L, Maehr R, Mirny LA et al. 2020. Ultrastructural details of mammalian chromosome architecture. Mol Cell. 78(3):554-565.e557.

Kyrchanova O, Ibragimov A, Postika N, Georgiev P, Schedl P. 2023. Boundary bypass activity in the abdominal-b region of the drosophila bithorax complex is position dependent and regulated. Open Biol. 13(8):230035.

Kyrchanova O, Kurbidaeva A, Sabirov M, Postika N, Wolle D, Aoki T, Maksimenko O, Mogila V, Schedl P, Georgiev P. 2018. The bithorax complex iab-7 polycomb response element has a novel role in the functioning of the fab-7 chromatin boundary. PLoS Genet. 14(8):e1007442.

Kyrchanova O, Mogila V, Wolle D, Deshpande G, Parshikov A, Cleard F, Karch F, Schedl P, Georgiev P. 2016. Functional dissection of the blocking and bypass activities of the fab-8 boundary in the drosophila bithorax complex. PLoS Genet. 12(7):e1006188.

Kyrchanova O, Sabirov M, Mogila V, Kurbidaeva A, Postika N, Maksimenko O, Schedl P, Georgiev P.

2019a. Complete reconstitution of bypass and blocking functions in a minimal artificial fab7 insulator from drosophila bithorax complex. Proceedings of the National Academy of Sciences.201907190.

Kyrchanova O, Wolle D, Sabirov M, Kurbidaeva A, Aoki T, Maksimenko O, Kyrchanova M, Georgiev P, Schedl P. 2019b. Distinct elements confer the blocking and bypass functions of the bithorax fab-8 boundary. Genetics.genetics. 302694.302019.

Li H-B, Muller M, Bahechar IA, Kyrchanova O, Ohno K, Georgiev P, Pirrotta V. 2011. Insulators, not polycomb response elements, are required for long-range interactions between polycomb targets in drosophila melanogaster. Mol Cell Biol. 31(4):616-625.

Li X, Tang X, Bing X, Catalano C, Li T, Dolsten G, Wu C, Levine M. 2023. Gaga-associated factor fosters loop formation in the drosophila genome. Mol Cell. 83(9):1519-1526.e1514.

Lim B, Heist T, Levine M, Fukaya T. 2018. Visualization of transvection in living drosophila embryos. Mol Cell. 70(2):287-296. e286.

Link N, Kurtz P, O'Neal M, Garcia-Hughes G, Abrams JM. 2013. A p53 enhancer region regulates target genes through chromatin conformations in cis and in trans. Genes Dev. 27(22):24332438.

Mohana G, Dorier J, Li X, Mouginot M, Smith RC, Malek H, Leleu M, Rodriguez D, Khadka J, Rosa P et al. 2023. Chromosome-level organization of the regulatory genome in the drosophila nervous system. Cell. 186(18):3826-3844.e3826.

Muller M, Hagstrom K, Gyurkovics H, Pirrotta V, Schedl P. 1999. The mcp element from the drosophila melanogaster bithorax complex mediates long-distance regulatory interactions. Genetics. 153(3):1333-1356.

Postika N, Metzler M, Affolter M, Müller M, Schedl P, Georgiev P, Kyrchanova O. 2018. Boundaries mediate long-distance interactions between enhancers and promoters in the drosophila bithorax complex. PLoS Genet. 14(12):e1007702.

Rollins RA, Morcillo P, Dorsett D. 1999. Nipped-b, a drosophila homologue of chromosomal adherins, participates in activation by remote enhancers in the cut and ultrabithorax genes. Genetics. 152(2):577-593.

Samal B, Worcel A, Louis C, Schedl P. 1981. Chromatin structure of the histone genes of d. Melanogaster. Cell. 23(2):401-409.

Shermoen AW, McCleland ML, O'Farrell PH. 2010. Developmental control of late replication and s phase length. Curr Biol. 20(23):2067-2077.

Shidlovskii YV, Bylino OV, Shaposhnikov AV, Kachaev ZM, Lebedeva LA, Kolesnik VV, Amendola D, De Simone G, Formicola N, Schedl P et al. 2021. Subunits of the pbap chromatin remodeler are capable of mediating enhancer-driven transcription in drosophila. Int J Mol Sci. 22(6).

Sigrist CJ, Pirrotta V. 1997. Chromatin insulator elements block the silencing of a target gene by the drosophila polycomb response element (pre) but allow trans interactions between pres on different chromosomes. Genetics. 147(1):209-221.

Udvardy A, Schedl P. 1984. Chromatin organization of the 87a7 heat shock locus of drosophila melanogaster. J Mol Biol. 172(4):385-403.

Vazquez J, Muller M, Pirrotta V, Sedat JW. 2006. The mcp element mediates stable long-range chromosome-chromosome interactions in drosophila. Molecular Biology of the Cell. 17(5):2158-2165.

Wolle D, Cleard F, Aoki T, Deshpande G, Schedl P, Karch F. 2015. Functional requirements for fab-7 boundary activity in the bithorax complex. Mol Cell Biol. 35(21):3739-3752.

Version 2

Abstract

Abstract

Two different models have been proposed to explain how the endpoints of chromatin looped domains (“TADs”) in eukaryotic chromosomes are determined. In the first, a cohesin complex extrudes a loop until it encounters a boundary element roadblock, generating a stem-loop. In this model, boundaries are functionally autonomous: they have an intrinsic ability to halt the movement of incoming cohesin complexes that is independent of the properties of neighboring boundaries. In the second, loops are generated by boundary:boundary pairing. In this model, boundaries are functionally non-autonomous, and their ability to form a loop depends upon how well they match with their neighbors. Moreover, unlike the loop-extrusion model, pairing interactions can generate both stem-loops and circle-loops. We have used a combination of MicroC to analyze how TADs are organized, and experimental manipulations of the even skipped TAD boundary, homie, to test the predictions of the “loop-extrusion” and the “boundary-pairing” models. Our findings are incompatible with the loop-extrusion model and instead suggest that the endpoints of TADs in flies are determined by a mechanism in which boundary elements physically pair with their partners, either head-to-head or head-to-tail, with varying degrees of specificity. Although our experiments do not address how partners find each other, the mechanism is unlikely to require loop extrusion.

Introduction

The chromosomes of multicellular animals have a regular and inheritable physical organization. This was first recognized in studies on the lampbrush chromosomes in amphibian oocytes and the polytene chromosomes of insects (Callan 1963; Nora et al. 2020; Zhimulev and Belyaeva 1991; Zhimulev et al. 1983). Subsequent research has shown that the architectural features inferred from analysis of lampbrush and polytene chromosomes are present in chromosomes throughout much of the animal kingdom. The key organizational principle is the subdivision of the chromatin fiber into a series of independent looped domains, commonly called “TADs” (Cavalheiro et al. 2021; Chetverina et al. 2017; Jerković et al. 2020; Matthews and White 2019; Rowley and Corces 2018). With important exceptions, the arrangement of TADs along a given chromosome tend to be invariant and are largely independent of the cell type or developmental stage. This regular and inheritable organization is a reflection of the underlying mechanism of TAD formation. TADs are separated from each other by special elements called boundaries or insulators. While these elements have been found in many different species, they have been most fully characterized in Drosophila (Cavalheiro et al. 2021; Chetverina et al. 2017). Fly boundaries span DNA sequences of 150 bp to 1.5 kb in length and contain one or more nucleosome-free nuclease-hypersensitive regions. These nuclease-hypersensitive regions are targets for a large collection of DNA binding proteins that have been implicated in boundary function. Some of the fly boundary factors are widely conserved (CTCF, BEN family proteins, GAF), while others appear to be restricted to insect linages (Su(Hw), Pita, Zw5, Zipic, Mod(mdg4), BEAF) (Heger et al. 2013; Heger and Wiehe 2014; Schoborg and Labrador 2010).

Boundary elements in flies are not only responsible for organizing the chromatin fiber, they also have genetic activities. When interposed between enhancers or silencers and target promoters, boundary elements block regulatory interactions (Bell et al. 2001; Chetverina et al. 2014; Chetverina et al. 2017). This insulating activity provides a mechanism for delimiting units of independent gene activity: genes located between a pair of compatible boundaries are subject to regulatory interactions with enhancers/silencers present in the same chromosomal interval, while they are insulated from the effects of enhancers/silencers located beyond either boundary in adjacent neighborhoods. Genetic studies suggest that the insulating activity of boundary elements is a consequence of subdividing the chromosome into a series of topologically independent domains (Cai and Shen 2001; Gohl et al. 2011; Muravyova et al. 2001). Organizing the chromatin fiber into looped domains enhances contacts between sequences within the loop, while it suppresses contacts with sequences outside of the loop. While it is not known whether a similar insulation mechanism is at play in mammals, the fact that mammalian chromosomes are also subdivided into TADs by boundary-like elements suggests that it might be.

A critical question in chromosome biology is the mechanism(s) responsible for determining the endpoints of the loop domains, the TADs. In mammals, a novel loop-extrusion model has been proposed to not only generate chromosomal loops but also determine the endpoints of those loops (Alipour and Marko 2012; Fudenberg et al. 2016; Guo et al. 2012; Guo et al. 2015; Nuebler et al. 2018; Sanborn et al. 2015). In this model, a cohesin complex initiates loop formation at a loading site within the “loop-to-be” and then extrudes a chromatin loop until it bumps into CTCF-dependent roadblocks, one on each side of the extruding loop. The location of these roadblocks combined with the processive action of the cohesin complex determines the endpoints of each TAD (Davidson and Peters 2021; Ghosh and Meyer 2021; Mirny and Dekker 2022; Perea-Resa et al. 2021). A key assumption of the loop-extrusion model is that mammalian boundaries are fully autonomous: they are roadblocks, and their physical presence in and of itself is sufficient to define the loop endpoint, independent of the functional properties of neighboring boundaries. In a more specific variant of this model, the relative orientation of the paired CTCF roadblocks is also important—in order to halt cohesin-mediated extrusion, the CTCF sites on each side of the loop must have a convergent orientation. In this model the CTCF roadblocks are able to block an oncoming cohesin complex in only one direction; however, their intrinsic blocking activity is independent of the orientation of other CTCF sites in their neighborhood.

While this loop extrusion mechanism is widely thought to be operative in mammals, the evidence regarding TAD formation and function in flies is seemingly inconsistent with this mechanism. Genetics studies have shown that fly boundaries are functionally non-autonomous and that their activities in both loop formation and gene regulation depend upon their ability to engage in direct physical interactions with other boundaries (Chetverina et al. 2014; Chetverina et al. 2017). That fly boundaries might function by physical pairing was first suggested by studies which showed that the gypsy transposon su(Hw) boundary and the bithorax complex (BX-C) Mcp boundary can mediate regulatory interactions (enhancer/silencer: reporter) between transgenes inserted at sites separated by 15-32 MB or more (Muller et al. 1999; Sigrist and Pirrotta 1997; Vazquez et al. 1993). Consistent with direct physical interactions, these distant transgenes co-localize in vivo, and in living tissue remain in contact for extended periods of time (Chen et al. 2018; Li et al. 2011; Vazquez et al. 2006).

Some of the parameters that govern pairing interactions have been defined in two different transgene assays, insulator bypass and boundary competition. In one version of the bypass assay, a set of enhancers are placed upstream of two different reporters (Cai and Shen 2001; Kyrchanova et al. 2008a; Muravyova et al. 2001). When a single boundary is introduced between the enhancers and the closest reporter, the enhancers are unable to activate either reporter. However, if the same boundary is placed downstream of the closest reporter, bypass is observed. In this case the closest reporter, which is bracketed by the two boundaries, is still insulated from the enhancers; however, the downstream reporter is activated (Muravyova et al. 2001). Kyrchanova et al. (Kyrchanova et al. 2008a) showed that bypass activity is, in most cases, orientation-dependent. When the same boundary is placed in the upstream and downstream position, they typically need to be introduced in the opposite orientation. The reason for this is that self-pairing is head-to-head, and this configuration generates a stem-loop topology, bringing the upstream enhancers into close proximity to the downstream reporter. When the two boundaries are introduced in the same orientation, the enhancers do not activate the downstream reporter because a circle-loop rather than a stem-loop is formed (Chetverina et al. 2017; Kyrchanova et al. 2008a). Head-to-head pairing seems to be a common feature of self-pairing interactions between fly boundaries and was observed for scs, scs’, iA2 and wari (Kyrchanova et al. 2008a). A preference for head-to-head interactions is also observed for heterologous pairing interactions between different boundaries in the Abdominal-B (Abd-B) region of the BX-C (Kyrchanova et al. 2011; Kyrchanova et al. 2008b). There are some exceptions. The BX-C Fab-7 boundary pairs with itself and with its neighbor Fab-8 in both orientations (Kyrchanova et al. 2011; Kyrchanova et al. 2008b). Bypass activity also seems to depend upon finding the proper match. For example, when multimerized dCTCF or Zw5 binding sites are placed both upstream and downstream of the closest reporter, bypass is observed. However, there is no bypass when multimerized dCTCF sites are tested with multimerized Zw5 sites (Kyrchanova et al. 2008a). Similar results have been reported for the Zipic and Pita factors (Zolotarev et al. 2016). Boundary:boundary pairing preferences are also observed in competition assays in which different boundaries are introduced into the same transgene (Gohl et al. 2011). In these assays, the insulating activity of a boundary in the blocking position (between an enhancer and a reporter that is flanked by a 3’ boundary) is challenged by placing a heterologous boundary upstream of the enhancer. If the boundary upstream of the enhancer is a better match for the boundary located downstream of the reporter, then the insulating activity of the boundary in the blocking position can be compromised or lost altogether. These and other experiments argue that fly boundaries are functionally non-autonomous—that is, their activities depend upon their ability to physically interact with other boundaries.

However, the differences between the two models are not limited to the mechanisms for determining the endpoints of TADs. The models also differ in the possible topologies of the chromatin loops that form TADs. Only two loop topologies can be generated by the cohesin-extrusion mechanism. One is a stem-loop while the other is an unanchored loop. Depending upon the location of the road-blocks that halt cohesin extrusion and the cohesin loading sites, the chromatin fiber would be organized into a series of stem-loops. These stem-loops will be separated from each other by unanchored loops (see Fig. 1B). This is the configuration that is often illustrated in articles discussing the loop-extrusion model. If the relevant road-blocks are very closely juxtaposed, then the unanchored connecting loops will disappear and be replaced by a series of stem-loops which point in opposite orientations, as illustrated in Fig. 1C. Stem-loops are also possible in the boundary-pairing model, and they will be formed when two heterologous boundaries pair with each other head-to-tail in cis (see Fig. 1D). Like loop-extrusion, the stem-loops could be separated from each other by unanchored loops (Fig. 1E). If the boundaries in the neighborhood pair with each other head-to-tail, a series of connected stem-loops pointing in opposite orientations will be generated. In this case, the main axis of the chromosome would be a series of paired boundaries (Fig. 1F).

Diagrams of possible loop topologies generated by loop extrusion and boundary:boundary pairing.

A) Cohesin embraces a loop at a loading site somewhere within the TAD-to-be and then extrudes a loop at an equal rate on both strands. Extrusion continues until cohesin encounters boundary roadblocks on both strands. B) In one model, the orientation of the roadblock is important. As a consequence, the chromatin fiber will be organized into a series of stem-loops separated from each other by unanchored loops. C) If the presence of a boundary roadblock, but not its orientation, is sufficient to halt extrusion, the chromatin fiber will be organized into a series of linked stem-loops. The main axis of the chromosome will be defined by the cohesin:CTCF roadblocks. D) The pairing of two boundaries head-to-tail generates a stem-loop. E) If boundary pairing interactions are strictly pairwise, head-to-tail pairing will generate a series of stem-loops separated from each other by unanchored loops. F) If boundaries can engage in multiple head-to-tail pairing interactions, the chromosome will be organized into a series of linked stem-loops. The main axis of the chromosome will be defined by a series of paired boundaries. G) The pairing of two boundaries head-to-head generates a circle-loop. H) If boundary pairing interactions are strictly pairwise there will be a series of circle-loops separated from each other by unanchored loops. I) If boundaries can engage in multiple head-to-head pairing interactions, the chromatin fiber will be organized into a series of circle-loops connected to each other at their base. Pairing interactions between boundaries #1 and #2 need not be in register with pairing of boundaries #2 and #3. In this case, the main axis of the chromosome may bend and twist, and this could impact the relative orientations of the circle-loops.

Stem-loops are not, however, the only loop topology that can be generated in the boundary-pairing model. If the neighboring boundaries pair with each other head-to-head instead of head-to-tail, a circle-loop structure will be generated (Fig. 1G). Like stem-loops, the circle-loops could be connected by unanchored loops (Fig. 1H). Alternatively, if boundaries pair with both neighbors head-to-head this will generate a series of linked circle-loops oriented in (more or less) the same direction (see Fig. 1I).

In the studies reported here, we have critically evaluated these two models. We have used a combination of MicroC to analyze how TADs are organized and experimental manipulations to test the predictions of the “loop-extrusion” and the “boundary-pairing” models. For these experimental manipulations, we have used the well-characterized homie boundary from the even skipped (eve) locus. homie together with nhomie flank the 16 kb eve locus, and these two elements share many of the properties of other fly boundaries (Fujioka et al. 2013; Fujioka et al. 2009). Like the gypsy and Mcp boundaries, nhomie and homie can mediate long-distance regulatory interactions. For example, a lacZ reporter transgene containing homie (or nhomie) inserted on one homolog can be activated by an eve enhancer in a homie- (or nhomie-) containing transgene inserted on the other homolog over a distance of 2 Mb (Fujioka et al. 2016). Long-distance activation is also observed in heterologous combinations of nhomie and homie, but not in combinations with phage lambda DNA. Similarly, regulatory interactions are observed when a homie- or nhomie- containing reporter is inserted at an attP landing site in the hebe gene, 142 kb upstream of the eve locus (Chen et al. 2018; Fujioka et al. 2016; Fujioka et al. 2009). In this case, the eve enhancers drive reporter expression during development in a pattern that mimics the stage- and tissue-specific expression of the endogenous eve gene. Activation also depends upon the orientation of the boundary (5’→3’) relative to the reporter in the transgene. For homie, the reporter must be located “upstream” of the boundary (5’→3’), just like the eve enhancers (and eve gene) are “upstream” of endogenous homie (see Fig. 2B). Little or no activation is observed when the relative orientation is reversed so that the reporter is located “downstream” of homie. For nhomie, the reporter must be located “downstream”, in the same relative position as that in which the eve enhancers and eve gene are located with respect to endogenous nhomie (5’→3’: Fig. 2B).

TAD organization of the even-skipped locus and neighboring sequences.

A) The eve TAD is a volcano with a plume that is anchored by nhomie and homie. ChIP-seq data below the MicroC map indicate that many of the known fly chromosomal architectural proteins are associated with the two eve boundary elements in vivo (Cuartero et al. 2014; Duan et al. 2021; Gaskill et al. 2021; Li et al. 2015; Sun et al. 2015; Ueberschär et al. 2019; Zolotarev et al. 2016). B) The eve locus forms a stem-loop structure. In this illustration, nhomie pairs with homie head-to-tail, and this forms a stem-loop which brings sequences upstream of nhomie and downstream of homie into contact, as is observed in insulator bypass assays (Kyrchanova et al. 2008a). The eve locus is shown assembled into a coiled “30 nM” chromatin fiber. C) MicroC contact pattern for the chromosomal region spanning the attP site at –142 kb and the eve locus. Like the eve volcano, this contact pattern was generated using aggregated previously published NC14 data (Batut et al. 2022; Levo et al. 2022). A black arrow indicates the –142kb locus where transgenes are integrated into the genome. A blue arrow indicates the position of the Etf-QO gene. Note the numerous TADs between the –142kb hebe locus from the eve “volcano TAD”. Individual TADs are labeled TA—TM. Black boxes indicate positions of sequences that, based on ChIP experiments, are bound by one or more known insulator proteins in vivo.

Results

The eve TAD

We used MicroC to probe the TAD organization of the eve locus and the genomic region that extends from eve to just beyond the attP site at –142 kb. For this purpose, we chose nuclear cycle 14 embryos. DNA replication commences soon after nuclei exit the previous mitosis and continues for nearly an hour until the nuclei arrest in G2, and the process of cellularization is completed. As sister chromosomes in flies pair with each other in register, there will be two juxtaposed copies of this region of the chromosome in most nuclei. The sister chromosomes should be linked to each other by cohesin in preparation for the asynchronous cellular divisions at the end of NC14 (Collier and Nasmyth 2022; Gligoris et al. 2014). While homolog pairing is limited in pre-cellular blastoderm embryos, there are regions of the chromosome that already show evidence of pairing (Erceg et al. 2019; Fung et al. 1998). In fact, transvection mediated by transgenes containing the gypsy insulator and the homie and Fab-8 boundaries has been observed in live imaging studies of nuclear cycle 14 embryos (Lim et al. 2018). Since the NC14 nuclei are just emerging from an earlier mitosis and are undergoing DNA synthesis, one would imagine that other key features of chromosome organization might also be in the process of being established. Thus, it should be possible to capture evidence of intermediates like those expected to be generated by extruding cohesin complexes as they move through the eve TAD and other nearby loci at this stage of development.

Shown in Fig. 2A is a blow-up of the TAD organization of the eve locus. As expected, the left and right boundaries of the eve TAD correspond to nhomie and homie, respectively, and sequences containing these two boundaries converge at the apex of the eve interaction triangle. All of the sequences within the TAD appear to come into contact with each other and define a high density of internal contact domain (HDIC). These contacts are expected to be generated, at least in part, by the sliding of the chromatin fiber within the TAD against itself, much like sequences in circular/supercoiled DNA can interact with each other (Fig. 1E). There are also sequences within the eve TAD that show enhanced interactions. For example, there is a darker interaction triangle connecting nhomie and the promoter region, while another triangle connects homie and the PRE adjacent to homie to the promoter region. These enhanced internal contacts would be expected to help insulate the eve gene from external regulatory elements and at the same time facilitate interactions between eve and its upstream and downstream enhancers.

As predicted by both the loop-extrusion and boundary-pairing models, the eve TAD is a stem-loop, not a circle-loop (Fig. 2B). This is evident from the prominent “volcano plume” at the apex of the eve TAD (volcano triangle). A volcano plume is observed because sequences flanking eve are brought into close proximity when homie and nhomie are linked to each other by either cohesin or by head-to-tail pairing (see Fig.1: see also Ke et al., 2024). Fig. Supplemental 1 shows the hierarchy of interactions between TADs to either side of eve (L-M, K-M and J-M in Fig. Supplemental 1). This conclusion is supported by previous transgene studies. As first described in the boundary bypass experiments of Cai and Shen (Cai and Shen 2001) and Muravyova et al. (Muravyova et al. 2001), the formation of a stem-loop structure when two boundaries are linked together brings sequences immediately beyond the linked boundaries into close proximity with each other. In this assay, a stem-loop structure is required, as it enables enhancers flanking the “upstream” boundary to interact with reporters flanking the downstream boundary (Kyrchanova et al. 2008a). In contrast, a circle-loop configuration does not bring the neighbors to either side of eve into close proximity.

While a stem-loop is the topology expected for the eve TAD in both models, one feature of the TAD is inconsistent with the loop-extrusion model. According to this model, loops (TADs) are generated by a cohesin complex that forms a small bubble at a loading site somewhere in the TAD-to-be and then extrudes both strands of the growing loop at an equal rate until it bumps into the boundary road-blocks (Davidson and Peters 2021; Ghosh and Meyer 2021; Mirny and Dekker 2022; Perea-Resa et al. 2021). If the loading site were located in the middle of the TAD-to-be, and the extrusion intermediates in a population of nuclei were captured by crosslinking, one should observe a crosslink-generated stripe that begins at the loading site and extends perpendicular to the chromosomal DNA until it reaches the apex that links the two TAD boundaries with imaginary lines of 45° (Fig. Supplemental 2B). This is the pattern that would be generated by a “broken zipper” when it transiently links the zipper teeth equidistant from the initial “loading site”. If the loading site is off center to the left, the crosslinking of the passing cohesin complex in a population of nuclei will generate a perpendicular stripe until cohesin come to a halt when it encounters the roadblock on the left (Fig. Supplemental 2C). Assuming that the complex continues to extrude the right strand, it will then generate a stripe at 45° degrees that extends to the apex of the TAD. If the loading site corresponds to one of the CTCF roadblocks or is located close to the roadblock, the cohesin complex will generate a 45° stripe that comes to a halt at the apex of the triangle when it encounters the neighboring CTCF roadblock (Fig. Supplemental 2D and E). If there are multiple loading sites within the TAD, there should be a series of perpendicular stripes that generate and sequentially reinforce the stripe(s) at 45° (Fig. Supplemental 2F). However, in spite of the fact that NC14 fly embryos should provide by far the best prospect of actually capturing loop extrusion intermediates, there are no vertical stripes in the eve TAD, nor are there stripes at 45° that extend to the apex of the eve TAD. Nor do we observe a population of perpendicular stripes that convert into a series of reinforced 45° stripes. There is also no evidence of stripes that are initially tilted to the left or right of a “loading site”, as might be observed if the rate of extrusion were unequal on the two strands, and then subsequently convert to a 45° stripe as cohesin comes to a halt at the first boundary encountered.

TADs in the eve environs

The attP insertion site at –142 kb is near the end of the first exon of the hebe transcription unit (Fig. 2C). It is located just within a ∼14 kb TAD, TB, that has a high density of internal contacts, and extends from the hebe promoter to the 3’ end of the neighboring dila gene. In between hebe and the eve locus there are at least ten distinct TADs, TC-TL (Fig. 2C). The endpoints of most of these TADs correspond to sequences that are associated with one or more known chromosomal architectural factors (dots along the horizontal axis in Fig. 2C).

These TADs correspond to the fundamental building blocks for organizing the 3D architecture of this chromosomal segment. Superimposed upon these TADs are regions that exhibit lower density of contacts (LDC). For example, the mef2 gene spans two TADs, TJ and TK. TK contains the mef2 distal promoter region and extends to the internal mef2 promoter, while TJ extends to near the promoter of the divergently transcribed Etf-QO gene (blue arrow). Both of these TADs are linked together by a rectangular LDC domain, J-K (see Fig. Supplemental 1). Likewise, TJ is linked to its immediate neighbor TI by the LDC domain I-J. In the next level, TADs separated by a single TAD are linked together. In the region immediately above TJ, the LDC domain I-K links TI to TK (Fig. Supplemental 1). Similarly, TJ is linked to TL by J-L while TJ is also linked to the large TAD TM (which contains TER94 and the neighboring gene, pka-R2) on the other side of eve by J-M. The same pattern of a hierarchical series of LDC domains linking TAD neighbors, TAD next-door neighbors and next-next-door neighbors is observed in the DNA segment that includes the hebe gene.

--cohesin-mediated loop extrusion: Like eve, the TADs in the region between eve and the attP site in the hebe locus are defined by right-angle triangles (volcanos) with a high density of internal contacts. However, unlike eve, these volcano triangles do not have a plume, and instead are surrounded by a series of LDC rectangular “clouds”. This is not the contact pattern that one might expect for either a series of stem-loops connected by unanchored loops or a series of stem-loops connected to each other. Moreover, there are no perpendicular or angled crosslinking stripes (broken zippers) emanating from the base of these triangles, nor are their stripes along their 45° legs. Thus, in spite of the fact that the NC14 nuclei used in this analysis are just emerging from an earlier mitosis and a round of DNA synthesis and should be in the process of assembling TADs, the crosslinking signatures that should be generated by the embrace of passing cohesin complexes are completely absent.

In the loop-extrusion model, the LDC domains would arise in a subset of nuclei because the extruding cohesin complex breaks through one or more boundary roadblocks before its progression is halted (Fig. Supplemental 3A versus B, C and D) (Hsieh et al. 2020; Krietenstein et al. 2020). Breakthroughs could occur at only one roadblock on one side of the extruding loop (Fig. Supplemental 3B and C), or at multiple roadblocks on one or both sides (Fig. Supplemental 3D), giving rise to a set of overlying LDC domains like that evident in the region between hebe and eve. However, while there would have to be multiple breakthrough events to account for the hierarchical array of LDC domains observed in these NC14 embryos, there is no evidence of the 45° crosslinking stripes that would be expected to mark the legs of the LDC domains following the breakthrough. For example, the I-J LDC domain (Fig. Supplemental 1) could be generated by cohesin complexes that initiated extrusion in either TI or TJ and then failed to halt at the TI:TJ boundary (green arrowhead). In either case, there should be a 45° stripe of crosslinking that marks (at least part of) the left (TI) and/or the right (TJ) leg of the I-J LDC domain and extends to the apex of the TI-TJ LDC triangle; however, there is no evidence of such stripes. Nor is there evidence of crosslinking stripes marking one or both legs of the J-K or I-K LDC domains (Fig. Supplemental 1).

While the LDC domains associated with the TADs located between eve and the hebe locus could potentially be explained by cohesin breaking through one or more boundaries (even though we do not observe the expected stripe intermediates), it is important to note that the eve boundaries, nhomie and homie, do not appear to be subject to as frequent break-through events. This is because, unlike the other TADs in the neighborhood, the LDC domains that link eve to TA and TL are much more lightly populated compared to the LDC domain that links TA to TL. --boundary:boundary pairing: At least two different mechanisms are expected to account for the TADs and LDC domains in the region between eve and the attP insertion site at -142 kb. First, since none of the TADs in this DNA segment (with the possible exception of TA) are topped by volcano plumes, they could correspond to circle-loops. As circle-loops are generated by head-to-head pairing, the boundaries in the region between nhomie and hebe would be expected to pair with each other head-to-head. (The TAD to TAD contact maps generated by circle-loops and stem-loops are considered further in the accompanying paper (Ke et al. 2024).)

As illustrated in Fig. Supplemental 3E-G, the coiled circle-loop TADs, though topologically independent, are expected to be in relatively close proximity to each other. This means that in addition to crosslinking events within the TADs, there will be crosslinking events between neighboring TADs. These crosslinking events could generate the LDCs seen in Fig. 2C and Fig. Supplemental 1. Since TADs next to each other (Fig. Supplemental 3E and F; Fig. Supplemental 1, I-J or J-K) would typically be expected to interact more frequently than TADs separated by one, two or more TADs (Fig. Supplemental 3G; Fig. Supplemental 1, I-K or J-L), there should be a progressive reduction in contact frequency at each step. This is generally what is seen (see Fig. Supplemental 1). While connected stem-loops generated by cohesin extrusion should also bump into one another, contacts between next-door neighbors might be expected to be less frequent than contacts between next-next-door neighbors (Fig. 1B, C and Fig. Supplemental 3A). (TAD-to-TAD crosslinking and the impact of loop topology is further examined in Ke et al. (Ke et al. 2024).

The second mechanism for generating LDC domains would be switching and/or combining pairing partners (Fig. Supplemental 3H-J). Though imaging studies suggest that pairing interactions can be of relatively long duration (>30 min: Chen et al. 2018; Vazquez et al. 2006), they are not permanent, and other nearby boundaries can compete for pairing interactions (Gohl et al. 2011). If switching/combining occurs in NC14 embryos (see Fig. Supplemental 3H-J), then the precise pattern of TADs and LDC domains could also be impacted by the relative avidity of potential pairing partners in the neighborhood and also the distances separating potential partners. For example, the rectangle linking TH and TI (H-I in Fig. Supplemental 1) could represent partner switching, so that the boundary between TI and TJ (green arrowhead in Fig. Supplemental 1) sometimes pairs with the boundary separating TH and TI (blue arrowhead in Fig. Supplemental 1) and sometimes with the boundary separating TG and TH (red arrowhead in Fig. Supplemental 1). Distant boundary elements could also occasionally interact with each other, generating a supra-TAD that contains multiple TAD domains. For example, at the apex of the LDC domain B-I there is an interaction dot (small black arrow, Fig. Supplemental 1) that links the TA-TB boundary to the TI-TJ boundary (green arrowhead).

Activation of a distant reporter by the eve enhancers

In previous studies we found that a minimal 367 bp homie fragment can orchestrate regulatory interactions between the eve enhancers and reporters inserted at an attP site in the first intron of the hebe transcription unit, 142 kb upstream of the eve gene (Fujioka et al. 2016). This minimal fragment lacks the one potential homie CTCF site, but has sites for several other generic boundary proteins including BEAF, Zipic, Su(Hw), Pita and Ibf2. In blastoderm stage embryos, lacZ reporter expression was observed in a 7-stripe pattern that coincided with the stripes of the endogenous gene. However, at this stage only a subset of eve-expressing nuclei also expressed the reporter. Later in development, during mid-embryogenesis, reporter expression was observed in cells in the dorsal mesoderm, the ventral CNS, and in the anal plate region, in a pattern that recapitulated eve gene expression (Fujioka et al. 2016). At these later stages, most of the eve-expressing cells also appear to express the reporter. The eve-dependent activation of the reporter in a stripe pattern has also been visualized in live imaging of pre-cellular blastoderm-stage embryos (Chen et al. 2018). Within the 7 eve stripes, the majority of the nuclei (∼80-85%) did not express the lacZ reporter. In these nuclei, the transgene was found to be at a considerable average physical distance from the eve locus (∼700 nM). In the remaining nuclei, the transgene was located in closer proximity to the eve locus (∼330 nM, on average). Within this subset, the transgene was expressed in most of the nuclei. Moreover, in most cases in which the transgene was active, it remained in close proximity and continued to express lacZ for the duration of the experiment (∼30 min).

In the two models for TAD formation, quite different mechanisms must be invoked to account for activation of the reporter at –142 kb by the eve enhancers. In the boundary-pairing model, the transgene homie boundary at –142 kb loops over the intervening TADs and pairs with the nhomie:homie complex flanking the eve TAD (c.f., Fig. 3B and C). In the loop-extrusion model, a cohesin complex initiating loop extrusion in the eve TAD must break through the nhomie roadblock at the upstream end of the eve TAD. It must then make its way past the boundaries that separate eve from the attP site in the hebe gene, and come to a halt at the homie boundary associated with the lacZ reporter. This would generate a novel TAD, eveMammoth (eveMa), that extends from the eve homie all the way to the homie fragment at –142 kb, and encompasses both the reporter and the eve gene, including its enhancers (c.f., Fig. 3D). Of course, the eveMa TAD could also be generated by a cohesin complex that initiated in, for example, TF. However, in this case, the runaway cohesin complex would have to break through the intervening boundary roadblocks in both directions. In both the boundary-pairing and loop-extrusion models, the configuration of the chromatin fiber would lead to the activation of lacZ expression by the eve enhancers, while the reporter would be protected from the hebe enhancers by the homie boundary.

Schematics of boundary pairing and loop extrusion.

A) GeimohL and GhomieL transgenes and the eve locus on a linear map. The same color codes are used throughout. In the eve locus, nhomie (blue arrow) and homie (red arrow) are oriented, by convention, so that they are pointing toward the right. This convention is maintained in the two transgenes. In GeimohL, homie (top) is in the opposite orientation from homie in the eve TAD and so is pointing away from the eve TAD. In GhomieL (bottom) homie is in the same orientation as homie in the eve locus and so is pointing towards the eve TAD. B) Predicted boundary pairing interactions between GeimohL and the eve TAD. homie in the transgene pairs with homie in the eve TAD head-to-head. Since homie in the transgene is pointing in the opposite orientation from homie in the eve TAD, a stem-loop will be generated. C) If homie in the transgene also pairs with nhomie in the eve TAD head-to-tail, a loop structure like that shown in C will be generated. In this topology, eve-lacZ is in close proximity to eve enhancers, and eve-gfp is in contact with the hebe enhancer. D) Loop-extrusion model for GeimohL. Transgene homie and endogenous homie determine the endpoints of the extruded eveMammoth (eveMa) loop. Like B, the topology of eveMa is a stem-loop. This topology brings eve-lacZ into close proximity to the eve enhancers, while eve-gfp is in another loop that contains the hebe enhancers. E) Predicted boundary pairing between GhomieL and the eve locus. The GhomieL transgene is inserted in the same chromosomal orientation as GeimohL; however, the homie boundary is inverted so that it is in the same orientation as the eve homie, and so it is pointing towards the eve TAD. homie in the transgene will pair with homie in the eve locus head-to-head, and this generates a circle-loop. F) If homie in the transgene also pairs with nhomie in the eve TAD, a loop structure like that shown in F will be generated. In this topology eve-gfp will be activated by both the eve and hebe enhancers. G) Loop-extrusion model for GhomieL. The cohesin complex bypasses transgenic homie and is stopped at an upstream boundary X, to generate a novel eveGarantuan (eveGa) loop. Both eve-gfp and eve-lacZ are located within the same eveGa TAD, and thus should interact with eve. Since the hebe enhancers are within this TAD as well, they would also be able to activate both reporters.

To test these two models, we used a transgene that has two divergently transcribed reporters, lacZ and gfp, that are each under the control of an eve promoter (Fig. 3A). Two different versions of the transgene were generated. In the first, the eve-lacZ reporter is located “upstream” of homie (with respect to the 5’→3’orientation of homie in the eve locus), while eve-gfp is located downstream. In the second, the orientation of homie within the transgene is reversed, so that eve-gfp is upstream of homie, while eve-lacZ is downstream. These two transgenes were then individually inserted into the attP site at –142 kb. For the pair shown in Fig. 3A, GeimohL and GhomieL, the transgenes were inserted so that eve-lacZ is on the eve side of the homie boundary. The eve-gfp reporter is on the opposite side of the homie boundary and is farther away from the eve locus. This places the eve-gfp reporter next to a series of hebe enhancers located farther upstream (Fig. 2C), in the intron of the hebe gene, and thus it should be subject to regulation by these enhancers. In the other pair, LhomieG and LeimohG, the entire transgene is inserted in the opposite orientation in the chromosome so that eve-gfp is on the eve side of homie, while eve-lacZ is next to the hebe enhancers (Fig. 6A). As a control, we also generated a transgene in which lambda DNA instead of homie was inserted between the eve-gfp and eve-lacZ reporters. This transgene was oriented so that eve-lacZ was on the eve side of the lambda DNA, and eve-gfp was close to the hebe enhancers.

The eve enhancers drive lacZ expression in the GeimohL insert

The results for transgenes GeimohL and the control GlambdaL are most straightforward and will be considered first. In both the loop-extrusion and boundary-pairing models, the GlambdaL control is not expected to display any regulatory interactions with the eve locus. Fig. 4C shows that this is the case: the eve-lacZ and eve-gfp reporters in GlambdaL embryos are silent at the blastoderm stage, and transcripts are only very rarely detected. Later in development, both reporters are expressed in a repeating pattern along the dorsal midline (Fig. 4E). This expression is driven by the hebe gene enhancers located just beyond the gfp-reporter (Fujioka et al. 2009).

The lacZ reporter is activated by eve enhancers in the GeimohL insert.

A) Expression patterns of endogenous eve in early (stage 4-5) and late (stage 14-16) embryos by smFISH. eve (Atto 633) probes were used to hybridize with eve mRNA. eve is shown in yellow and DAPI in blue. N>3. B) Schematics of transgenes. C) Expression patterns of transgenic reporters in early-stage embryos. Top: GlambdaL; bottom: GeimohL. GFP is in green, lacZ is in orange, and DAPI is in blue, here and in E. D) Quantification of normalized stripe signals for transgenes shown in C. The reporter smFISH intensity (signal to background), here and in F, was measured, normalized, and plotted as described in Methods. N>3. n=27 for GlambdaL and n=42 for GeimohL. E) Expression patterns of transgenic reporters in late-stage embryos. Top: GlambdaL; bottom: GeimohL. F) Quantification of normalized stripe signals for transgenes shown in E. N>3. n=27 for GlambdaL, n=56 for GeimohL. Scale bars = 200μm. N = # of independent biological replicates. n = # of embryos. The paired two-tailed t-test was used for statistical analysis. ****p < 0.0001, ns: not significant. Raw measurements are available in the Source Data files.

In the case of the GeimohL insert, regulatory interactions between the transgene and the eve locus will be observed if a stem-loop is established that links the transgene to eve. In the loop-extrusion model, the breakthrough cohesin complexes would generate a stem-loop, eveMammoth (eveMa), by coming to a halt at the homie boundary in the transgene and the homie boundary in eve, as illustrated in Fig. 3D. This will place the lacZ reporter in close proximity to the eve enhancers. It will also disrupt the eve TAD, as nhomie is no longer linked to the homie boundary by cohesin. In the boundary-pairing model, the lacZ reporter in the transgene is “upstream” of homie just like the eve gene and its enhancers are “upstream” of homie in the eve locus. Since homie pairs with itself head-to-head, pairing between homie in the transgene and homie in the eve locus will generate the same stem-loop as that generated by loop-extrusion (Fig. 3B). However, because homie pairs with nhomie head-to-tail, a more complex multi-loop structure like that in Fig. 3C would be expected if the pairing interactions with the transgene do not disrupt the eve TAD. Consistent with a tripartite structure of the sort shown here, previous studies have shown that three Mcp transgenes on three different chromosomes can interact with each other genetically (Muller et al. 1999). Subsequent direct visualization of Mcp-mediated pairing interactions in imaginal discs by Vazquez et al. (Vazquez et al. 2006) showed that four Mcp-containing transgenes separated by Mbs (at Bridges cytogenetic intervals ∼65 on the left arm of chromosome 3, and ∼83 and ∼95 on the right arm of chromosome 3) and/or located on different homologs, clustered in the same nuclear foci in 94% of the nuclei.

As predicted from the formation of a stem-loop linking the transgene to the eve locus, lacZ is expressed in blastoderm-stage embryos in 7 stripes that coincide with the 7 stripes of the endogenous eve gene. However, unlike eve, not all nuclei in the seven stripes express lacZ (compare with the eve control in Fig. 4A). Later, during mid-embryogenesis, lacZ is expressed in the mesoderm, the CNS, and the anal plate in a pattern that mimics the endogenous eve gene. At this stage almost all eve-positive cells are also positive for lacZ. Unlike the GlambdaL transgene, the hebe enhancers do not drive eve-lacZ expression, as they are beyond the homie boundary (Fig. 4E). The eve-gfp reporter in the GeimohL transgene responds differently. As was observed for the GlambdaL control, the eve enhancers drive little if any gfp expression, either at the blastoderm stage or later in development (Fig. 4C, E). Instead, gfp is expressed in the midline during mid-embryogenesis under the control of the hebe enhancers (Fig. 4E), which should be in the same TAD as eve-gfp. These results would be consistent with both the loop-extrusion (eveMa) and boundary-pairing models (Fig. 3B/C and D).

The eve enhancers drive gfp expression in the GhomieL insert

The situation is more complicated for the GhomieL transgene. This transgene is inserted in the same chromosomal orientation as GeimohL: eve-lacZ is on the eve side of the homie boundary and eve-gfp is on the hebe enhancer side of the boundary. However, the orientation of the boundary within the transgene is switched so that eve-gfp rather than eve-lacZ is “upstream” of homie. In the boundary-pairing model, flipping the orientation of homie in the transgene but keeping the same orientation of the transgene in the chromosome will have two consequences. First, as illustrated in Fig. 3E, the topology of the chromatin loop connecting the homie boundary in the transgene and the homie boundary in the eve locus will change from a stem-loop to a circle-loop. The reason for the change in loop topology is that homie pairs with homie head-to-head (Fujioka et al. 2016). As before, the transgene homie will pair with eve homie head-to-head to generate a simple circle-loop (Fig. 3E); however, if, as expected, it also pairs with nhomie so that the eve TAD remains intact, a more complicated loop structure would be generated (Fig. 3F). Second, the reporter that is preferentially activated by the eve enhancers will switch from eve-lacZ to eve-gfp. In order to be activated by the eve enhancer, the reporter must be “upstream” of the transgene homie. This physical relationship means that the eve-gfp reporter will be activated by the eve enhancers independent of the orientation of the transgene in the chromosome. On the other hand, the eve-lacZ reporter will not be activated by the eve enhancers. In addition, it will still be insulated from the hebe enhancers by the homie boundary.

It is not possible to form a circle-loop in the loop-extrusion model. Instead, the breakthrough cohesin complex will come to a halt as before when it encounters the inverted homie boundary at –142 kb (Fig. 3D). This means that the eve-lacZ reporter will be in the same TAD (eveMa) as the eve enhancers and will be activated by the eve enhancers. In contrast, the eve-gfp reporter will be in the neighboring TAD and will not be activated by the eve enhancers. It will, however, be regulated by the nearby hebe enhancers.

Fig. 5 shows that the four predictions of the boundary-pairing model are correct: a) the eve enhancers drive eve-gfp expression, b) the hebe enhancers drive eve-gfp expression, c) eve-lacZ is not subject to regulation by the eve enhancers, and d) eve-lacZ is insulated from the hebe enhancers. This would also mean that pairing of the transgene homie in the GhomieL insert with homie in the eve locus generates a circle-loop (or a more complicated multi-loop structure if it also pairs with nhomie), not a stem-loop, as was the case for GeimohL. In contrast, since loop-extrusion cannot generate circle-loops, the key predictions of this model, namely that eve-gfp will be silent while eve-lacZ will be regulated by the eve enhancers, are not satisfied. Instead, eve-gfp is activated by the eve enhancers, while eve-lacZ is not.

The GFP reporter is activated by both eve and hebe enhancers in the GhomieL insert.

A) Schematics of transgenes. B) Expression patterns of transgenic reporters in early-stage embryos. Top: GlambdaL; bottom: GhomieL. GFP is in green, lacZ is in orange, and DAPI is in blue, here and in D. C) Quantification of normalized stripe signals as represented in B. The reporter smFISH intensity (signal to background) was measured, normalized, and plotted as described in Methods, here and in E. N>3. n=27 for GlambdaL, n=46 for GeimohL. D) Expression patterns of transgenic reporters in late-stage embryos. Top: GlambdaL; bottom: GhomieL. E) Quantification of normalized stripe signals for transgenes shown in D. N>3. n=27 for GlambdaL, n=59 for GeimohL. Scale bars = 200μm. N = # of independent biological replicates. n = # of embryos. The paired two-tailed t-test were used for statistical analysis. ****p < 0.0001, ns: not significant. Raw measurements are available in the Source Data files.

While the results described above for the GhomieL transgene are inconsistent with a “simple” loop-extrusion model, there is one potential caveat: the 5’→3’ orientation of the homie boundary is inverted so that it is “pointing” towards the eve locus rather than away from it like the endogenous homie boundary. In this orientation, it is possible that homie no longer functions as an effective roadblock, and instead the cohesin complex emanating from the eve locus (or from a loading site somewhere in between) slips past this homie fragment and continues until it is blocked by a properly oriented boundary further upstream. In this case the newly formed stem-loop, eveGargantuan (eveGa) (Fig. 3G), would include the eve-gfp reporter, and it would then be activated by the eve enhancers. However, there are two problems with postulating this novel eveGa TAD. First, eve-lacZ would be in the same eveGa TAD as eve-gfp, and it should also be activated by the eve enhancers. Second, since the homie boundary was bypassed by the cohesin complex to form the larger eveGa TAD, it should not be able to block the hebe enhancers from activating eve-lacZ. Neither of these predictions is correct.

LeimohG and LhomieG: Orientation of homie in the transgenes determines which reporter is activated

To confirm these conclusions, we examined the expression patterns of the reporters when the transgene is inserted in the opposite orientation in the –142 kb attP site. In this transgene orientation, eve-gfp is on the eve side of the homie boundary while eve-lacZ is close to the hebe enhancers. In the boundary-pairing model, the orientation of homie in the transgene determines which reporter will be preferentially activated, while the orientation of the transgene homie in the chromosome determines topology of the resulting loop(s). The transgene homie will pair with the endogenous homie head-to-head. Since the transgene homie in LeimohG is in the opposite orientation from the endogenous homie, this pairing interaction will generate a stem-loop, if pairing interactions are exclusively pairwise. A more complicated multi-loop structure will be formed if the transgene homie interacts simultaneously with both endogenous boundaries (Fig. 6B). In both cases, eve-gfp will be activated by the eve enhancers. On the other hand, since this orientation places eve-lacZ on the same side of the transgene homie as the hebe enhancers, they will activate it rather than the eve-gfp reporter (Fig. 6B). In LhomieG, the topology of the loop (or multi-loop) will switch, and eve-lacZ will be activated by both the eve and hebe enhancers (Fig. 6C).

Expression of reporters in the LeimohG and LhomieG inserts.

A) Schematics of GE transgenes. B) Boundary pairing for LeimohG transgene long-range interactions. In this topology, GFP is in close proximity to eve enhancers, and lacZ is in contact with the hebe enhancer region. C) Boundary pairing for the LhomieG transgene. In this topology, lacZ is in close proximity to both the eve enhancers and the hebe enhancers. D) Loop-extrusion model for LeimohG. Transgenic homie and endogenous homie flank the eveMammoth loop. In this topology, GFP is in close proximity to eve enhancers, and lacZ is close to the hebe enhancers. E) Loop-extrusion model for the LhomieG transgene. The cohesin complex bypasses transgenic homie and is stopped at an upstream boundary X, to create the eveGarantuan loop. Both GFP and lacZ are at a similar physical distance from eve enhancers and hebe enhancers in this topology. F) Expression patterns of transgenic reporters in early-stage embryos. Top: LeimohG; bottom: LhomieG. GFP is in green, lacZ is in orange, and DAPI is in blue. N=3, n=24 for both LeimohG and LhomieG. G) Expression patterns of transgenic reporters in late-stage embryos. Top: LeimohG; bottom: LhomieG. N=3, n=24 for both LeimohG and LhomieG. Scale bars = 200μm, N = # of independent biological replicates, n = # of embryos. Statistical analysis is in Table Supplemental #1.

In the first version of the loop-extrusion model, the eve-gfp reporter in both LeimohG (Fig. 6D) and LhomieG (not shown) should be activated by the eve enhancers, since it is included in the eveMa TAD when the transgenes are inserted so that eve-gfp is on the eve side of the homie boundary, and eve-lacZ will be regulated by the hebe enhancers (Fig. 6D). In the second version of the loop-extrusion model, the functioning of the homie roadblock depends upon its orientation relative to endogenous homie. In this model, LeimohG is expected to form the eveMa TAD (Fig. 6D), while LhomieG will form the larger eveGa TAD (Fig. 6E).

As would be predicted by all three models, the eve enhancers activate the eve-gfp reporter in the LeimohG insert (Fig. 6F and G). Likewise, homie blocks the hebe enhancers from activating the eve-gfp reporter, while they turn on eve-lacZ expression instead. As predicted by the boundary-pairing model, eve-dependent expression of the reporters switches from gfp to lacZ in the LhomieG insert (Fig. 6F and G). This would not be expected in the first version of the loop-extrusion model, but it is predicted in the second version. However, as indicated in the diagram in Fig. 6E, the two reporters, along with the hebe enhancers, are included in the larger eveGa TAD, and for this reason both reporters should be activated equally by both the eve and hebe enhancers. This is not the case. The hebe enhancers are blocked by the intervening homie boundary, while there is little or no eve-dependent gfp expression.

TAD formation by transgenes inserted at –142 kb

We used MicroC to examine the TAD organization of the genomic region extending from upstream of the attP site through the eve locus in 12-16 hr embryos carrying different transgene insertions. At this stage, eve is expressed in only a small number of cells in the CNS, mesoderm and anal plate, and hebe is also expressed in only a small subset of cells. This means that most of the interactions detected by MicroC are in cells in which both the eve gene and the two reporters are inactive. In addition, while the transgene co-localizes with eve in less than about a fifth of blastoderm stage nuclei (based on the fraction of cells within eve stripes that activate a transgene reporter) the frequency of physical contact is expected to be much higher in 12-16 hr embryos (c.f., Vazquez et al.: Vazquez et al. 2006). This is because at this stage of development, cell cycles are much longer, providing more time for transgene – endogenous eve pairing to occur. The key findings are summarized below.

GlambdaL: While there is no evidence of contact between sequences in eve and sequences around the attP site at –142 kb in wild-type embryos, this not true for embryos carrying the dual reporter transgene. Perhaps the most unexpected finding is a weak, but clearly discernable pattern of interaction linking the GlambdaL transgene to the eve locus (Fig. 7A). However, since the GlambdaL transgene does not respond to the eve enhancers, this low level of physical interaction is not sufficient to drive detectable expression of the reporters. Though our experiments do not allow us to unequivocally identify which sequences in the transgene are responsible for this long-distance interaction, the most likely candidates are the two eve promoters in the GlambdaL transgene.

homie in transgenes LhomieG and LhomieG mediates long-range physical interactions with the eve TAD.

Diagrams for panels A, B, and C. yellow box: 5’ end of the hebe transcription unit. Transgene: green box: eve-gfp reporter, orange box: eve-lacZ reporter, black box: lambda DNA, red block arrow: homie DNA. eve TAD: blue block arrow: endogenous nhomie, red block arrow: endogenous homie; the directions of block arrows follows the established convention for nhomie and homie; gray boxes: eve enhancers, yellow box: eve transcription unit. A) MicroC contact profile of the control GlambdaL. Inset shows a blow-up of the contacts between the GlambdaL transgene and the eve TAD. Note a slight increase in interaction frequency (compare to Figure 1A). B, C) MicroC map of GeimohL and GhomieL, respectively. The key difference between the two inserts is the contacts between transgenes and sequences in the eve TAD. Note the changed pattern of interaction with the endogenous eve locus due to the orientation switch. D) “Virtual 4C” maps obtained from MicroC maps of “B” (top panel) and “C” (bottom panel). Viewpoints are shown in both panels from either the lacZ gene (orange) or the GFP gene (green).

GeimohL: Unlike GlambdaL, the eve-lacZ reporter in the GeimohL insert is activated by the eve enhancers, and this regulatory interaction is perfectly paralleled by the strong pattern of crosslinking between the transgene and the eve locus (Fig. 7B). On the eve side, the interactions between the transgene and the eve locus generate a heavily populated band of crosslinked sequences that span the 16 kb eve TAD. On the transgene side, there is an unequal distribution of crosslinked sequences to either side of the transgene homie boundary. As shown in the inset, the heaviest density of crosslinked sequences is on the eve-lacZ side of the transgene, consistent with the activation of this reporter by eve enhancers. To confirm that the interaction bias mimics the differences in activity of the two reporters in cells in which the reporters are expressed, we analyzed the interaction profiles from viewpoints within either the eve-lacZ or eve-gfp genes. Fig. 7D shows that eve-lacZ interacts with sequences extending across the eve TAD, while eve-gfp interactions are largely restricted to the eve homie element. While these results would seemingly be consistent with both the boundary-pairing and loop-extrusion models, there is no evidence of cohesin breaking through multiple intervening boundaries into order to establish the novel eve(Mam) TAD. Instead, the organization of the TADs and LDC domains in the interval in between the transgene and eve (Fig. 7B) resemble that seen in wild-type NC14 embryos (see Fig. 2) (with the caveat that there are significantly fewer reads in the transgene experiment). Equally important, the same is true for the eve TAD. In order to generate the eve(Mam) TAD, the cohesin complex must break through the nhomie boundary and, in so doing, disrupt the eve TAD. However, there are no obvious alterations in either the eve volcano or plume with respect to their structure or relative contact density compared to the GlambdaL and wild type NC14 embryos. Likewise, there is no evidence of novel vertical and 45° crosslinking stripes that would reflect the boundary breakthrough events that are expected in the loop-extrusion model to accompany the formation of a physical connection between the transgene and the eve locus.

We next used the transgene homie as the viewpoint to test the predictions of the boundary-pairing and loop-extrusion models. Since homie can pair with itself (head-to-head) and with nhomie (head-to-tail) in transvection assays (Fujioka et al. 2016), a prediction of the boundary-pairing model is that there will be physical interactions between the transgene homie and both homie and nhomie in the eve locus (c.f., Vazquez et al. 2006). Also based on what is known about how fly boundaries pair with each other (Chetverina et al. 2014; Chetverina et al. 2017; Kyrchanova et al. 2008a), self-interactions are expected to be stronger than heterologous interactions. In contrast, in the loop-extrusion model, the cohesin complex forms the eveMa TAD by linking homie in the transgene to homie in the eve locus, and this requires the cohesin complex to break through nhomie and disrupt the connections between nhomie and homie in the eve locus. In this case, transgene:homie←→eve:homie interactions should be observed, while transgene:homie←→-eve:nhomie should not. Fig. 8A shows that, as predicted by the boundary-pairing model, the transgene homie interacts with both homie and nhomie in the eve locus.

homie in the transgene interacts with eve homie and nhomie.

“Virtual 4C” maps of the GeimohL and LhomieG (A) or GhomieL and LeimohG (B) transgenes oriented so that lacZ (blue) or gfp (orange) is located on the eve side of the homie boundary. Viewpoints are taken from the transgenic homie sequence. Note that transgenic homie preferentially interacts with endogenous homie in both orientations.

Since the eve TAD does not appear to be disturbed by the presence of the GeimohL transgene, these findings would suggest that the transgene homie likely pairs with both nhomie and homie in a pattern somewhat like that illustrated in Fig. 3C. While the physical interactions between the transgene and eve boundaries are inconsistent with the loop-extrusion model, one could make the ad hoc assumption that runaway cohesin complexes form not one but two different novel TADs. One would be eveMa. In the other, eveElephant (eveEl), a cohesin complex that initiates in, for example, the TE TAD would come to a halt on one side at the homie boundary in the transgene, and on the other side at the nhomie boundary in the eve locus (Fig. Supplemental 4). However, this configuration would not explain how the transgene reporter is activated by eve enhancers. Since the eve enhancers in eveEl are located in a different TAD than the TADs containing eve-lacZ and eve-gfp, they would not be able to activate either reporter.

GhomieL: When homie is reversed in the GhomieL transgene, the eve-gfp reporter responds to the eve enhancers instead of eve-lacZ. In the boundary-pairing model, eve-gfp is activated because self- and heterologous-pairing interactions between homie and nhomie are orientation-dependent. As a result, the topology of the loop formed between the transgene and eve is transformed in a way that brings the eve-gfp reporter instead of the eve-lacZ reporter into close proximity to the eve enhancers (see Fig. 3E, F). Since only stem-loops can be generated in the loop-extrusion model, activation of the distal eve-gfp reporter would require the formation of a larger eveGa TAD in order to encompass both the transgene and the eve enhancers in the same looped domain (Fig. 3G). However, as shown in Fig. 7C, we do not observe the cross-linking pattern expected for a new eveGa TAD that includes the entire transgene at –142 kb and then extends to an even more distal boundary. Nor are there any novel vertical/45° crosslinking stripes or obvious perturbations in the basic TAD organization in the region beyond –142 kb, or between –142 kb and the eve locus, as might be expected for cohesin complexes breaking through multiple intervening roadblocks. Instead, as was the case for GeimohL, the GhomieL transgene is physically linked to the eve locus, and there is a prominent band of crosslinked transgene sequences that spans the eve TAD (Fig. 7C). Consistent with the expression patterns of the two reporters, the heaviest density of crosslinked sequences is on the eve-gfp side of the transgene (see inset). This is confirmed by the interaction profiles generated using either the eve-gfp or eve-lacZ gene body as the viewpoint: the most frequent contacts are between sequences within eve and the eve-gfp reporter (Fig. 7D). In addition, as was the case for GeimohL, when the transgene homie is used as the viewpoint, interactions are observed between both homie and nhomie in the eve locus, with the most frequent corresponding to self-pairing interactions (Fig. 8B). Other than postulating the ad hoc eveEl TAD (Fig. Supplemental 4), this would not be expected for either version of the loop-extrusion model.

LhomieG and LeimohG: To confirm that the orientation of the homie boundary in the transgene is the key determinant of the physical interactions between the transgene reporters and the eve TAD, we used MicroC to analyze the TAD organization of LhomieG and LeimohG (Figs. Supplemental 5 and 6). Like their counterparts GhomieL and GeimohL, both show a strong band of interaction between the transgene and sequences spanning the eve locus. In the case of LeimohG, the pairing of the transgene homie with the endogenous homie is predicted to generate a stem-loop, while a more complicated multi-loop structure will be generated if it also pairs with nhomie (see Fig. 6B). As shown in Fig. Supplemental 5A, these pairing interactions bring eve-gfp into contact with the eve enhancers and, as expected, the eve-gfp sequences interact more frequently with the eve locus than does eve-lacZ (inset Fig. Supplemental 5A). When homie is inserted in the transgene in opposite orientation, homie self-pairing is expected to generate a circle-loop, while the more complicated structure shown in Fig. 6C will be formed if it also pairs with nhomie. While the sample preparation for MicroC was of poorer quality than the others, the inset in Fig. Supplemental 6A shows that, as expected, eve-lacZ interacts more frequently with sequences in the eve locus than does eve-gfp. On the other hand, there is no evidence of a larger eveGa TAD, even though eve-lacZ is activated by the eve enhancers. These interactions are confirmed when either eve-lacZ or eve-GFP are used as viewpoints to plot the physical contacts between the eve TAD and the two reporters in LhomieG and LeimohG (Figs. Supplemental 5B and 6B). In addition, as was the case for the GhomieL and GeimohL transgenes, when the transgene homie is used as the viewpoint, interactions are observed between homie in the transgene and both homie and nhomie in the eve locus (Fig. 8A and B). Again, the most frequent interactions are self-pairing interactions between homie in the transgene and the eve homie.

Loop topology and interactions with neighbors: The results in the previous sections demonstrate that the orientation of homie in the transgene determines which of the two reporters physically interacts with sequences in the eve TAD, and this is independent of the orientation of the transgene in the chromosome. On the other hand, the orientation of the transgene homie in the chromosome impacts how sequences flanking the transgene homie and the eve TAD interact with each other. With the important caveat that the depth of our MicroC sequencing is limited, the interaction profiles reflect the topology of the transgene-induced loop. For homie pointing away from eve (GeimohL and LeimohG), sequences to the TER94 side of eve and to the hebe enhancer side of the transgene are more frequently crosslinked with each other (Fig. Supplemental 7). These interactions would be expected for either a stem-loop structure in Fig. 3B or the more complicated multi-loop structure shown in Fig. 3C. For homie pointing towards eve, sequences on the TER94 side of eve interact with sequences on the eve side of the transgene. This would be expected for either a circle-loop structure in Fig. 3E or the more complicated multi-loop structure shown in Fig. 3F (neither of which can be generated by a simple loop-extrusion mechanism).

Discussion

Two different mechanisms have been proposed to explain how TADs are formed and their endpoints determined. In the first, a cohesin complex first induces a small loop at a loading site within the TAD-to-be and then progressively extrudes one or both strands until the complex encounters boundary roadblocks. If the loading site is located asymmetrically within the TAD-to-be, extrusion will continue on the unblocked strand until a roadblock is encountered. The endpoints of the loops were initially thought to be determined by the location of the first two roadblocks encountered by the cohesin complex (Alipour and Marko 2012). In a refinement of this model, the orientation of a given roadblock relative to the direction of movement of the cohesin complex determines whether it will come to a halt (Davidson and Peters 2021; Fudenberg et al. 2016; Guo et al. 2012; Guo et al. 2015; Sanborn et al. 2015). It has also been suggested that cohesin complexes can break through a roadblock and come to a halt at more distant roadblocks, so that in a population of cells analyzed by Hi-C or MicroC, the basic units of organization, the TADs, are overlaid by a hierarchy of LDC domains (Hsieh et al. 2020; Krietenstein et al. 2020). However, the only loop topologies that are possible in the loop-extrusion model are a stem-loop and an unanchored loop.

In the other mechanism, TAD formation is governed by boundary:boundary pairing interactions. In this case, the critical factors are the pairing and orientation preferences of potential partners, together with proximity. Since boundary pairing interactions often exhibit an orientation preference, the topology of the loop formed by paired boundaries can be either a stem-loop or a circle-loop, depending on their relative orientation in the chromosome. In this respect it differs from the loop-extrusion model in that switching the orientation of the boundary does not preclude the formation of a TAD, whereas it can in some versions of the loop-extrusion model.

In the studies reported here, we have used MicroC to analyze the TAD organization of a ∼150 kb chromosomal segment flanking the eve locus, and have experimentally tested the predictions of these two models for the mechanisms of TAD formation. Our studies do not support a loop-extrusion mechanism, and are inconsistent with this model on multiple levels. Manipulating “TAD” formation: In our experimental paradigm, a transgene containing two divergently transcribed reporters and the eve boundary homie is inserted 142 kb upstream of the eve TAD. Depending upon the orientation of the homie boundary within the transgene, either eve-gfp or eve-lacZ is activated by developmentally regulated enhancers in the eve TAD. Since there are a series of well-defined TADs and their associated boundary elements separating the transgene from eve, a version of the loop-extrusion model in which the cohesin complex invariably arrests upon encountering nhomie and homie at the ends of the eve TAD cannot account for this long-distance regulation. Instead, the extruding cohesin complex must be able to break through the boundaries located between homie in the transgene and homie in the eve TAD. This would generate a novel TAD, eveMa, with endpoints corresponding to the two homie boundaries (Fig. 3D). In this case, the reporter on the eve side of the transgene homie would be in the same domain as the eve enhancers, and thus potentially subject to regulation. While eveMa could account for reporter expression in inserts in which the transgene homie is pointing away from eve (and also away from the reporter on the eve side of the transgene insert), it does not explain reporter expression when the transgene homie is pointing towards eve (and in this case towards the reporter on the eve side of the transgene insert). In the latter case, the reporter activated by the eve enhancers lies outside of the eveMa (cf., Fig. 3D), and thus is not included in the same TAD as the eve enhancers. In order to account for activation of the distal reporter, one must imagine that the cohesin complex breaks through the homie boundary in the transgene and continues on until extrusion is halted by an even more distal boundary. In this revision of the loop-extrusion model, both reporters would be included in the same TAD as the eve enhancers (c.f., Fig. 3G, eveGa), and both reporters should be activated. However, this is not observed. TAD organization: loop extrusion versus boundary pairing: Analysis of the TAD organization in the region spanning the transgene insertion site and eve also does not fit with the expectations of the loop-extrusion model. To account for reporter activation, the cohesin complex must be able to break through the nhomie boundary. This breakthrough could be engineered by a cohesin complex that initiated loop extrusion either from within the eve TAD or from one of the TADs located between the nhomie boundary and the transgene at –142 kb. However, in the absence of the transgene, there is no indication that nhomie is particularly prone to break-through events. In fact, the eve TAD differs from most of the TADs in between eve and the transgene in that it is incorporated into larger LDC domains much less frequently. There is also no reason to imagine that the presence of a transgene carrying homie (or nhomie: see Fujioka et al. 2016) at –142 kb would somehow induce nhomie break-through events.

As expected from the regulatory interactions, homie-containing transgenes at –142 kb generate a strong band of crosslinking events that physically link the transgene to the eve locus. In the loop-extrusion model, this strong band of interaction should generate a novel interaction triangle (LDC or HDIC) with endpoints corresponding to the homie elements in the transgene and eve. Contrary to this expectation, a triangle of increased interaction frequencies spanning the entire region between –142 kb and homie is not observed for any of the transgene inserts. In viewpoints from the homie element in the GeimohL and LeimohG transgenes, there are two peaks in the eve TAD. The more prominent peak maps to homie, while the less prominent peak maps to nhomie. To account for these interaction peaks, the loop-extrusion model would have to assume that the presence of the GeimohL and LeimohG transgenes induces the formation of two novel TADs, the eveMa (homie:homie) TAD in Fig. 3D and another TAD, eveElephant, that links the transgene homie to the nhomie element in the eve locus (Fig. Supplemental 4B). However, while the eveMa TAD could potentially explain the activation of reporters on the proximal side of the transgene homie by the eve enhancers, these TADs do not explain how the eve enhancers are able to activate reporters on the distal (hebe enhancer) side of the GhomieL and LhomieG transgene. This would require the formation of yet another TAD, eveGa, that encompasses eve, the transgene, and one or more TADs beyond the hebe gene (Fig. 3G). However, there is no indication of physical interactions between the homie boundary in eve and a boundary located beyond the hebe gene. Moreover, like eveMa, eveGa would also have to be accompanied by the formation of eveEl, since the homie element in GhomieL and LhomieG interacts with both nhomie and homie.

While these results are inconsistent with the expectations of the loop-extrusion model, they dovetail nicely with many of the predictions of the boundary-pairing model. First, in all four transgenes, homie interacts with both homie and nhomie in the eve locus, but does not interact with the multiple boundary elements located in between. This is consistent with other studies which show that boundaries often have strong partner preferences (Gohl et al. 2011; Kyrchanova et al. 2011; Kyrchanova et al. 2008b). Second, interaction frequencies are greater for self-pairing than for heterologous pairing. This fits with what is known about how partner preferences are determined (Blanton et al. 2003; Chetverina et al. 2017; Erokhin et al. 2021; Kyrchanova et al. 2008a; Zolotarev et al. 2016). Third, the pattern of reporter activation in the four transgenes is precisely as expected from previous studies on the orientation dependence of homie pairing with itself (head-to-head) and with nhomie (head-to-tail) (Chen et al. 2018; Fujioka et al. 2016). This orientation dependence means that the reporter located “upstream” of homie is the one that is activated, independent of the orientation of the transgene in the chromosome. Fourth, the pattern of physical interactions between the eve locus and the transgenes recapitulates the pattern of reporter activation. That is, when eve-lacZ sequences are preferentially crosslinked to sequences in the eve locus, then the eve-lacZ reporter is activated by the eve enhancers. Likewise, eve-gfp is preferentially crosslinked to sequences in eve when it is activated by the eve enhancers. Again, this is dependent on the orientation of homie in the transgene, but not on the orientation of the transgene in the chromosome. Fifth, unlike the loop-extrusion model where only stem-loops or unanchored loops can be generated, stem-loops, circle-loops and even more complicated structures can be, and, as we have shown here, actually are generated by boundary:boundary pairing.

Are the pairing interactions of homie and nhomie unusual? There are two reasons to think that the properties of most fly boundaries are similar to those of homie and nhomie. First, all of the fly boundaries that have been tested in assays that are expected to require pairing do have this activity (Blanton et al. 2003; Cai and Shen 2001; Erokhin et al. 2011; Gohl et al. 2011; Kyrchanova et al. 2008a; Kyrchanova et al. 2011; Kyrchanova et al. 2007; Kyrchanova et al. 2008b; Muravyova et al. 2001).

Second, multimerized binding sites for generic polydactyl zinc finger proteins like Su(Hw), Pita, dCTCF, Zw5, and Zipic that are found associated with many fly boundaries are not only able to block enhancer/silencer action but can also mediate pairing interactions in insulator bypass and transvection assays (Erokhin et al. 2021; Kyrchanova et al. 2008a; Zolotarev et al. 2016). Moreover, as is the case for homie and nhomie, pairing interactions require the appropriate partners: multimerized sites for Zw5 do not pair with multimerized sites for dCTCF (Erokhin et al. 2021; Kyrchanova et al. 2008a; Zolotarev et al. 2016). This makes good sense, since these proteins form dimers/multimers, as do other fly boundary factors like BEAF, Mod(mdg4), the BEN-domain protein Insensitive and the Elba complex (Avva and Hart 2016; Bonchuk et al. 2021; Bonchuk et al. 2011; Bonchuk et al. 2020; Fedotova et al. 2018; Fedotova et al. 2019; Fedotova et al. 2017; Hart et al. 1997). The ability to multimerize means that boundaries that share binding sites for the same protein, i.e., for Pita, Zipic or CTCF, can be physically linked to each other by a single protein complex.

Third, homie and nhomie are not the only examples of fly boundary elements that engage in specific orientation-dependent physical pairing interactions. In addition to gypsy, Mcp and Fab-7, Mohana, et al. (Mohana et al. 2023) identified nearly 60 meta-loops in which distant TADs are linked together by what appears in most instances to be orientation-dependent boundary pairing interactions. Fig. 9 shows the MicroC contact profiles generated for two such meta-loops in 12-16 hr embryos. As observed in our homie transgene experiments, the meta-loop in panel A is formed by specific and orientation-dependent boundary:boundary (blue:purple) pairing interactions that bring together distant TADs. The blue boundary splits the Leukocyte-antigen-related like (Lar) gene into a small upstream TAD that contains the most distal promoter (black arrowhead), and a larger downstream TAD whose endpoint is close to an internal lar promoter (green arrowhead). The purple boundary is located ∼600 kb away on the left arm of the 2nd chromosome between two TADs, Pa and Pb, in a gene poor region. The pairing of the blue and purple boundaries generates two rectangles of enhanced contacts. The one on the upper left corresponds to interactions of the sequences in the small TAD just upstream of the blue boundary with sequences in the Pa TAD located upstream of the purple boundary. The other box on the lower right is generated by interactions between sequences in the TAD downstream of the blue boundary and sequences downstream of the purple boundary. Based on the locations of interacting TADs, the blue and purple boundaries pair with each other head-to-head, generating the circle-loop in the diagram.

Circle-loop and stem-loop meta-loops.

Distant boundary elements can find and pair with each other, forming large meta-loops. Panel A shows a circle-loop meta-loop formed by the head-to-head pairing of the blue (blue arrow) and purple (purple arrow) boundaries. As described in the text, the blue boundary separates a small TAD that contains the most distal Lar promoter (indicated by the black arrowhead) and a larger TAD that contains part of the Lar transcription unit (and is upstream of an internal Lar promoter: green arrowhead). The purple boundary is located between two TADS, Pa and Pb, in a gene-poor region of the 2nd chromosome approximately 600 kb away downstream of the Lar gene. When the blue and purple boundaries pair, sequences in the small TAD upstream of the blue boundary come into contact with sequences in the Pa TAD upstream of the purple boundary. This interaction generates the dark rectangle on the upper left side of the interaction plot. Sequences downstream of the blue boundary in the TAD containing part of the Lar transcription unit are also brought into contact with sequences in the Pb TAD downstream of the purple boundary. This interaction generates the larger rectangular box on the bottom right side of the interaction plot. Shown in the diagram on the right is a schematic illustrating that the head-to-head pairing of the blue and purple boundaries generates a ∼600 kb circle-loop. In the circle loop configuration, sequences in the TADs upstream of the blue and purple boundaries come into contact with each other, as do sequences downstream of the blue and purple boundaries (blue double arrows). Panel B shows a stem-loop meta-loop formed by the head-to-tail pairing of a boundary (blue arrow) located upstream of the beat IV gene with another boundary (purple arrow) located ∼6 Mb away, between the Sid and snu genes. The blue boundary separates a small TAD containing three UDP-glycosyltransferase genes from a larger TAD containing the CG10164 gene and the distal-most promoter of the beat IV gene. The purple boundary separates a ∼20 kb TAD containing Sid and four other genes from a small TAD that contains the most distal snu promoter (green arrowhead). When the blue and purple boundaries pair with each other head-to-tail (see diagram on right), this interaction brings sequences in the large TAD downstream of the blue boundary into contact with sequences upstream of the purple boundary in the 20 kb Sid TAD, and this generates the large rectangular box to the lower left of interaction map. The stem-loop configuration also brings sequences upstream of blue boundary in the small TAD containing the three UDP-glycosyltransferase genes into contact with the TAD containing the distal-most snu promoter. This interaction gives a small rectangular box with a high contact density. Sequences in the TAD upstream of the UDP-glycosyltransferase TAD are also brought into contact with the small TAD containing the snu promoter. These two interacting “zones” are indicated by blue double arrows in the diagram on right. In addition to boundary:boundary pairing, both meta-loops have a prominent dot in the larger regular boxes indicated by the green (panel A) and blue (panel B) arrows. Based on the ChIP signature of the sequences giving rise to these dots, they likely correspond to elements that are bound by the GAF containing LBC complex, as indicated in the figure (“LBC” in both panels). LBC elements in other contexts function to bring enhancers into contact with promoters (Batut et al. 2022; Kyrchanova et al. 2023; Kyrchanova et al. 2019a; Kyrchanova et al. 2019b; Levo et al. 2022).

The meta-loop in panel B links a boundary (blue arrow) to a boundary (purple arrow) ∼5 Mb away on the right arm of the 3rd chromosome. The blue boundary is located between a small TAD that contains three genes encoding UDP-glycosyltransferases and a larger TAD that includes CG10164 and the most distal beat IV (beaten path IV) promoter. The purple boundary separates a large TAD containing five genes (Sid, CG33346, CG31050, CG14062 and CG9988) from a small TAD that contains the most distal promoter of the snu (snustorr) ABC transporter. The blue and purple boundaries pair with each other head-to-tail, and this generates a stem-loop meta-loop (see diagram). In this pairing configuration, sequences upstream of the blue boundary come into contact with sequences downstream of the purple boundary. This generates a box of interactions between the small TAD that contains the UDP-glycosyltransferase gene cluster and the small TAD that contains the most distal snu promoter. (Another weaker rectangular block of interactions links sequences in the TAD immediately upstream of the UDP-glycosyltransferase gene cluster, which contains CG10175, and the small TAD containing the distal snu promoter). A second, larger rectangular box is generated by contacts between sequences in the TADs downstream of the blue boundary and upstream of the purple boundary. Note that the positioning of the rectangular interaction boxes differs for circle-loops and stem-loops.

The pairing interactions in the lar and beat IV meta-loops are much simpler than those seen for homie in the dual reporter. In both cases, there is only a single contact point (blue:purple boundaries). In contrast, as shown in Fig. 7 and in Figs. Supplemental 5 and 6, the interaction between the dual reporter homie and the eve TAD generates a “stripe” that spans the eve TAD together with crosslinking events between TADs upstream and downstream of the dual reporters and TADs located to either side of the eve TAD. This more complicated interaction pattern is consistent with the viewpoint analysis (Fig. 8), which shows that the transgene homie physically interacts with both homie and nhomie in the eve TAD.

Boundary pairing orientation and loop topology: Our analysis of the TAD organization in the eve neighborhood suggests that the predominant loop topology may be a circle-loop, not a stem-loop. Only the eve TAD (and perhaps the very small TAD TA) is clearly a stem-loop. It is topped by a distinctive volcano plume and is not overlaid by a hierarchy of LDC domains, indicating that it must be “insulated” from crosslinkable interactions with other nearby TADs. By contrast, none of the TADs in the region spanning eve and the hebe gene have volcano plumes at their apex. Instead, all of these TADs are overlaid by a hierarchal organization of LDC domains (clouds) that diminish in crosslinking intensity as the number of in-between TADs increases. As discussed above, the contact pattern of the TADs to the left of eve are likely to arise from partner switching and/or head-to-head boundary pairing interactions that generate circle-loops. Pairing orientation versus loop topology is considered further in the accompanying paper (Ke et al. 2024).

Orientation and long-distance regulatory interactions: Our experiments highlight an important constraint on long-distance regulatory interactions that are mediated not only by boundary elements like homie, nhomie, and the meta-loop boundaries in Fig. 9, but also by the GAF-dependent (LBC) tethering and bypass elements (Batut et al. 2022; Kyrchanova et al. 2023; Kyrchanova et al. 2018; Kyrchanova et al. 2019a; Kyrchanova et al. 2019b; Levo et al. 2022; Li et al. 2023; Mohana et al. 2023), namely the orientation of the physical interactions that bring distant enhancers (or silencers) into close proximity with their gene targets. As we have shown here, the orientation of the physical interactions that bring distant sequences close together determines whether the enhancer will be able to activate its target gene. In principle the effects of orientation could be mitigated by increasing the distance between the boundary/tethering elements and either the enhancers (silencers) or the promoter. However, distance can be a critical, rate-limiting step in transcriptional activation. For example, live imaging experiments have shown that the frequency and amplitude of transcriptional bursts depend on the distance between the enhancer and target promoter. Nearly continuous bursting and high amplitude transcription was observed when the enhancer was located close the promoter (Fukaya et al. 2016). However, at distances as little as 6 kb, transcription became discontinuous with discrete bursts, while at 9 kb, the bursts were much less frequent, and the amplitude/output was substantially reduced (Yokoshi et al. 2020). If similar distance-dependent reductions in transcriptional efficiency are common, then it would not be possible to overcome the topological constraints on regulatory interactions by simply increasing the distance between the boundary/tethering elements and the enhancers/promoters. While our studies do not directly address how chromosome architecture impacts regulatory interactions in vertebrates, it is clear that long-distance regulatory interactions in vertebrates (c.f., Tan et al. 2023) will be subject to the same types of topological constraints as those that apply in flies.